### 2014-05-12 abstract.tex
first commit

We have developed a virtual machine for Linear Meld, a forward-chaining linear logic programming language
where programs can be executed in parallel. In Linear Meld, a program is seen as a database of logical facts and a set of derivation rules.
The database of facts is partitioned by the nodes of a graph structure which leads
to parallelism when nodes are executed simultaneously.
Due to the foundations on linear logic, rules can retract facts in a declarative and structured fashion, leading
to more expressive programs.
We present the design and implementation of the virtual machine that executes on multicores,
with a focus on thread management, code organization, fact indexing, rule execution, and database organization for efficient fact insertion,
lookup and deletion. Our results show that the virtual machine is not only
capable of scaling programs with up to 16 threads but also exhibits interesting performance results when
compared against other programming languages.

### 2014-05-12 conclusions.tex
first commit

We have presented a parallel virtual machine for executing forward-chaining linear logic programs, with particular focus
on parallel scheduling on multicores, rule execution and database organization for fast insertion,
lookup, and deletion of linear facts.
Our experimental results show that the VM is able to scale the execution of programs when run with up to 16 threads.
Although this is still a work in progress, the VM also fairs well against other programming languages.
Moreover, since LM programs are concurrent by default, we can easily get better performance from the start by executing
them with multiple threads.

In the future, we intend to take advantage of linear logic to perform whole-program optimizations,
including computing program invariants, loop detection in rules and bypass of rule priorities. For instance,
we want to detect during compile time which might be the best data structure for each predicate depending in how it
is used in the rules.

### 2014-05-12 language.tex
first commit

\newcommand{\selector}[0]{[~S~\Rightarrow~y;~BE~] \lolli HE}
\newcommand{\comprehension}[0]{\{~\widehat{x};~BE;~SH~\}}
\newcommand{\aggregate}[0]{[~A~\Rightarrow~y;~\widehat{x};~BE;~SH_1;~SH_2~]}

Linear Meld (LM) is a logic programming language that offers a declarative and structured way to manage state.
A program consists of a database of facts and a set of derivation rules. The database includes persistent
and linear facts. Persistent facts cannot be deleted, while linear facts can be asserted and retracted.

The dynamic (or operational) semantics of LM are identical to Datalog.
Initially, we populate the database with the program's axiom facts and then determine which derivation rules can be applied by using the current database. Once a rule is applied, we derive new facts, which are then added to the database.
If a rule uses linear facts, they are retracted from the database.
The program stops when \emph{quiescence} is achieved, that is, when rules no longer apply.

Each fact is a predicate on a tuple of \emph{values}, where the type of the predicate prescribes the types of the arguments.
LM rules are type-checked using the predicate declarations in the header of the program. LM has a simple type system that includes types such as
\emph{node}, \emph{int}, \emph{float}, \emph{string}, \emph{bool}. Recursive types such as \emph{list X} and \emph{pair X; Y} are
also allowed.
Each rule in LM has a defined priority that is inferred from its position in the source file.
Rules at the beginning of the file have higher priority. We consider all
the new facts that have been not considered yet to create a set of \emph{candidate rules}.
The set of candidate rules is then applied (by priority) and updated as new facts are derived.

\subsection{Example}

\begin{figure}[t]
{\footnotesize
\begin{Verbatim}[numbers=right]
type left(node, node).
type right(node, node).
type linear value(node, int, int).
type linear replace(node, int, int).

replace(A, K, New),
value(A, K, Old)
   -o value(A, K, New). // we found our key

replace(A, RKey, RValue),
value(A, Key, Value),
!left(A, B),
RKey < Key // go left
   -o value(A, Key, Value),
      replace(B, RKey, RValue).

replace(A, RKey, RValue),
value(A, Key, Value),
!right(A, B),
RKey > Key // go right
   -o value(A, Key, Value),
      replace(B, RKey, RValue).

// binary tree configuration
value(@3, 3, 3). value(@1, 1, 1). value(@0, 0, 0).
value(@2, 2, 2). value(@5, 5, 5). value(@4, 4, 4).
value(@6, 6, 6).
!left(@1, @0). !left(@3, @1). !left(@5, @4). 
!right(@1, @2). !right(@3, @5). !right(@5, @6).

replace(@3, 6, 7). // replace value of key 6 to 7
\end{Verbatim}
}
\caption{Binary tree dictionary: replacing a key's value.}
  \label{code:btree_replace}
\end{figure}

We now present an example LM program in Fig.~\ref{code:btree_replace} that implements the key update operation for a binary tree
represented as a key/value dictionary.
We first declare all the predicates (lines 1-4), which represent the kinds of facts we are going to use.
Predicate \texttt{left/2} and \texttt{right/2} are persistent while \texttt{value/3} and \texttt{replace/3} are linear.
The \texttt{value/3} predicate assigns a key/value pair to a binary tree node and the \texttt{replace/3} predicate
represents an update operation that updates the key in the second argument to the value in the third argument.

The algorithm uses three rules for the three cases of updating a key's value: the first rule performs the update
(lines 6-8); the second rule recursively picks the left branch for the update operation (lines 10-15); and the third
rule picks the right branch (lines 17-22).
The axioms of this program are presented in lines 24-29 and they describe the initial binary tree configuration,
including keys and values. By having the \texttt{update(@3, 6, 7)} axiom instantiated at the root node $@3$, we intend to
change the value of key 6 to 7.
Note that when writing rules or axioms, persistent facts are preceded with a \texttt{!}.

Figure~\ref{fig:btree_trace} represents the trace of the algorithm. Note that the program database is partitioned
by the tree nodes using the first argument of each fact. In Fig.~\ref{fig:btree_trace}a we present the database
filled with the program's axioms. Next, we follow the right branch using rule 3 since $6 > 3$ (Fig.~\ref{fig:btree_trace}b).
We then use the same rule again in Fig.~\ref{fig:btree_trace}c where we finally reach the key 6. Here, we apply rule 1 and
\texttt{value(@6, 6, 6)} is updated to \texttt{value(@6, 6, 7)}.

\begin{figure*}[]
        \centering
        \begin{subfigure}[b]{0.45\textwidth}
                \includegraphics[width=\textwidth]{btree_trace1}
                \caption{Initial database. Replace axiom instantiated at root node $@3$.}
                \label{fig:btree_trace1}
        \end{subfigure}%
        ~
        \begin{subfigure}[b]{0.45\textwidth}
                \includegraphics[width=\textwidth]{btree_trace2}
                \caption{After applying rule 3 at node $@3$. Replace fact sent to node $@5$.}
                \label{fig:btree_trace2}
        \end{subfigure}\\
        \begin{subfigure}[b]{0.45\textwidth}
                \includegraphics[width=\textwidth]{btree_trace3}
                \caption{After applying rule 3 at node $@5$. Replace fact reaches node $@6$.}
                \label{fig:btree_trace3}
        \end{subfigure}%
        ~
        \begin{subfigure}[b]{0.45\textwidth}
                  \includegraphics[width=\textwidth]{btree_trace4}
                  \caption{After applying rule 1 at node $@6$. Value of key 6 has changed to 7.}
                  \label{fig:btree_trace4}
          \end{subfigure}
        \caption{An execution trace for the binary tree dictionary algorithm.}\label{fig:btree_trace}
\end{figure*}

\subsection{Syntax}

\renewcommand{\arraystretch}{1.2}
\begin{table}[h]
\centering
{\scriptsize
\begin{tabular}{ l l c l }
  Program & $Prog$ & $::=$ & $\Sigma, D$ \\
  Set Of Rules & $\Sigma$ & $::=$ & $\cdot \; | \; \Sigma, R$\\
  Database & $D$ & $::=$ & $\Gamma; \Delta$ \\
  Rule & $R$ & $::=$ & $BE \lolli HE \; | \; \forall_{x}. R$ \\
  Body Expression & $BE$ & $::=$ & $L \; | \; P \; | \; C \; | \; BE, BE \; | \; \exists_{x}. BE \; | \; 1$\\
  Head Expression & $HE$ & $::=$ & $L \; | \; P \; | \; HE, HE \; | \; CE \; | \; AE \; | \; 1$\\

  Linear Fact & $L$ & $::=$ & $l(\hat{x})$\\
  Persistent Fact & $P$ & $::=$ & $\bang p(\hat{x})$\\
  Constraint & $C$ & $::=$ & $c(\hat{x})$ \\

  Comprehension & $CE$ & $::=$ & $\comprehension$ \\
  Aggregate & $AE$ & $::=$ & $\aggregate$ \\
  Aggregate Operation & $A$ & $::=$ & $\mathtt{min} \; | \; \mathtt{max} \; | \; \mathtt{sum} \; | \; \mathtt{count}$ \\

  Sub-Head & $SH$ & $::=$ & $L \; | \; P \; | \; SH, SH \; | \; 1$\\

  Linear Facts & $\Delta$ & $::=$ & $\cdot \; | \; \Delta, l(\hat{t})$ \\
  Persistent Facts & $\Gamma$ & $::=$ & $\cdot \; | \; \Gamma, \bang p(\hat{t})$ \\
\end{tabular}
}
\caption{Abstract syntax of LM.}
\label{tbl:ast}
\end{table}
\renewcommand{\arraystretch}{1.0}

Table~\ref{tbl:ast} shows the abstract syntax for rules in LM.
A LM program $Prog$ consists of a set of derivation rules $\Sigma$ and a database $D$.
Each derivation rule $R$ can be written as $BE \lolli HE$ where $BE$ is the body of a rule and
$HE$ is the head. Rules without bodies are allowed in LM and they are called \textit{axioms}. Rules without heads are also allowed.
The body of a rule, $BE$, may contain linear ($L$) and persistent ($P$) \emph{fact expressions}
and constraints ($C$). Fact expressions are template facts that instantiate variables
(from facts in the database). Variables can be used again in the body for matching and
also in the head when instantiating facts. Constraints are boolean expressions that must
be true in order for the rule to be fired. Constraints use variables from fact expressions and are built using a small functional language that includes mathematical operations, boolean operations, external functions and literal values.
The head of a rule, $HE$, contains linear ($L$) and persistent ($P$) \emph{fact templates} which are uninstantiated facts to derive new facts. The head can also have \emph{comprehensions} ($CE$) and \emph{aggregates} ($AE$). Head
expressions may use the variables instantiated in the body.

\subsubsection{Comprehensions}

Sometimes we need to consume a linear fact and then immediately generate several facts depending on
the contents of the database. To solve this particular need, we created the concept of comprehensions, which are
sub-rules that are applied with all possible combinations of facts from the database. In a comprehension $\comprehension$,
$\widehat{x}$ is a list of variables, $BE$ is the body of the comprehension and $SH$ is the head.
The body $BE$ is used to generate all possible combinations for the head $SH$, according to the facts
in the database.

The following example illustrates a simple program that uses comprehensions:

{\footnotesize
\begin{Verbatim}
!edge(@1, @2).
!edge(@1, @3).
iterate(@1).
iterate(A) -o {B | !edge(A, B) | perform(B)}.
\end{Verbatim}
}

When the rule is fired, we consume \texttt{iterate(@1)} and then generate the comprehension. Here, we iterate through
all the \texttt{edge/2} facts that match \texttt{!edge(@1, B)}, which are: \texttt{!edge(@1, @2)} and \texttt{!edge(@1, @3)}.
For each fact, we derive \texttt{perform(B)}, namely: \texttt{perform(@2)} and \texttt{perform(@3)}.

\subsubsection{Aggregates}

Another useful feature in logic programs is the ability to reduce several facts into a single fact.
In LM we have aggregates ($AE$), a special kind of sub-rule that works very similarly to comprehensions.
In the abstract syntax of the aggregate $\aggregate$, $A$ is the aggregate operation, $\widehat{x}$ is the list of variables
introduced in $BE$, $SH_1$ and $SH_2$ and $y$ is the variable in the body
$BE$ that represents the values to be aggregated using $A$. Like comprehensions,
we use $\widehat{x}$ to try all the combinations of $BE$, but, in addition to deriving $SH_1$ for each combination,
we aggregate the values represented by $y$ and derive $SH_2$ only once using $y$.
As an example, consider the following program:

{\footnotesize
\begin{Verbatim}
price(@1, 3).
price(@1, 4).
price(@1, 5).
count-prices(@1).
count-prices(A) -o
   [sum => P | . | price(A, P) | 1 | total(A, P)].
\end{Verbatim}
}

By applying the rule, we consume \texttt{count-prices(@1)} and
derive the aggregate which consumes all the \texttt{price(@1, P)} facts.
These are summed and \texttt{total(@1,~12)} is derived.  
LM provides several aggregate operations, including the \emph{minimum}, \emph{maximum}, \emph{sum}, and \emph{count}.

\subsection{Concurrency}

LM is at its core a concurrent programming language.
The database of facts can be seen as a graph data structure where each node contains a fraction of the database.
To accomplish this, we force the first argument of each predicate to be typed as a \emph{node}. We then
restrict the derivation rules to only manipulate facts belonging to a single node.
However, the expressions in the head may refer to other nodes, as long as those nodes are instantiated in the body of the rule.

Due to the restrictions on LM rules, nodes are able to
run rules independently without using other node's facts. Node computation follows a
\emph{don't care} or \emph{committed choice} non-determinism
since any node can be picked to run as long as it contains enough facts to fire a derivation rule.
Facts coming from other nodes will arrive in order of derivation but may be considered
partially and there is no particular order among the neighborhood. To improve concurrency,
the programmer is encouraged to write rules that take advantage of the non-deterministic nature of execution.

### 2014-05-12 paper.tex
first commit

% This is LLNCS.DEM the demonstration file of
% the LaTeX macro package from Springer-Verlag
% for Lecture Notes in Computer Science,
% version 2.4 for LaTeX2e as of 16. April 2010
%
\documentclass{acm_proc_article-sp}
%
\usepackage{makeidx}  % allows for indexgeneration
\usepackage{fancyvrb}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{wrapfig}
\usepackage{nth}
\usepackage[backref,pageanchor=true,plainpages=false, pdfpagelabels, bookmarks,bookmarksnumbered]{hyperref}
\input{fp-macros}

\begin{document}
\newcommand{\cmu}{\ensuremath{^\dag}}
\newcommand{\fcup}{\ensuremath{^\ddag}}
  \makeatletter
    \let\@copyrightspace\relax
      \makeatother
\numberofauthors{3} %  in this sample file, there are a *total*

\title{Design and implementation of a multithreaded virtual machine for executing linear logic programs}

\author{
\alignauthor
   Flavio Cruz \\
\affaddr{Carnegie Mellon University, Pittsburgh, PA 15213, USA}\\
\email{fmfernan@cs.cmu.edu} \\
\alignauthor Ricardo Rocha \\
\affaddr{CRACS \& INESC TEC, Faculty of Sciences, University Of Porto} \\
\affaddr{Rua do Campo Alegre, 1021/1055, 4169-007 Porto, Portugal} \\
\email{ricroc@dcc.fc.up.pt} \\
\alignauthor Seth Copen Goldstein \\
\affaddr{Carnegie Mellon University, Pittsburgh, PA 15213, USA}\\
\email{seth@cs.cmu.edu} \\
}

%\institute{Carnegie Mellon University, Pittsburgh, PA 15213, USA\\
%\email{\{fmfernan, seth\}@cs.cmu.edu}
%\and
%CRACS \& INESC TEC, Faculty of Sciences, University Of Porto\\
%Rua do Campo Alegre, 1021/1055, 4169-007 Porto, Portugal\\
%\email{\{flavioc, ricroc\}@dcc.fc.up.pt}}

\maketitle              % typeset the title of the contribution

\begin{abstract}
\input{abstract}
\keywords{linear logic programming, virtual machine, implementation}
\end{abstract}

\section{Introduction}
\input{introduction}

\section{The LM Language}\label{lm_language}
\input{language}

\section{The Virtual Machine}\label{virtual_machine}
\input{virtual_machine}

\section{Experimental Results}\label{results}
\input{preliminary_results}

\section{Related Work}\label{related_work}
\input{related_work}

\section{Conclusions}
\input{conclusions}

\bibliographystyle{abbrv}
\bibliography{refs}
\end{document}

### 2014-05-12 preliminary_results.tex
first commit

This section presents preliminary results for our VM.
First, we present scalability results in order to show that LM programs can take advantage of multicore architectures.
Next, we present a comparison with similar programs written in other programming languages
in order to show evidence that our VM is viable.

For our experimental setup, we used a machine 
with two 16 (32) Core AMD Opteron
(tm) Processor 6274 $@$ 2.2 GHz with 32 GBytes of RAM memory and running the Linux
kernel 3.8.3-1.fc17.x86\_64.
     We compiled our VM using GCC 4.7.2 (g++) with the flags \texttt{-O3 -std=c+0x -march=x86-64}.
     We run all experiments 3 times and then averaged the execution time.

\subsection{Scalability Results}

For this section, we run each program using 1, 2, 4, 6, 8, 10, 12, 14 and 16 threads and compared the runtime against the execution of the sequential version of the VM. We used the following programs:

\newcommand{\figsize}[0]{6.5cm}
\captionsetup[sub]{              % The subcaption settings
       font=scriptsize}

\begin{itemize}
   \item Greedy Graph Coloring (GGC) colors nodes in a graph so that no two adjacent nodes have the same color. We start with a small number of colors and then we expand the number of colors when we cannot color the graph.
   \item PageRank implements a PageRank algorithm without synchronization between iterations. Every time a node sends a new rank to its neighbors and the change was significant, the neighbors are scheduled to recompute their ranks.
   \item N-Queens, the classic puzzle for a 13x13 board.
   \item Belief Propagation, a machine learning algorithm to denoise images.
\end{itemize}

Figure~\ref{exp:graph_coloring} presents the speedup results for the GGC program using 2 different datasets. In Fig.~\ref{exp:graph_coloring}a we show the speedup for a search engine graph of 12,000 webpages\footnote{Available from \url{http://www.cs.toronto.edu/~tsap/experiments/download/download.html}}. Since this dataset follows the power law, that is, there is a small number of pages with a lots of links (1\% of the nodes have 75\% of the edges), the speedup is slightly worse than the benchmark shown in Fig.~\ref{exp:graph_coloring}b, where we use a random dataset of 2,000 nodes with an uniform distribution of edges.

\newcommand{\plotsize}{0.38\textwidth}

\begin{figure*}[t]
   \centering
   \begin{subfigure}[b]{\plotsize}
      \includegraphics[width=\textwidth]{speedup_greedy-graph-coloring-search_engines.pdf}
      \caption{Using a graph of web pages with around 12,000 nodes and 292,000 edges.}
   \end{subfigure}
   ~~
   \begin{subfigure}[b]{\plotsize}
      \includegraphics[width=\textwidth]{speedup_greedy-graph-coloring-2000.pdf}
      \caption{Using a random graph with 2,000 nodes and 600,000 edges.}
   \end{subfigure}
   \caption{Experimental results for the GGC algorithm.}
   \label{exp:graph_coloring}
\end{figure*}

The PageRank results are shown in Fig.~\ref{exp:pagerank}. We used the same search engine dataset as before and a new random dataset with 5,000 nodes and 500,000 edges. Although the search engine graph (Fig.~\ref{exp:pagerank}a) has half the edges (around 250,000), it scales better than the random graph (Fig.~\ref{exp:pagerank}b), meaning that the PageRank program depends on the number of nodes to be more scalable.

\begin{figure*}[ht]
   \centering
   \begin{subfigure}[b]{\plotsize}
      \includegraphics[width=\textwidth]{speedup_pagerank-search_engines.pdf}
      \caption{Using a graph of web pages collected from a search engine (around 12,000 nodes and 292,000 edges)}
   \end{subfigure}
   ~~
   \begin{subfigure}[b]{\plotsize}
      \includegraphics[width=\textwidth]{speedup_pagerank-5000.pdf}
      \caption{Using a random graph with 5,000 nodes and 500,000 edges.}
   \end{subfigure}
   \caption{Experimental results for the asynchronous PageRank algorithm.}
   \label{exp:pagerank}
\end{figure*}

The results for the N-Queens program are shown in Fig.~\ref{exp:8queens}. The program is not regular since computation starts at the top of the grid and then rolls down, until only the last row be doing computation. Because the number of valid states for the nodes in the upper rows is much less than the nodes in the lower rows, this may potentially lead to
load balancing problems. The results show that our system is able to scale well due to work stealing.

\begin{figure*}[ht]
   \centering
   \begin{subfigure}[b]{\plotsize}
      \includegraphics[width=\textwidth]{speedup_8queens-13.pdf}
      \caption{N-Queens program (13x13~board).}
      \label{exp:8queens}
   \end{subfigure}
   ~~
   \begin{subfigure}[b]{\plotsize}
      \includegraphics[width=\textwidth]{speedup_bp-400.pdf}
      \caption{BP (400x400~image).}
      \label{exp:bp}
   \end{subfigure}
   \caption{Experimental Results for N-Queens and Belief Propagation.}
\end{figure*}

Finally, we shown the results for the Belief Propagation~(BP) program in Fig.~\ref{exp:bp}. BP is a regular and asynchronous program
and benefits (as expected) from having multiple threads executing since the belief values of each node will converge faster.
The super-linear results prove this assertion.

\subsection{Absolute Execution Time}

As we have seen, our VM scales reasonably well, but how does it compare in terms of absolute execution time
with other competing systems? We next present such comparison for the execution time using one thread.

In Fig.~\ref{comp:nqueens} we compare the LM's N-Queens version against 3 other versions: a straightforward sequential program implemented
in C using backtracking; a sequential Python~\cite{vanRossum:1995:PRM}
implementation; and a Prolog implementation executed in
YAP Prolog~\cite{DBLP:journals/corr/abs-1102-3896}, an efficient implementation of Prolog. Numbers less than 1 mean that LM
is faster and larger than 1 mean that LM is slower. We see that LM easily beats Python, but is 5 to 10 times slower than YAP
and around 15 times slower than C.
However, note that if we use at least 16 threads in LM, we can beat the sequential implementation written in C.

\begin{figure*}[ht]
   \centering

   \begin{subfigure}[b]{0.45\textwidth}
      \resizebox{7cm}{!} {
      \begin{tabular}{ | l | l | l | l |}
       \hline

       Size & C & Python & YAP Prolog \\ \hline\hline
       10 & 16.92 & \textbf{0,62} & 5,42 \\
       11 & 21.59 & \textbf{0.64} & 6.47 \\
       12 & 10.32 & \textbf{0.73} & 7.61 \\
       13 & 14.35 & \textbf{0.88} & 10.38 \\
       \hline
       \end{tabular}}
       \caption{N-Queens problem.}
      \label{comp:nqueens}
   \end{subfigure}
   ~~
   \begin{subfigure}[b]{0.45\textwidth}
      \resizebox{6.5cm}{!} {
      \begin{tabular}{ | l | l | l | l |}
       \hline

       Size & C & Python & GraphLab \\ \hline\hline
       10 & \textbf{0.67} & \textbf{0,03} & \textbf{1.00} \\
       50 & \textit{1.77} & \textbf{0.04} & \textit{1.73} \\
       200 & \textit{1.99} & \textbf{0.05} & \textit{1.79} \\
       400 & \textit{2.00} & \textbf{0.04} & \textit{1.80} \\
       \hline
       \end{tabular}}
       \caption{Belief Propagation program.}
       \label{comp:bp}
   \end{subfigure}
   \caption{Comparing absolute execution times (System Time / LM Time).}
\end{figure*}

In Fig.~\ref{comp:bp} we compare LM's Belief Propagation program against a sequential C version, a Python version and a GraphLab version. GraphLab~\cite{GraphLab2010} is a parallel C++ library used to solve graph-based problems in machine learning. C and GraphLab
perform about the same since both use C/C++. Python runs very slowly since it is a dynamic programming language and BP has many
mathematical computations. We should note, however, that the LM version uses some external functions written in C++ in order
to improve execution time, therefore the comparison is not totally fair.

We also compared the PageRank program against a similar GraphLab version and LM is around 4 to 6 times slower.
Finally, our version of the all-pairs shortest distance algorithm is 50 times slower than a C sequential implementation of the Dijkstra algorithm, but it is almost twice as fast when compared to the same implementation in Python.

### 2014-05-12 related_work.tex
first commit

Virtual machines are a popular technique for implementing interpreters for high level programming languages.
Due to the increased availability of parallel machines and distributed architectures, several machine
models have been developed with parallelism in mind~\cite{Kara:1997:AMM:265274}.
One example of such machine is the Parallel Virtual Machine (PVM)~\cite{Sunderam90pvm:a}, which serves as an abstraction
to program heterogeneous computers as a single machine. Another important machine is the Threaded Abstract Machine (TAM)~\cite{CullerGSvE93,goldstein-tr94},
which defines a self-scheduled machine language of parallel threads where a program is represented as conventional control flow.

Prolog, the most prominent logic programming language, has a rich history of virtual machine research centered
around the Warren Abstract Machine WAM~\cite{AICPub641:1983}. The WAM offers special purpose instructions, including
unification instructions for different kinds of data and control flow instructions to implement backtracking.
The WAM is fully sequential and uses four memory areas: heap, stack, trail and the push down list.
Much research has been done to improve the speed and efficiency of the original WAM design~\cite{Costa07demand-drivenindexing,167005,Turk-Logic:1986fk}.

Prolog is naturally parallel because several clauses for the same goal (AND-parallelism) or all goals in a clause (OR-parallelism) can be tried
in parallel. Different abstract machines for AND-parallelism has been developed on top of the WAM~\cite{Hermenegildo:1986:AMB:913061,Lin:1988:AEL:900478}.
For OR-parallelism we have several models such as: the SRI model~\cite{Warren:1987:OEM:67683.67699},
the Argonne model~\cite{ButlerDLOOS88}, the MUSE model~\cite{Ali:1990fk} and the BC machine~\cite{Ali88}. While all those models have been
developed using the WAM, there are some parallel machines totally different from the WAM such as the PPAM~\cite{Kacsuk:1990:EMP:533578},
which is based on a data-flow model.

### 2014-05-12 virtual_machine.tex
first commit

We developed a compiler that compiles LM programs to byte-code and a multi-threaded
virtual machine (VM) using the Pthreads library to run the byte-code.
The goal of our system is to keep the threads as busy as possible and to reduce inter-thread communication.

The load balancing aspect of the system is performed by our work scheduler that is based on a work
stealing algorithm. Initially, the system will partition the application
graph of $N$ nodes into $P$ subgraphs (the number of threads) and then each thread will work on their own subgraph.
During execution, threads can steal nodes of other threads to keep themselves busy.

Reduction of inter-thread communication is achieved by first ordering the node addresses present
in the code in such a way that closer nodes are clustered together and then partitioning them
to threads. During compilation, we take note of predicates that are used in rules for
communication rules (arguments with type \emph{node}) and then build a graph of nodes from
the program's axioms. The nodes of the graph are then ordered by using a breadth-first search
algorithm that changes the nodes of addresses to the domain $[0, n[$, where $n$ is the number of
nodes. Once the VM starts, we simply partition the range
$[0, n[$.

\subsection{Threads}

When the VM starts, it reads the byte-code file and starts all threads.
The partitioning of nodes is performed by dividing the domain of node addresses across the number of available threads.
In Fig.~\ref{fig:overview} we present the overview of a virtual machine running 2 threads and a program with 6 nodes.
Node that the dotted arrows represent the edges between nodes. In each thread space, we place the nodes owned by each thread.
Initially, the threads grab their own nodes and assign the \texttt{owner} property of each.
Because only one thread is allowed to do computation on a node at any giving time, the owner property
defines the thread with such permission.

\begin{figure*}[ht]
   \centering
   \scalebox{0.60}{\includegraphics[]{overview.pdf}}
   \caption{Overview of the virtual machine (showing 2 threads).}
   \label{fig:overview}
\end{figure*}

Included in the thread space is the \emph{Work Queue}, a single linked list containing \emph{active nodes}, i.e.,
nodes that have new facts to process. Initially, the \emph{Work Queue} is filled with all the nodes of the thread
in order to derive the axioms.

The main thread loop is shown in Fig.~\ref{code:work_loop}, where the threads
inspects their work queue for active nodes. Procedure \texttt{process\_node()} takes a node with new candidate rules
and executes them. If the work queue is empty, the thread attempts to steal one node from another thread before
becoming idle (\emph{Steal node} in Fig.~\ref{fig:overview}). Starting from a random thread, it cycles through all the threads to find one active node.
Eventually, there will be no more work to do and the threads will go idle. There is a global atomic counter, a global
boolean flag and one boolean flag for each thread that are used to detect termination.
Once a thread goes idle, it decrements the global counter and changes its flag to idle. If the counter
reaches zero, the global flag is set to idle. Since every thread will be busy-waiting and checking
the global flag, they will detect the change and stop executing.

\begin{figure}[h!]
\scriptsize\begin{Verbatim}
void work_loop(thread_id tid):
   while (true):
      current_node = NULL;
      if(has_work(tid)):
         current_node = pop_work(tid); // take node from the queue
      else:
         target_thread = random(NUM_THREADS);
         for (i = 0; i < NUM_THREADS && current_node == NULL; ++i):
            // need to steal a node
            target_thread = (target_thread + 1) % NUM_THREADS;
            current_node = steal_node_from_thread(target_thread)
      if(current_node == NULL):
         become_idle(tid);
         if(!synchronize_termination(tid)):
            return;
         become_active(tid);
      else:
         process_node(current_node, tid);
\end{Verbatim}
  \caption{Thread work loop.}
  \label{code:work_loop}
\end{figure}

\subsection{Nodes}

In Fig.~\ref{fig:overview}, we also present the internal structure of a node (node \texttt{@1}). Each node
contains the following: a database composed of \emph{Linear DB} (linear facts) and \emph{Persistent DB} (persistent facts) (further explained in \ref{sec:database});
rule matching structures (\emph{Rule Matching Structures}); and a \emph{Fact Buffer}, for storing intermediate facts coming
from other threads.

Whenever a new fact is derived through rule derivation, we need to update the data structures for the corresponding node.
This is trivial if the thread that owns the node derived the fact also. However, if that is not the case, then we have to synchronize
since multiple threads need to update the same data structures. We added a lock and a boolean flag to each node to protect the access to its data structures.
For example, in Fig.~\ref{fig:overview}, if thread 2 derives a fact to node \texttt{@1} (owned by thread 1), then thread 2 checks the node's flag
and if not activated, will lock node \texttt{@1} and then perform the required updates (\emph{New fact (1)}). If the flag is activated, it means
that thread 1 is currently executing the node, therefore it will not touch the main node data structures, but instead will add the new fact
to \emph{Fact Buffer} (\emph{New fact (2)}). The facts stored in \emph{Fact Buffer} will then be processed by thread 1 whenever the node is about to be executed.

Another thread interaction happens during fact derivation. In the example, after the fact is derived and sent to node \texttt{@1}, thread 2 also may need
to activate node \texttt{@1} by pushing it to the \emph{Work Queue} of thread 1 (\emph{Activate node}). Each node has another flag called \emph{In Queue} that indicates
if the node is currently placed in the \emph{Work Queue}. If the node is not in the queue, then we push it to the queue and then check if the target thread is currently idle.
After this synchronization point, the target thread will be active and with a new node to process.

\subsubsection{Database Data Structures}\label{sec:database}

We said before that LM rules are constrained by the first argument. Because nodes can execute
independently, our database is indexed by the node address and each sub-database does not
need to deal with synchronization issues since at any given point, only one thread will be using
the database. Note that the first argument of each fact is not stored.

The database must be implemented efficiently because during matching of rules we need
to restrict the facts using a given \emph{match object}, which fixes arguments of the target predicate to instantiated values.
Each sub-database is implemented using three kinds of data structures:

\begin{itemize}
   \item \emph{Trie Data Structures} are used exclusively to store persistent facts.
   Tries are trees where facts are indexed by the common arguments.

   \item \emph{Doubly Linked List Data Structures} are used to store linear facts.
   We use a double linked list because it is very efficient to add and remove facts.

   \item \emph{Hash Table Data Structures} are used to improve lookup when linked lists are too long and when we need to do search filtered by a fixed argument. The virtual machine decides which arguments are best to be indexed
   (see "Indexing") and then
   uses an hash table indexed by the appropriate argument. If we need to go through all the facts, we just iterate through all the facts in the table. For collisions, we use the above doubly linked list data structure.
\end{itemize}

\begin{figure}[]
   \centering
   \includegraphics[width=0.40\textwidth]{hash_table.pdf}
   \caption{\small{Hash table data structure for storing predicate \texttt{a(int,int)}.}}
   \label{fig:hash_table}
\end{figure}

Figure~\ref{fig:hash_table} shows an example for a hash table data structure
with 3 linear facts indexed by the second argument and stored as doubly linked list
in bucket \texttt{2}. Each linear fact contains the regular list pointers, a \texttt{flags} field
and the fact arguments. Those are all stored continuously to improve data
locality. One use of the \texttt{flags} field is to mark that a fact is already being used. For example,
consider the rule body \texttt{a(A,B), a(C,D) -o ...}. When we first pick a fact for \texttt{a(A, B)} from the hash table,
we mark it as being used in order to ensure that, when we retrieve facts for \texttt{a(C, D)}, the first one
cannot be used since that would violate linearity.

\subsubsection{Rule Engine}\label{rule_engine}

The rule engine decides which rules may need to be executed while taking into account rule priorities.
There are 5 main data structures for scheduling rule execution;
\texttt{Rule Queue} is the bitmap representing the rules that will be run;
\texttt{Active Bitmap} contains the rules that have enough facts to be fired;
\texttt{Inactive Bitmap} contains the rules that must be dropped from \texttt{Rule Queue};
\texttt{Predicates Bitmap} marks the newly derived facts;
and \texttt{Predicates Count} counts the number of facts per predicate.
To understand how our engine works, consider the following rules and axioms:

{\footnotesize\begin{Verbatim}
// rules
a, e(1) -o b.
a -o c.
b -o d.
e(0) -o f.
c -o e(1).

// axioms
a.
e(0).
\end{Verbatim}
}

\begin{figure*}[]
   \centering
   \scalebox{0.7}{\includegraphics[]{rule_queue1.pdf}}
   \caption{Program and rule engine data structures: before and after applying the \nth{2} rule.}
   \label{fig:rule_engine}
\end{figure*}

In our example, since we have facts \texttt{a} and \texttt{e(0)},
we will execute the second rule \texttt{a -o c}.
In order to pick such rule, we need to take the least significant bit from the \texttt{Rule Queue} bitmap.
Because the derivation is successful, we will consume \texttt{a} and derive \texttt{c}.
We thus mark the \texttt{c} predicate in the \texttt{Predicates Bitmap} and the first and second rules
in \texttt{Dropped Bitmap} since such rules are no longer applicable (\texttt{a} is gone). To update the \texttt{Rule Queue},
we remove the bits marked in \texttt{Dropped Bitmap} and add the active rules marked in \texttt{Active Bitmap} that are affected
by predicates in \texttt{Predicates Map}. The engine thus schedules the fourth and fifth rules to run (Fig.~\ref{fig:rule_engine}b).

Note that every node in the program has the same set of data structures present in Fig.~\ref{fig:rule_engine}.
We use 32 bits integers to implement bitmaps and an array of 16 bits integers to count facts, resulting in
$32 + 2P$ bytes per node, where $P$ is the number of predicates.

We do a small optimization to reduce the number of derivations of persistent facts. We
divide the program rules into two sets: \emph{persistent rules} and \emph{non persistent rules}.
Persistent rules are rules where only persistent facts are involved. We compile such rules
incrementally, that is, we attempt to fire all rules where a persistent fact is used. This is called
the \emph{pipelined semi-naive} evaluation and it originated in the P2 system~\cite{Loo-condie-garofalakis-p2}.
This evaluation method avoids excessing re-derivations of the same fact. The order of derivation does not matter for those rules, since
only persistent facts are used.

\subsection{Byte-Code}

A byte-code file contains meta-data about the program's predicates, initial nodes, partitioning
information, and code for each rule.
Each VM thread has 32 registers that are used during rule execution.
Registers can store facts, integers, floats, node addresses and pointers to runtime 
data structures (lists and structures). When registers store facts, we can reference
fields in the fact through the register.

Consider a rule \texttt{!a(X,Y), b(X,Z), c(X,Y) -o d(Y)} and a database with
\texttt{!a(1,2)}, \texttt{!a(2,3)}, \texttt{b(1,3)}, \texttt{b(5,3)}, \texttt{c(1,2)}, \texttt{c(1,3)},
\texttt{c(5,3)}. Rule execution proceeds in a series of recursive loops, as follows: the first loop retrieves an
iterator for the persistent facts of \texttt{!a/2} and moves the first valid fact, \texttt{!a(1,2)},
to register 0; the inner loop retrieves linear facts that match \texttt{b(1,Z)} (from the
\emph{join constraint}) and moves \texttt{b(1,3)} to register 1; in the final
loop we move \texttt{c(1,2)} to register 2 and the body of the rule is successfully matched. Next, we
derive \texttt{d(2)}, where \texttt{2} comes from register 0.
Fig.~\ref{fig:byte_code} shows the byte-code for this example.

\begin{figure}[]
\scriptsize\begin{Verbatim}
PERSISTENT ITERATE a MATCHING TO reg 0
  LINEAR ITERATE b MATCHING TO reg 1
      (match).0=0.0
    LINEAR ITERATE c MATCHING TO reg 2
        (match).0=0.0
        (match).1=0.1
      ALLOC d TO reg 3
      MVFIELDFIELD 0.1 TO 3.0
      ADDLINEAR reg 3
      REMOVE reg 2
      REMOVE reg 1
      TRY NEXT
    NEXT
  NEXT
RETURN
\end{Verbatim}
\caption{\small{Byte-code for rule \texttt{!a(X,Y), b(X,Z), c(X,Y) -o d(Y).}}}
\label{fig:byte_code}
\end{figure}

In case of failure, we jump to the previous outer loop in order to try the next candidate fact.
If a rule matches and the head is derived, we backtrack to the inner most \emph{valid loop}, i.e.,
the first inner loop that uses linear facts or, if there are no linear facts involved, to the previous 
inner loop. We need to jump to a valid loop because we may have loops with linear facts that are now invalid.
In our example, we would jump to the
loop of \texttt{b(X,Z)} and not \texttt{c(X,Y)}, since \texttt{b(1,3)} was consumed.

\begin{figure}[]
\scriptsize\begin{Verbatim}
LINEAR ITERATE a MATCHING TO reg 0
  MVFIELDREG 0.0 TO reg 1
  MVINTREG INT 1 TO reg 2
  reg 1 INT PLUS reg 2 TO reg 3
  MVREGFIELD reg 3 TO 0.0
  UPDATE reg 0
  TRY NEXT
RETURN
\end{Verbatim}
\caption{\small{Byte-code for rule \texttt{a(N) -o a(N+1)}.}}
\label{code:update}
\end{figure}

The compiler re-orders the fact expressions used in the body in order to make execution more
efficient. For example, it forces the join constraints in rules to appear at the beginning so
that matching will fail sooner rather than later. It also does the same for constraints.
Note that for every loop, the compiler adds the \emph{match object}, which contains information
about which arguments need to match, so that runtime matching is efficient.

Our compiler also detects cases where we re-derive a linear fact with new arguments.
For example, as shown in Fig.~\ref{code:update}, the rule \texttt{a(N) -o a(N+1)}
will compile to code that reuses the old \texttt{a(N)} fact.
We use a \texttt{flags} field to mark updated nodes (presented next).

\subsection{Indexing}\label{indexing}

To improve fact lookup,
the VM employs a fully dynamic mechanism to decide which argument may be optimal to index.
The algorithm is performed in the beginning of the execution and empirically tries to assess the argument
of each predicate that more equally spreads the database across the values of the argument.
A single thread performs the algorithm for all predicates.

The indexing algorithm is performed in three main steps. First, it gathers statistics of lookup data by keeping a counter
for each predicate's argument.
Every time a fact search is performed where arguments are fixed to a value, the counter of such arguments is incremented. This phase is performed during rule execution for a small fraction of the nodes in the program.

The second step of the algorithm then decides the candidate arguments of each predicate.
If a predicate was not searched with any fixed arguments, then it will be not indexed.
If only one argument was fixed, then such argument is set as the indexing argument. Otherwise, the top 2 arguments
are selected for the third phase, where \emph{entropy statistics} are collected dynamically.

During the third phase, each candidate argument has an entropy score.
Before a node is executed, the facts of the target predicate
are used in the following formula applied for the two arguments:

{\scriptsize
\[
Entropy(A, F) = - \sum_{v \in values(F, A)} \frac{count(F, A = v)}{total(F)} 	\log_2 \frac{count(F, A = v)}{total(F)}
\]
}

Where $A$ is the target argument, $F$ is the multi-set of linear facts for the target predicate, $values(F, A)$ is set of values of the argument $A$, $count(F, A = v)$ counts the number
of linear facts where argument $A$ is equal to $v$ and $total(F)$ counts the number of linear facts in $F$.
The entropy value is a good metric because it tells us how much information is needed to describe an argument.
If more information is needed, then that must be the best argument to index.

For one of the arguments to score, $Entropy(A, F)$ multiplied by the number of times it has been used for lookup must be larger than the other argument.

The argument with the best score is selected and then
a global variable called \texttt{indexing\_epoch} is updated.
In order to convert the node's linked lists into hash tables, each node also has a local variable called \texttt{indexing\_epoch}
that is compared to the global variable in order to rebuild the node database according to the new indexing
information.

Our VM also dynamically resizes the hash table if necessary. When the hash table becomes
too dense, it is resized to the double. When it becomes too sparse, it is reduced in half
or simply transformed back into a doubly linked list. This is done once in a while, before a node executes.

We have seen very good results with this scheme. For example, for the all-pairs shortest paths program,
we obtained a 2 to 5-fold improvement in sequential execution time.
The overhead of dynamic indexing is negligible since programs run almost as fast
as if the indices have been added from the start.

\subsection{Runtime Data Structures}

LM supports recursive types such as lists and pairs. These complex data structures are stored in
the heap of the VM and are managed through reference counting. For instance, each list
is a \emph{cons cell} with 3 fields: \texttt{tail}, the pointer to the next element of the list;
\texttt{head}, the element stored by this element of the list; and \texttt{refs} that counts the
number of pointers to this list element in the VM. The list is deleted from the heap whenever
\texttt{refs} is decremented to 0.

### 2014-07-25 abstract.tex
Update.

@@
-We have developed a virtual machine for Linear Meld, a forward-chaining linear logic programming language
-where programs can be executed in parallel. In Linear Meld, a program is seen as a database of logical facts and a set of derivation rules.
-The database of facts is partitioned by the nodes of a graph structure which leads
-to parallelism when nodes are executed simultaneously.
-Due to the foundations on linear logic, rules can retract facts in a declarative and structured fashion, leading
-to more expressive programs.
-We present the design and implementation of the virtual machine that executes on multicores,
-with a focus on thread management, code organization, fact indexing, rule execution, and database organization for efficient fact insertion,
-lookup and deletion. Our results show that the virtual machine is not only
-capable of scaling programs with up to 16 threads but also exhibits interesting performance results when
-compared against other programming languages.
+\begin{abstract}
+Linear Meld is a concurrent forward-chaining linear logic programming
+language where logical facts can be asserted and retracted in a
+structured way. In Linear Meld, a program is seen as a database of
+logical facts and a set of derivation rules. The database of facts is
+partitioned by the nodes of a graph structure which leads to
+parallelism when nodes are executed simultaneously. Due to the
+foundations on linear logic, rules can retract facts in a declarative
+and structured fashion, leading to more expressive programs. We
+present the design and implementation of the virtual machine that we
+implemented to run Linear Meld on multicores, with particular focus on
+thread management, code organization, fact indexing, rule execution,
+and database organization for efficient fact insertion, lookup and
+deletion. Our results show that the virtual machine is
+capable of scaling programs with up to 16 threads and also exhibits
+interesting scalar performance results due to our indexing optimizations.
+\end{abstract}
+
+\category{D.3.4}{PROCESSORS}{Interpreters}
+\category{D.3.4}{PROCESSORS}{Run-time environments}
+\category{D.1.3}{PROGRAMMING TECHNIQUES}{Concurrent Programming}[Parallel Programming]
+
+\terms{Design, Languages, Performance}
+
+\keywords{Linear Logic, Virtual Machine, Implementation}

### 2014-07-25 conclusions.tex
Update.

@@
-We have presented a parallel virtual machine for executing forward-chaining linear logic programs, with particular focus
-on parallel scheduling on multicores, rule execution and database organization for fast insertion,
-lookup, and deletion of linear facts.
-Our experimental results show that the VM is able to scale the execution of programs when run with up to 16 threads.
-Although this is still a work in progress, the VM also fairs well against other programming languages.
-Moreover, since LM programs are concurrent by default, we can easily get better performance from the start by executing
-them with multiple threads.
+\section{Conclusions}
 
-In the future, we intend to take advantage of linear logic to perform whole-program optimizations,
-including computing program invariants, loop detection in rules and bypass of rule priorities. For instance,
-we want to detect during compile time which might be the best data structure for each predicate depending in how it
-is used in the rules.
+We have presented a parallel virtual machine for executing
+forward-chaining linear logic programs, with particular focus on
+thread management, code organization, fact indexing, rule execution
+and database organization for fast insertion, lookup, and deletion of
+linear facts. Experimental results show that our VM is able to scale
+the execution of programs when run with up to 16 threads.
+Our results also show the importance of having an efficient indexing mechanisms
+for facts. With our dynamic indexing, the VM automatically detects which predicates
+need to be indexed in order to improve performance.
+Due to these and other optimizations, the VM fairs relatively well against other programming languages, including
+compiled languages. Moreover, since LM programs are
+concurrent by default, we can easily get better performance from the
+start by executing them with multiple threads.
+As further work, we want to improve parallel scalability and
+take advantage of linear logic to perform whole-program
+optimizations, including computing program invariants, loop detection
+in rules and bypass of rule priorities.
+
+We think that our virtual machine is an excellent starting point to make
+logic programming more desirable in the data-mining, machine learning and
+distributed/parallel programming community. Moreover, our virtual machine can
+be easily extended to execute over computer networks or to execute programs on really big datasets.
+In a nutshell, LM provides a concise way to describe
+graph-based algorithms that can be more easily reasoned about, a clear advantage over
+competing systems.
+%% For instance, we want to detect during compile time which might be the
+%% best data structure for each predicate depending in how it is used in
+%% the rules.

### 2014-07-25 experiments.tex
Update.

\newcommand{\plotsize}{0.45\textwidth}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Experimental Results}
\label{results}

This section presents initial results for our VM. First, we present a
comparison with similar programs written in other programming
languages in order to show evidence that our VM is viable. Then, we
present scalability results in order to show that LM programs can take
advantage of multicore architectures.

For our experimental setup, we used a machine with 32 (2x16) Core AMD
Opteron (tm) Processor 6274 $@$ 2.2 GHz with 32 GBytes of RAM memory
and running the Linux kernel 3.8.3-1.fc17.x86\_64.  We compiled our VM
using GCC 4.7.2 (g++) with the flags \texttt{-O3 -std=c+0x
  -march=x86-64}. We run all experiments 3 times and averaged the
execution time.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Absolute Execution Time}

To put our VM in perspective, we first compare it in terms of absolute
execution time with other competing systems using a single thread.

In Table~\ref{tbl:comp_nqueens}, we compare LM's version of the
classic N-Queens puzzle against 3 other versions: a straightforward
sequential program implemented in C using backtracking; a sequential
Python implementation~\cite{vanRossum:1995:PRM}; and a Prolog
implementation executed in YAP
Prolog~\cite{DBLP:journals/corr/abs-1102-3896}, an efficient
implementation of Prolog. Numbers less than 1 mean that LM is faster
and larger than 1 mean that LM is slower. We can observe that LM
easily beats Python, but is 5 to 10 times slower than YAP Prolog and
around 15 times slower than C. Note however that, as we will see next,
if we use at least 16 threads in LM, we can beat the sequential
implementation written in C.

\begin{table}[ht]
\centering
{\begin{tabular}{c|c|c|c}
\textbf{Problem} & \multicolumn{3}{c}{\textbf{System}} \\
\textbf{Size} & \textbf{C} & \textbf{Python} & \textbf{YAP Prolog} \\
\hline\hline
\textbf{10x10} & 16.92 & 0.62 & 5.42 \\
\textbf{11x11} & 21.59 & 0.64 & 6.47 \\
\textbf{12x12} & 10.32 & 0.73 & 7.61 \\
\textbf{13x13} & 14.35 & 0.88 & 10.38 \\
\end{tabular}}
\caption{Comparing the absolute execution times (LM/System) for the
  N-Queens program}
\label{tbl:comp_nqueens}
\end{table}

In Table~\ref{tbl:comp_bp}, we compare LM's Belief Propagation (BP)
program, a machine learning algorithm to denoise images, against a
sequential C, Python and GraphLab~\cite{GraphLab2010} version of the
algorithm. GraphLab is a parallel C++ library used to solve
graph-based problems in machine learning. C and GraphLab perform about
the same since both use C/C++. Python runs very slowly since it is a
dynamic programming language and BP has many mathematical
computations. We should note, however, that LM's version uses some
external functions written in C++ in order to improve execution time,
therefore the comparison is not totally fair.

\begin{table}[ht]
\centering
{\begin{tabular}{c|c|c|c}
\textbf{Problem} & \multicolumn{3}{c}{\textbf{System}} \\
\textbf{Size} & \textbf{C} & \textbf{Python} & \textbf{GraphLab} \\
\hline\hline
\textbf{10}  & 0.67 & 0.03 & 1.00 \\
\textbf{50}  & 1.77 & 0.04 & 1.73 \\
\textbf{200} & 1.99 & 0.05 & 1.79 \\
\textbf{400} & 2.00 & 0.04 & 1.80 \\
\end{tabular}}
\caption{Comparing the absolute execution times (LM/System) for the
  Belief Propagation program}
\label{tbl:comp_bp}
\end{table}

We also compared a LM's version of the PageRank program against a
similar GraphLab version and LM showed to be around 4 to 6 times
slower. Our worse results were obtained for the all-pairs shortest
distance algorithm where a LM's version of the problem was around 50
times slower than a C sequential implementation of the Dijkstra
algorithm, but almost twice as fast when compared to the same
implementation in Python.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Scalability}

\begin{figure*}[ht]
\centering
\subfigure[PageRank using a graph of web pages with around 12,000 nodes and 292,000 edges]
   {\label{exp:pagerank}\includegraphics[width=\plotsize]{figures/results-pagerank-search-engines.pdf}}
\hspace{0.5cm}
\subfigure[GGC using a random graph with 2,000 nodes and 600,000 edges]
   {\label{exp:ggc}\includegraphics[width=\plotsize]{figures/results-ggc.pdf}}
\caption{Experimental results for the GGC and PageRank algorithms}
\end{figure*}

\begin{figure*}[ht]
\centering
\subfigure[Shortest Distance for a graph with around 5,000 nodes and 13,000 edges]
   {\label{exp:sd}\includegraphics[width=\plotsize]{figures/results-shortest-uspowergrid.pdf}}
\hspace{0.5cm}
\subfigure[MiniMax algorithm for the Tic-Tac-Toe game (complete tree)]
   {\label{exp:minimax}\includegraphics[width=\plotsize]{figures/results-minimax.pdf}}
\caption{Experimental results for the Shortest Distance and MiniMax algorithm}
\end{figure*}

\begin{figure*}[ht]
\centering
\subfigure[N-Queens program (13x13~board)]
   {\label{exp:13queens}\includegraphics[width=\plotsize]{figures/results-13queens.pdf}}
\hspace{0.5cm}
\subfigure[BP (400x400~image)]
   {\label{exp:bp}\includegraphics[width=\plotsize]{figures/results-bp.pdf}}
\caption{Experimental results for the N-Queens and Belief Propagation programs}
\end{figure*}

In this section we measure the scalability of the VM along with the
performance gains due to work stealing and dynamic indexing.  For
this purpose, we used 4 configurations for the VM: \textbf{WI}, the
full configuration that includes work stealing and dynamic indexing;
\textbf{WN}, with work stealing but without dynamic
indexing; \textbf{NI}, with indexing but without work stealing; and
\textbf{NN}, without work stealing and without dynamic indexing.  We
ran each configuration using 1, 2, 4, 6, 8, 10, 12, 14 and 16 threads
and compared the run time against the sequential execution (1 thread)
of \textbf{WI}. We used the following set of programs:

\begin{itemize}
\item PageRank implements a PageRank algorithm without synchronization
  between iterations. Every time a node sends a new rank to its
  neighbors and the change was significant, the neighbors are
  scheduled to recompute their ranks.
\item Greedy Graph Coloring (GGC) colors nodes in a graph so that no
  two adjacent nodes have the same color. We start with a small number
  of colors and then we expand the number of colors when we cannot
  color the graph.
\item Shortest Distance (SD) computes the shortest distance of all
  nodes to all nodes.
\item MiniMax, the AI algorithm for selecting the best player move in a game of Tic-Tac-Toe.
\item N-Queens, the classic puzzle for a 13x13 board.
\item Belief Propagation, a machine learning algorithm to denoise
  images.
\end{itemize}

The PageRank results are shown in Fig.~\ref{exp:pagerank}. We used a
search engine graph of 12,000 webpages\footnote{Available from
  \url{http://www.cs.toronto.edu/~tsap/experiments/download/download.html}}. Since
this dataset follows the power law, that is, there is a small number
of pages with a lots of links (1\% of the nodes have 75\% of the
edges), it can be difficult to parallelize. Our results show that the
VM is able to scale the program with up to 14 threads.  We also notice
the huge performance drop when we run the VM without work
stealing. Dynamic indexing is also an advantage, since it detects that
the facts for the pagerank of neighboring nodes need to be indexed
efficiently.

Figure~\ref{exp:ggc} presents the results for the GGC program with a
random dataset of 2,000 nodes with an uniform distribution of edges.
There is a slight drop in scalability as the number of threads goes
up, but the VM is still capable of reducing the run time. We note that
in this program, the work available is reduced as the graph becomes
increasingly colored.

In Fig.~\ref{exp:sd} we show the results for the Shortest Distance program.
We attain a 13-fold speedup for 16 threads with both work stealing and dynamic
indexing (\textbf{WI}). We note that indexing is more advantageous than work stealing
because indexing the distance facts according to the source node is more crucial
than improved load balancing.

The results for the MiniMax algorithm are presented in Fig.~\ref{exp:minimax}.
MiniMax is very different than the other algorithms because the graph of nodes is dynamic
and is created during program execution. The load balancing is also problematic since there
is little work to do in the initial and final phases of the algorithm. Still, our VM has decent
performance, with almost a 7-fold speedup for 14 threads. The scalability drops with 16 threads
but we think that is due to the simplicity of the current work stealing algorithm.
Dynamic indexing has no effects in this program.

The results for the N-Queens program are shown in
Fig.~\ref{exp:13queens}. The program is not regular since computation
starts at the top of the grid and then rolls down, until only the last
row be doing computation. Because the number of valid states for the
nodes in the upper rows is much less than the nodes in the lower rows,
this may potentially lead to load balancing problems. The results show
that our system is able to scale well. When work stealing is left out
(\textbf{NI} and \textbf{NN}), we see a serious drop in performance with 16 threads.

Finally, we shown the results for the Belief Propagation~(BP) program
in Fig.~\ref{exp:bp}. BP is a regular and asynchronous program and
benefits (as expected) from having multiple threads executing since
the belief values of each node will converge faster.  The super-linear
results prove this assertion.

\subsection{Memory Statistics}

In this section, we present the amount of \textbf{Memory} used by the VM after the
programs presented in the previous section complete. We also count the number of
\textbf{Facts} stored in the database. This serves as an indication of
how much memory is needed in terms of the number of facts stored in the database.
Results are shown in Table~\ref{tbl:memory}.

\begin{table}[ht]
\centering
{\begin{tabular}{c|c|c|c}
\textbf{Program} & \textbf{Facts} & \textbf{Memory} & \textbf{Per Fact} \\
\hline\hline
PageRank & 1,180,603 & 203,832 KB & 176 bytes \\
GGC & 2,363,536 & 292,682 KB & 127 bytes \\
SD & 502,347 & 45,387 KB & 92 bytes \\
MiniMax & 3 & 886,443 KB & 295,481 KB \\
N-Queens & 74,557 & 55,408 KB & 760 bytes \\
BP & 2,235,200 & 617,417 KB & 283 bytes \\
\hline\hline
\multicolumn{3}{l}{Average (W/ MiniMax) } & 288 bytes \\
\end{tabular}}
\caption{Memory usage of programs after completion using 1 thread.}
\label{tbl:memory}
\end{table}

The most unexpected result in Table~\ref{tbl:memory} is that of the MiniMax program. There is a
very low number of facts (that indicate the final player decision) and a huge
amount of memory. How can this be explained? We note that MiniMax builds a huge
decision tree of nodes with almost $9!$ leaves. Although these nodes do not participate
in the computation after the MiniMax result is computed for them, the VM does not garbage collect
nodes that do not contain facts and that do not have external references from other nodes.
This is a potential improvement that can be done in the future.

PageRank, GGC and SD show a low memory usage per fact. This makes sense since the predicates
used in those programs are relatively simple with only a few arguments.
This changes with the N-Queens problem where the predicates have lists representing the valid
positions of the queens in the board. Storing lists is far more expensive than storing integral
values such as integers or floating point numbers. The same thing happens with BP because
the belief values are stored as lists of floating point numbers.

### 2014-07-25 introduction.tex
Update.

\section{Introduction}

The last decade has seen a tremendous growth in content available in the World Wide Web, and, more specifically,
in information generated from online social networks. The structure of such content is usually a graph, a very
flexible structure suited to represent content where pairs of items are linked.
In order to process such information, there has been an increased interest in running graph-based
algorithms concurrently and efficiently on top of distributed networks and computer architectures (multicores).
Currently available libraries and frameworks are built on top of imperative programming languages, which
require the programmer to know how to properly use the framework and the language. Reasoning about such
programs requires knowing the intricacies of the framework, how computation is scheduled and how processing
units coordinate between each other.

Some well known frameworks include Dryad, Pregel and GraphLab.
The Dryad system~\cite{Isard:2007:DDD:1272996.1273005} combines computational vertices
with communication channels (edges) to form a data-flow graph. The program is scheduled to
run on multiple processors or cores and data is partitioned during runtime. Routines that run on computational vertices
are sequential, with no synchronization.
The Pregel system~\cite{Malewicz:2010:PSL:1807167.1807184} is also graph based, although programs have a more strict
structure. They must be represented as a sequence of iterations where each iteration is composed of computation and message passing.
Pregel is specially suited to work on big graphs and to scale to large architectures.
GraphLab~\cite{GraphLab2010} is a C++ framework for developing parallel machine learning
algorithms. While Pregel uses message passing, GraphLab allows nodes to have read/write
access to different scopes through different concurrent access models in order to balance
performance and data consistency. While some programs only need to access the local node's
data, others may need to update edge information. Each consistency model will provide different
guarantees that are better adapted to some algorithms. GraphLab also provides different
schedulers that dictate the order in which node's are computed.

Logic programming is an attractive approach to the graph-based algorithms,
since logic-based languages provide a high-level, declarative approach
to programming. An important characteristic of logic programming is
that it offers great potential for implicit parallelism, thus making
logic programs much easier to parallelize than imperative
programs. First, logic programs are easier to reason about since they
are based on logical foundations. Second, logic programmers do not
need to use low level programming constructs such as locks or
semaphores to coordinate parallel execution, because logic systems
hide such details from the programmer.

\iffalse
Logic programming languages split into two main
groups: \emph{backwards-chaining} and \emph{forward-chaining}. In a
backwards-chaining logic programming language, programs are composed
of a set of rules that can be activated by inputting a query. Given a
query $q(\hat{x})$, an interpreter will work backwards by matching
$q(\hat{x})$ against the head of a rule. If found, the interpreter
will then try to match the body of the rule, recursively, until it
finds the program axioms (rules without body). If the search procedure
succeeds, the interpreter finds a valid substitution for the $\hat{x}$
variables. A popular backwards-chaining programming language is
Prolog~\cite{Colmerauer:1993:BP:154766.155362}, which has been a
productive research language for executing logic programs in
parallel. Researchers took advantage of Prolog's non-determinism to
evaluate subgoals in parallel with models such
as \emph{And-parallelism}
and \emph{Or-parallelism}~\cite{Gupta:2001:PEP:504083.504085}.
\fi

In forward-chaining logic programming, programs start with
a database of facts (filled with the program's axioms) and a set of
logical rules. The database of facts is then used to fire the
program's rules and derive new facts that are then added to the
database. This process is repeated recursively until the database
reaches \emph{quiescence} and no more information can be derived from
the program. A popular forward-chaining programming language is
Datalog~\cite{Ramakrishnan93asurvey}.

We have designed Linear Meld~(LM), a forward-chaining logic programming
language that is specially suited for
concurrent programming over graph structures~\cite{cruz-iclp14}. LM
differs from Datalog-like languages because it integrates both
classical logic and linear logic~\cite{girard-87} into the language, allowing some
facts to be retracted and asserted in a logical fashion. Although most
Datalog and Prolog-like programming languages allow some kind of state
manipulation~\cite{Liu98extendingdatalog}, those features are
extra-logical, reducing the advantages brought forward by logic programming.

The roots of LM are the P2 system~\cite{Loo-condie-garofalakis-p2} and
the original Meld
language~\cite{ashley-rollman-derosa-iros07wksp,ashley-rollman-iclp09}. P2
is a Datalog-like language that maps a computer network to a graph,
where each computer node can perform computations locally and
communicate with neighbors. Meld is inspired by the P2 system
but adapted to the concept of massively distributed systems made of
modular robots with a dynamic topology. LM still follows the same
graph model of computation of Meld, which makes LM programs naturally
concurrent since the graph of nodes can be easily partitioned to be
executed by different threads. As a forward-chaining linear logic
programming language, LM also shares similarities with Constraint
Handling Rules
(CHR)~\cite{Betz:2005kx,DBLP:journals/corr/abs-1006-3039}.  CHR is a
concurrent committed-choice constraint language used to write
constraint solvers. A CHR program is a set of rules and a set of
constraints (which can be seen as facts). Constraints can be consumed
or generated during the application of rules. Some optimizations
used in LM such as join optimizations and the use of different data
structures for indexing facts were inspired by research done in
CHR~\cite{DBLP:journals/corr/cs-PL-0408025}.

In this paper, we present the design and implementation of the LM virtual machine
and compiler.
The LM virtual machine was designed from scratch to run LM programs on
multicore machines in an efficient manner\footnote{Source code available at \url{http://github.com/flavioc/meld}}
The virtual machine is multithreaded and executes byte-code that is generated
by the LM compiler. To test our language and virtual machine we have implemented several graph algorithms, search
algorithms and machine learning algorithms, including: belief
propagation~\cite{Gonzalez+al:aistats09paraml}, belief propagation
with residual splash~\cite{Gonzalez+al:aistats09paraml}, PageRank,
graph coloring, N-Queens, shortest path, diameter estimation, map
reduce, quick-sort, neural network training, minimax, etc.

Our results show that our virtual machine is scalable and presents some interesting
execution times when compared with other competing systems.
The virtual machine uses a simple, but effective, work stealing algorithm
that is able to balance the load across threads, improving scalability.
Another important feature of our virtual machine is the dynamic indexing algorithm.
It is a runtime algorithm that decides how to index logical facts and which
data structure is best to use, so that database lookup and insertion time during rule application
is effectively reduced.

This paper is organized as follows. First, we briefly introduce the LM
language. Then, we present an overview of the virtual machine and
describe in more detail the code organization, thread management, rule
execution and database organization. Finally, we present our experiments
and outline some conclusions.

### 2014-07-25 language.tex
update figs.
Update.

@@
 
-Linear Meld (LM) is a logic programming language that offers a declarative and structured way to manage state.
-A program consists of a database of facts and a set of derivation rules. The database includes persistent
-and linear facts. Persistent facts cannot be deleted, while linear facts can be asserted and retracted.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-The dynamic (or operational) semantics of LM are identical to Datalog.
-Initially, we populate the database with the program's axiom facts and then determine which derivation rules can be applied by using the current database.
-Once a rule is applied, we derive new facts, which are then added to the database.
-If a rule uses linear facts, they are retracted from the database.
-The program stops when \emph{quiescence} is achieved, that is, when rules no longer apply.
+\section{The LM Language}
+\label{lm_language}
 
-Each fact is a predicate on a tuple of \emph{values}, where the type of the predicate prescribes the types of the arguments.
-LM rules are type-checked using the predicate declarations in the header of the program. LM has a simple type system that includes types such as
-\emph{node}, \emph{int}, \emph{float}, \emph{string}, \emph{bool}. Recursive types such as \emph{list X} and \emph{pair X; Y} are
-also allowed.
-Each rule in LM has a defined priority that is inferred from its position in the source file.
-Rules at the beginning of the file have higher priority. We consider all
-the new facts that have been not considered yet to create a set of \emph{candidate rules}.
-The set of candidate rules is then applied (by priority) and updated as new facts are derived.
+Linear Meld (LM) is a logic programming language that offers a
+declarative and structured way to manage mutable state.  A program consists of
+a database of facts and a set of derivation rules. The database
+includes persistent and linear facts. Persistent facts cannot be
+deleted, while linear facts can be asserted and retracted.
+
+The dynamic (or operational) semantics of LM is identical to
+Datalog. Initially, we populate the database with the \emph{program's axioms} (initial facts)
+and then determine which derivation rules can be applied by
+using the current database. Once a rule is applied, new facts can be
+derive, which are then added to the database. If a rule uses linear
+facts, they are retracted from the database. The program stops when
+\emph{quiescence} is achieved, i.e., when rules no longer apply.
+
+Each fact is a predicate on a tuple of \emph{values}, where the type
+of the predicate prescribes the types of the arguments.  LM rules are
+type-checked using the predicate declarations in the header of the
+program. LM has a simple type system that includes types such as
+\emph{node}, \emph{int}, \emph{float}, \emph{string},
+\emph{bool}. Recursive types such as \emph{list X} and \emph{pair X;
+  Y} are also allowed.  Each rule in LM has a defined priority that is
+inferred from its position in the source file.  Rules at the beginning
+of the file have higher priority.
+We consider all the new facts that have been not used yet to create a
+set of \emph{candidate rules}.
+The set of candidate rules is then applied (by priority)
+and updated as new facts are derived.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 type right(node, node).
-type linear value(node, int, int).
-type linear replace(node, int, int).
+type linear value(node, int, string).
+type linear replace(node, int, string).
 
+// set of rules
 replace(A, K, New),
@@
 !left(A, B),
-RKey < Key // go left
+RKey < Key
    -o value(A, Key, Value),
-      replace(B, RKey, RValue).
+      replace(B, RKey, RValue). // go left
 
@@
 !right(A, B),
-RKey > Key // go right
+RKey > Key
    -o value(A, Key, Value),
-      replace(B, RKey, RValue).
+      replace(B, RKey, RValue). // go right
 
-// binary tree configuration
-value(@3, 3, 3). value(@1, 1, 1). value(@0, 0, 0).
-value(@2, 2, 2). value(@5, 5, 5). value(@4, 4, 4).
-value(@6, 6, 6).
-!left(@1, @0). !left(@3, @1). !left(@5, @4). 
-!right(@1, @2). !right(@3, @5). !right(@5, @6).
+// initial configuration
+!left(@0, @1).   !right(@0, @2).
+!left(@1, @3).   !right(@1, @4). 
+!left(@2, @5).   !right(@2, @6).
 
-replace(@3, 6, 7). // replace value of key 6 to 7
+value(@0, 3, a).   value(@1, 1, b).
+value(@2, 5, c).   value(@3, 0, d).
+value(@4, 2, e).   value(@5, 4, f).
+value(@6, 6, g).
+
+// update key 6 to value x
+replace(@0, 6, x).
 \end{Verbatim}
 }
-\caption{Binary tree dictionary: replacing a key's value.}
+\caption{LM program for replacing a key's value in a binary tree dictionary}
 \label{code:btree_replace}
@@
 
-We now present an example LM program in Fig.~\ref{code:btree_replace} that implements the key update operation for a binary tree
-represented as a key/value dictionary.
-We first declare all the predicates (lines 1-4), which represent the kinds of facts we are going to use.
-Predicate \texttt{left/2} and \texttt{right/2} are persistent while \texttt{value/3} and \texttt{replace/3} are linear.
-The \texttt{value/3} predicate assigns a key/value pair to a binary tree node and the \texttt{replace/3} predicate
-represents an update operation that updates the key in the second argument to the value in the third argument.
+We now present in Fig.~\ref{code:btree_replace}, a LM program that
+implements the update operation for a binary tree dictionary
+represented as key/value pairs. We first declare the predicates (lines
+1-4) which represent the facts we are going to use. Predicate
+\texttt{left/2} and \texttt{right/2} are persistent while predicates
+\texttt{value/3} and \texttt{replace/3} are linear. Predicate
+\texttt{value/3} assigns a key/value pair to a tree node and predicate
+\texttt{replace/3} represents an update operation that updates the key
+in the second argument to the value in the third argument.
 
-The algorithm uses three rules for the three cases of updating a key's value: the first rule performs the update
-(lines 6-8); the second rule recursively picks the left branch for the update operation (lines 10-15); and the third
-rule picks the right branch (lines 17-22).
-The axioms of this program are presented in lines 24-29 and they describe the initial binary tree configuration,
-including keys and values. By having the \texttt{update(@3, 6, 7)} axiom instantiated at the root node $@3$, we intend to
-change the value of key 6 to 7.
-Note that when writing rules or axioms, persistent facts are preceded with a \texttt{!}.
+The algorithm uses three rules for the three cases of updating a key's
+value: the first rule performs the update (lines 6-9); the second rule
+recursively picks the left branch for the update operation (lines
+11-16); and the third rule picks the right branch (lines 18-23). The
+initial axioms are presented in lines 26-33 and they describe the
+initial binary tree configuration, including keys and values.
+Finally, with the \texttt{replace(@0, 6, x)} axiom instantiated at the
+root node $@0$, we intend to change the value of key 6 to value
+`$x$'. Note that when writing rules or axioms, persistent facts are
+preceded with a \texttt{!}.
 
-Figure~\ref{fig:btree_trace} represents the trace of the algorithm. Note that the program database is partitioned
-by the tree nodes using the first argument of each fact. In Fig.~\ref{fig:btree_trace}a we present the database
-filled with the program's axioms. Next, we follow the right branch using rule 3 since $6 > 3$ (Fig.~\ref{fig:btree_trace}b).
-We then use the same rule again in Fig.~\ref{fig:btree_trace}c where we finally reach the key 6. Here, we apply rule 1 and
-\texttt{value(@6, 6, 6)} is updated to \texttt{value(@6, 6, 7)}.
+\begin{figure*}[t]
+\centering
+\subfigure[Initial database]
+   {\label{fig:btree_trace1}\includegraphics[width=0.44\textwidth]{figures/btree_trace1.pdf}}
+\subfigure[After applying rule 3 at node $@0$]
+   {\label{fig:btree_trace2}\includegraphics[width=0.44\textwidth]{figures/btree_trace2.pdf}}
+\subfigure[After applying rule 3 at node $@2$]
+   {\label{fig:btree_trace3}\includegraphics[width=0.44\textwidth]{figures/btree_trace3.pdf}}
+\subfigure[After applying rule 1 at node $@6$]
+   {\label{fig:btree_trace4}\includegraphics[width=0.44\textwidth]{figures/btree_trace4.pdf}}
+\caption{An execution trace for the binary tree dictionary program}
+\label{fig:btree_trace}
+\end{figure*}
 
-\begin{figure*}[]
-        \centering
-        \begin{subfigure}[b]{0.45\textwidth}
-                \includegraphics[width=\textwidth]{btree_trace1}
-                \caption{Initial database. Replace axiom instantiated at root node $@3$.}
-                \label{fig:btree_trace1}
-        \end{subfigure}%
-        ~
-        \begin{subfigure}[b]{0.45\textwidth}
-                \includegraphics[width=\textwidth]{btree_trace2}
-                \caption{After applying rule 3 at node $@3$. Replace fact sent to node $@5$.}
-                \label{fig:btree_trace2}
-        \end{subfigure}\\
-        \begin{subfigure}[b]{0.45\textwidth}
-                \includegraphics[width=\textwidth]{btree_trace3}
-                \caption{After applying rule 3 at node $@5$. Replace fact reaches node $@6$.}
-                \label{fig:btree_trace3}
-        \end{subfigure}%
-        ~
-        \begin{subfigure}[b]{0.45\textwidth}
-                  \includegraphics[width=\textwidth]{btree_trace4}
-                  \caption{After applying rule 1 at node $@6$. Value of key 6 has changed to 7.}
-                  \label{fig:btree_trace4}
-          \end{subfigure}
-        \caption{An execution trace for the binary tree dictionary algorithm.}\label{fig:btree_trace}
-\end{figure*}
+Figure~\ref{fig:btree_trace} illustrates the trace of the
+execution. Note that the program database is partitioned by the tree
+nodes using the first argument of each fact. In
+Fig.~\ref{fig:btree_trace1}, we present the database filled with the
+program's axioms. Next, we follow the right branch using rule 3 since
+$6 > 3$ (Fig.~\ref{fig:btree_trace2}). We then use the same rule
+again in Fig.~\ref{fig:btree_trace3} where we finally reach the key
+6. Here, we apply rule 1 and \texttt{value(@6, 6, g)} is updated to
+\texttt{value(@6, 6, x)} (Fig.~\ref{fig:btree_trace4}).
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-\renewcommand{\arraystretch}{1.2}
-\begin{table}[h]
+Table~\ref{tbl:ast} shows the abstract syntax for rules in LM.  An LM
+program $Prog$ consists of a set of derivation rules $\Sigma$ and a
+database $D$.  Each derivation rule $R$ can be written as $BE \lolli
+HE$ where $BE$ is the body of a rule and $HE$ is the head. Rules
+without bodies are allowed in LM and they are called
+\textit{axioms}. Rules without heads are also allowed.  The body of a
+rule, $BE$, may contain linear ($L$) and persistent ($P$) \emph{fact
+  expressions} and constraints ($C$). Fact expressions are template
+facts that instantiate variables (from facts in the
+database). Variables can be used again in the body for matching and
+also in the head when instantiating facts. Constraints are boolean
+expressions that must be true in order for the rule to be
+fired. Constraints use variables from fact expressions and are built
+using a small functional language that includes mathematical
+operations, boolean operations, external functions and literal values.
+The head of a rule, $HE$, contains linear ($L$) and persistent ($P$)
+\emph{fact templates} which are uninstantiated facts to derive new
+facts. Head expressions may use the variables instantiated in the
+body. The head can also have \emph{comprehensions} ($CE$) and
+\emph{aggregates} ($AE$).
+
+\begin{table}[ht]
 \centering
 {\scriptsize
-\begin{tabular}{ l l c l }
+\begin{tabular}{llcl}
 Program          & $Prog$ & $::=$ & $\Sigma, D$ \\
-  Set Of Rules & $\Sigma$ & $::=$ & $\cdot \; | \; \Sigma, R$\\
+Set of Rules     & $\Sigma$ & $::=$ & $\cdot \; | \; \Sigma, R$\\
 Database         & $D$ & $::=$ & $\Gamma; \Delta$ \\
@@
 Head Expression  & $HE$ & $::=$ & $L \; | \; P \; | \; HE, HE \; | \; CE \; | \; AE \; | \; 1$\\
-
 Linear Fact      & $L$ & $::=$ & $l(\hat{x})$\\
@@
 Constraint       & $C$ & $::=$ & $c(\hat{x})$ \\
-
 Comprehension    & $CE$ & $::=$ & $\comprehension$ \\
 Aggregate        & $AE$ & $::=$ & $\aggregate$ \\
-  Aggregate Operation & $A$ & $::=$ & $\mathtt{min} \; | \; \mathtt{max} \; | \; \mathtt{sum} \; | \; \mathtt{count}$ \\
-
+Operations       & $A$ & $::=$ & $\mathtt{min} \; | \; \mathtt{max} \; | \; \mathtt{sum} \; | \; \mathtt{count}$ \\
 Sub-Head         & $SH$ & $::=$ & $L \; | \; P \; | \; SH, SH \; | \; 1$\\
-
 Linear Facts     & $\Delta$ & $::=$ & $\cdot \; | \; \Delta, l(\hat{t})$ \\
@@
 }
-\caption{Abstract syntax of LM.}
+\caption{Abstract syntax of LM}
 \label{tbl:ast}
 \end{table}
-\renewcommand{\arraystretch}{1.0}
 
-Table~\ref{tbl:ast} shows the abstract syntax for rules in LM.
-A LM program $Prog$ consists of a set of derivation rules $\Sigma$ and a database $D$.
-Each derivation rule $R$ can be written as $BE \lolli HE$ where $BE$ is the body of a rule and
-$HE$ is the head. Rules without bodies are allowed in LM and they are called \textit{axioms}. Rules without heads are also allowed.
-The body of a rule, $BE$, may contain linear ($L$) and persistent ($P$) \emph{fact expressions}
-and constraints ($C$). Fact expressions are template facts that instantiate variables
-(from facts in the database). Variables can be used again in the body for matching and
-also in the head when instantiating facts. Constraints are boolean expressions that must
-be true in order for the rule to be fired.
-Constraints use variables from fact expressions and are built using a small functional language that includes mathematical operations, boolean operations, external functions and literal values.
-The head of a rule, $HE$, contains linear ($L$) and persistent ($P$) \emph{fact templates} which are uninstantiated facts to derive new facts.
-The head can also have \emph{comprehensions} ($CE$) and \emph{aggregates} ($AE$).
-Head
-expressions may use the variables instantiated in the body.
-
-\subsubsection{Comprehensions}
-
-Sometimes we need to consume a linear fact and then immediately generate several facts depending on
-the contents of the database. To solve this particular need, we created the concept of comprehensions, which are
-sub-rules that are applied with all possible combinations of facts from the database. In a comprehension $\comprehension$,
-$\widehat{x}$ is a list of variables, $BE$ is the body of the comprehension and $SH$ is the head.
-The body $BE$ is used to generate all possible combinations for the head $SH$, according to the facts
-in the database.
-
-The following example illustrates a simple program that uses comprehensions:
+We created the concept of comprehensions to be used when the
+consumption of a linear fact should generate a set of facts
+accordingly to the current contents of the database. In a
+comprehension $\comprehension$, $\widehat{x}$ is a list of variables,
+$BE$ is the body of the comprehension and $SH$ is the head. The body
+$BE$ is used to generate all possible combinations for the head $SH$,
+according to the facts in the database. The following program shows a
+simple example that uses comprehensions:
 
@@
 \begin{Verbatim}
-!edge(@1, @2).
-!edge(@1, @3).
+!edge(@1, @2).   !edge(@1, @3).
 iterate(@1).
-iterate(A) -o {B | !edge(A, B) | perform(B)}.
+
+iterate(A)
+   -o {B | !edge(A, B) | perform(B)}.
 \end{Verbatim}
@@
 
-When the rule is fired, we consume \texttt{iterate(@1)} and then generate the comprehension. Here, we iterate through
-all the \texttt{edge/2} facts that match \texttt{!edge(@1, B)}, which are: \texttt{!edge(@1, @2)} and \texttt{!edge(@1, @3)}.
-For each fact, we derive \texttt{perform(B)}, namely: \texttt{perform(@2)} and \texttt{perform(@3)}.
+When the rule is fired, we consume \texttt{iterate(@1)} and then
+generate the comprehension. Here, we iterate through all the
+\texttt{edge/2} facts that match \texttt{!edge(@1, B)}, which are
+\texttt{!edge(@1, @2)} and \texttt{!edge(@1, @3)}. For each fact, we
+then derive \texttt{perform(B)}, namely \texttt{perform(@2)} and
+\texttt{perform(@3)} in this example.
 
-\subsubsection{Aggregates}
-
-Another useful feature in logic programs is the ability to reduce several facts into a single fact.
-In LM we have aggregates ($AE$), a special kind of sub-rule that works very similarly to comprehensions.
-In the abstract syntax of the aggregate $\aggregate$, $A$ is the aggregate operation, $\widehat{x}$ is the list of variables
-introduced in $BE$, $SH_1$ and $SH_2$ and $y$ is the variable in the body
-$BE$ that represents the values to be aggregated using $A$. Like comprehensions,
-we use $\widehat{x}$ to try all the combinations of $BE$, but, in addition to deriving $SH_1$ for each combination,
-we aggregate the values represented by $y$ and derive $SH_2$ only once using $y$.
-As an example, consider the following program:
+Another useful feature is the ability to reduce several facts into a
+single fact. For that, we have aggregates, a special kind of sub-rule
+that works very similarly to comprehensions. In a aggregate
+$\aggregate$, $A$ is the aggregate operation, $\widehat{x}$ is the
+list of variables introduced in $BE$, $SH_1$ and $SH_2$ and $y$ is the
+variable in the body $BE$ that represents the values to be aggregated
+using $A$. Like comprehensions, we use $\widehat{x}$ to try all the
+combinations of $BE$, but, in addition to deriving $SH_1$ for each
+combination, we aggregate the values represented by $y$ and derive
+$SH_2$ only once using $y$. As an example, consider the following
+program:
 
@@
 \begin{Verbatim}
-price(@1, 3).
-price(@1, 4).
-price(@1, 5).
+price(@1, 3).   price(@1, 4).   price(@1, 5).
 count-prices(@1).
-count-prices(A) -o
-   [sum => P | . | price(A, P) | 1 | total(A, P)].
+
+count-prices(A)
+   -o [sum => P | . | price(A, P) | 1 | total(A, P)].
 \end{Verbatim}
@@
 
-By applying the rule, we consume \texttt{count-prices(@1)} and
-derive the aggregate which consumes all the \texttt{price(@1, P)} facts.
-These are summed and \texttt{total(@1,~12)} is derived.  
-LM provides several aggregate operations, including the \emph{minimum}, \emph{maximum}, \emph{sum}, and \emph{count}.
+By applying the rule, we consume \texttt{count-prices(@1)} and derive
+the aggregate which consumes all the \texttt{price(@1, P)} linear
+facts. These are summed up and \texttt{total(@1,~12)} is derived. LM
+provides several aggregate operations, including the \emph{minimum},
+\emph{maximum}, \emph{sum} and \emph{count}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-LM is at its core a concurrent programming language.
-The database of facts can be seen as a graph data structure where each node contains a fraction of the database.
-To accomplish this, we force the first argument of each predicate to be typed as a \emph{node}. We then
-restrict the derivation rules to only manipulate facts belonging to a single node.
-However, the expressions in the head may refer to other nodes, as long as those nodes are instantiated in the body of the rule.
+LM is at its core a concurrent programming language. The database of
+facts can be seen as a graph data structure where each node contains a
+fraction of the database.  To accomplish this, we force the first
+argument of each predicate to be typed as a \emph{node}. We then
+restrict the derivation rules to only manipulate facts belonging to a
+single node.  However, the expressions in the head may refer to other
+nodes, as long as those nodes are instantiated in the body of the
+rule.
 
-Due to the restrictions on LM rules, nodes are able to
-run rules independently without using other node's facts. Node computation follows a
-\emph{don't care} or \emph{committed choice} non-determinism
-since any node can be picked to run as long as it contains enough facts to fire a derivation rule.
-Facts coming from other nodes will arrive in order of derivation but may be considered
-partially and there is no particular order among the neighborhood. To improve concurrency,
-the programmer is encouraged to write rules that take advantage of the non-deterministic nature of execution.
+Due to the restrictions on LM rules, nodes are able to run rules
+independently without using other node's facts. Node computation
+follows a \emph{don't care} or \emph{committed choice} non-determinism
+since any node can be picked to run as long as it contains enough
+facts to fire a derivation rule.  Facts coming from other nodes will
+arrive in order of derivation but may be considered partially and
+there is no particular order among the neighborhood. To improve
+concurrency, the programmer is encouraged to write rules that take
+advantage of the non-deterministic nature of execution.

### 2014-07-25 paper.tex
Update.

@@
-% This is LLNCS.DEM the demonstration file of
-% the LaTeX macro package from Springer-Verlag
-% for Lecture Notes in Computer Science,
-% version 2.4 for LaTeX2e as of 16. April 2010
-%
 \documentclass{acm_proc_article-sp}
-%
-\usepackage{makeidx}  % allows for indexgeneration
+\usepackage[utf8]{inputenc}
+\usepackage{epsfig}
+\usepackage{multicol}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{subfigure}
 \usepackage{fancyvrb}
-\usepackage{amsmath}
-\usepackage{graphicx}
-\usepackage{caption}
-\usepackage{subcaption}
-\usepackage{wrapfig}
-\usepackage{nth}
-\usepackage[backref,pageanchor=true,plainpages=false, pdfpagelabels, bookmarks,bookmarksnumbered]{hyperref}
+\usepackage{url}
+\usepackage{balance}
+\usepackage{listings}
+\usepackage{multicol}
+\lstset{language=C, basicstyle=\ttfamily, breaklines=true, frame=lines}
 \input{fp-macros}
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \begin{document}
-\newcommand{\cmu}{\ensuremath{^\dag}}
-\newcommand{\fcup}{\ensuremath{^\ddag}}
-  \makeatletter
-    \let\@copyrightspace\relax
-      \makeatother
-\numberofauthors{3} %  in this sample file, there are a *total*
 
-\title{Design and implementation of a multithreaded virtual machine for executing linear logic programs}
+\title{Design and Implementation of a Multithreaded Virtual Machine for Executing Linear Logic Programs}
 
+\numberofauthors{3}
 \author{
-\alignauthor
-   Flavio Cruz \\
-\affaddr{Carnegie Mellon University, Pittsburgh, PA 15213, USA}\\
+  \alignauthor Flavio Cruz \\
+    \affaddr{Carnegie Mellon University}
+    \affaddr{Pittsburgh, PA 15213, USA}\\
     \email{fmfernan@cs.cmu.edu} \\
   \alignauthor Ricardo Rocha \\
-\affaddr{CRACS \& INESC TEC, Faculty of Sciences, University Of Porto} \\
-\affaddr{Rua do Campo Alegre, 1021/1055, 4169-007 Porto, Portugal} \\
-\email{ricroc@dcc.fc.up.pt} \\
+    \affaddr{CRACS \& INESC TEC}\\
+    \affaddr{University of Porto}\\
+    \affaddr{Rua Campo Alegre 1021/1055}\\
+    \affaddr{4169-007 Porto, Portugal}\\
+    \email{ricroc@dcc.fc.up.pt}
   \alignauthor Seth Copen Goldstein \\
-\affaddr{Carnegie Mellon University, Pittsburgh, PA 15213, USA}\\
+    \affaddr{Carnegie Mellon University}
+    \affaddr{Pittsburgh, PA 15213, USA}\\
     \email{seth@cs.cmu.edu} \\
@@
 
-%\institute{Carnegie Mellon University, Pittsburgh, PA 15213, USA\\
-%\email{\{fmfernan, seth\}@cs.cmu.edu}
-%\and
-%CRACS \& INESC TEC, Faculty of Sciences, University Of Porto\\
-%Rua do Campo Alegre, 1021/1055, 4169-007 Porto, Portugal\\
-%\email{\{flavioc, ricroc\}@dcc.fc.up.pt}}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\maketitle              % typeset the title of the contribution
+\maketitle
 
-\begin{abstract}
 \input{abstract}
-\keywords{linear logic programming, virtual machine, implementation}
-\end{abstract}
+\input{introduction}
+\input{language}
+\input{virtual_machine}
+\input{experiments}
+\input{related_work}
+\input{conclusions}
 
-\section{Introduction}
-\input{introduction}
+\balance
 
-\section{The LM Language}\label{lm_language}
-\input{language}
-
-\section{The Virtual Machine}\label{virtual_machine}
-\input{virtual_machine}
-
-\section{Experimental Results}\label{results}
-\input{preliminary_results}
-
-\section{Related Work}\label{related_work}
-\input{related_work}
-
-\section{Conclusions}
-\input{conclusions}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2014-07-25 related_work.tex
Update.

@@
-Virtual machines are a popular technique for implementing interpreters for high level programming languages.
-Due to the increased availability of parallel machines and distributed architectures, several machine
-models have been developed with parallelism in mind~\cite{Kara:1997:AMM:265274}.
-One example of such machine is the Parallel Virtual Machine (PVM)~\cite{Sunderam90pvm:a}, which serves as an abstraction
-to program heterogeneous computers as a single machine. Another important machine is the Threaded Abstract Machine (TAM)~\cite{CullerGSvE93,goldstein-tr94},
-which defines a self-scheduled machine language of parallel threads where a program is represented as conventional control flow.
+\section{Related Work}
+\label{related_work}
 
-Prolog, the most prominent logic programming language, has a rich history of virtual machine research centered
-around the Warren Abstract Machine WAM~\cite{AICPub641:1983}. The WAM offers special purpose instructions, including
-unification instructions for different kinds of data and control flow instructions to implement backtracking.
-The WAM is fully sequential and uses four memory areas: heap, stack, trail and the push down list.
-Much research has been done to improve the speed and efficiency of the original WAM design~\cite{Costa07demand-drivenindexing,167005,Turk-Logic:1986fk}.
+Virtual machines are a popular technique for implementing interpreters
+for high level programming languages.  Due to the increased
+availability of parallel machines and distributed architectures,
+several machine models have been developed with parallelism in
+mind~\cite{Kara:1997:AMM:265274}.  One example of such machine is the
+Parallel Virtual Machine (PVM)~\cite{Sunderam90pvm:a}, which serves as
+an abstraction to program heterogeneous computers as a single
+machine. Another important machine is the Threaded Abstract Machine
+(TAM)~\cite{CullerGSvE93,goldstein-tr94}, which defines a
+self-scheduled machine language of parallel threads where a program is
+represented as conventional control flow.
 
-Prolog is naturally parallel because several clauses for the same goal (AND-parallelism) or all goals in a clause (OR-parallelism) can be tried
-in parallel. Different abstract machines for AND-parallelism has been developed on top of the WAM~\cite{Hermenegildo:1986:AMB:913061,Lin:1988:AEL:900478}.
-For OR-parallelism we have several models such as: the SRI model~\cite{Warren:1987:OEM:67683.67699},
-the Argonne model~\cite{ButlerDLOOS88}, the MUSE model~\cite{Ali:1990fk} and the BC machine~\cite{Ali88}. While all those models have been
-developed using the WAM, there are some parallel machines totally different from the WAM such as the PPAM~\cite{Kacsuk:1990:EMP:533578},
-which is based on a data-flow model.
+Prolog, the most prominent logic programming language, has a rich
+history of virtual machine research centered around the Warren
+Abstract Machine (WAM)~\cite{AICPub641:1983}. 
+%% The WAM offers special purpose instructions, including unification
+%% instructions for different kinds of data and control flow instructions
+%% to implement backtracking.  The WAM is fully sequential and uses four
+%% memory areas: heap, stack, trail and the push down list.  Much
+%% research has been done to improve the speed and efficiency of the
+%% original WAM
+%% design~\cite{Costa07demand-drivenindexing,167005,Turk-Logic:1986fk}.
+Prolog is naturally parallel because several clauses for the same goal
+(AND-parallelism) or all goals in a clause (OR-parallelism) can be
+tried in parallel. Different abstract machines for AND-parallelism has
+been developed on top of the
+WAM~\cite{Hermenegildo:1986:AMB:913061,Lin:1988:AEL:900478}.  For
+OR-parallelism we have several models such as: the SRI
+model~\cite{Warren:1987:OEM:67683.67699}, the Argonne
+model~\cite{ButlerDLOOS88}, the MUSE model~\cite{Ali:1990fk} and the
+BC machine~\cite{Ali88}. While all those models have been developed
+using the WAM, there are also some parallel machines totally different
+from the WAM, such as the PPAM~\cite{Kacsuk:1990:EMP:533578}, which is
+based on a data-flow model.

### 2014-07-25 virtual_machine.tex
Update.

@@
-We developed a compiler that compiles LM programs to byte-code and a multi-threaded
-virtual machine (VM) using the Pthreads library to run the byte-code.
-The goal of our system is to keep the threads as busy as possible and to reduce inter-thread communication.
+\section{The Virtual Machine}
+\label{virtual_machine}
 
-The load balancing aspect of the system is performed by our work scheduler that is based on a work
-stealing algorithm. Initially, the system will partition the application
-graph of $N$ nodes into $P$ subgraphs (the number of threads) and then each thread will work on their own subgraph.
-During execution, threads can steal nodes of other threads to keep themselves busy.
+We have developed a compiler that compiles LM programs to byte-code
+and a multi-threaded virtual machine (VM) using POSIX threads
+that runs the byte-code.
 
-Reduction of inter-thread communication is achieved by first ordering the node addresses present
-in the code in such a way that closer nodes are clustered together and then partitioning them
-to threads. During compilation, we take note of predicates that are used in rules for
-communication rules (arguments with type \emph{node}) and then build a graph of nodes from
-the program's axioms. The nodes of the graph are then ordered by using a breadth-first search
-algorithm that changes the nodes of addresses to the domain $[0, n[$, where $n$ is the number of
-nodes. Once the VM starts, we simply partition the range
-$[0, n[$.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-When the VM starts, it reads the byte-code file and starts all threads.
-The partitioning of nodes is performed by dividing the domain of node addresses across the number of available threads.
-In Fig.~\ref{fig:overview} we present the overview of a virtual machine running 2 threads and a program with 6 nodes.
-Node that the dotted arrows represent the edges between nodes. In each thread space, we place the nodes owned by each thread.
-Initially, the threads grab their own nodes and assign the \texttt{owner} property of each.
-Because only one thread is allowed to do computation on a node at any giving time, the owner property
-defines the thread with such permission.
+A key goal of our design is to keep the threads as busy as possible
+and to reduce inter-thread communication. When the VM starts, it reads
+the byte-code file and starts all threads. Initially, the VM will
+partition the application graph of $N$ nodes into $T$ subgraphs (the
+number of threads) and then each thread will work on their own
+subgraph. Reduction of inter-thread communication is achieved by first
+ordering the node addresses present in the code in such a way that
+closer nodes are clustered together and then partitioning them to
+threads. During compilation, we take note of predicates that are used
+in rules with communication (arguments with type \emph{node}) and
+then build a graph of nodes from the program's axioms. The nodes of
+the graph are then ordered by using a breadth-first search algorithm
+that changes the nodes of addresses to the domain $[0, n[$, where $n$
+is the number of nodes. Once the VM starts, we simply partition the
+range $[0, n[$ into $T$ subgraphs.
 
-\begin{figure*}[ht]
+During execution, threads can steal nodes of other threads to keep
+themselves busy. The load balancing aspect of the system is performed
+by our work scheduler that is based on a simple work stealing
+algorithm. The pseudo-code for the main thread loop is shown in
+Fig.~\ref{code:work_loop}. In each round, a thread inspects its work
+queue for active nodes with new candidate rules and, if there is any,
+procedure \texttt{process\_node()} executes it. If the work queue is
+empty, the thread attempts to steal one node from another
+thread. Starting from a random thread, it cycles through all the
+threads to find one active node. Eventually, there will be no more
+work to do and the threads will go idle. There is a global atomic
+counter, a global boolean flag and one boolean flag for each thread
+that are used to detect termination. Once a thread goes idle, it
+decrements the global counter and changes its flag to idle. If the
+counter reaches zero, the global flag is set to idle. Since every
+thread will be busy-waiting and checking the global flag, they will
+detect the change and stop executing.
+
+\begin{figure}[ht]
+{\footnotesize
+\begin{Verbatim}
+void thread_work_loop(thread_id tid)
+   while (true)
+      node = pop_node_from_work_queue(tid)
+      if (node)
+         process_node(tid, node)
+      else
+         // need to steal a node
+         target = random(NUM_THREADS)
+         for (i = 0; i < NUM_THREADS && !node; i++)
+            target = (target + 1) % NUM_THREADS
+            node = steal_node_from_thread(target)
+         if (!node)
+            // try to terminate
+            become_idle(tid)
+            if (!synchronize_termination(tid))
+               return
+            become_active(tid)
+\end{Verbatim}
+}
+\caption{Thread work loop}
+\label{code:work_loop}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Nodes}
+
+\begin{figure*}[t]
 \centering
-   \scalebox{0.60}{\includegraphics[]{overview.pdf}}
-   \caption{Overview of the virtual machine (showing 2 threads).}
+\includegraphics[width=\textwidth]{figures/overview.pdf}
+\caption{Layout of the virtual machine}
 \label{fig:overview}
@@
 
-Included in the thread space is the \emph{Work Queue}, a single linked list containing \emph{active nodes}, i.e.,
-nodes that have new facts to process. Initially, the \emph{Work Queue} is filled with all the nodes of the thread
-in order to derive the axioms.
+Figure~\ref{fig:overview} presents the layout of our virtual machine
+for a program with six nodes and two running threads. Each thread
+space includes the nodes owned by the thread (the dotted arrows
+represent the edges between nodes) and a \emph{Work Queue}, a single
+linked list containing \emph{active nodes}, i.e., nodes that have new
+facts to process. Initially, the \emph{Work Queue} is filled with all
+the nodes of the thread in order to derive the
+axioms. Figure~\ref{fig:overview} also illustrates the internal
+structure layout of a node, which includes: the database of linear
+facts (\emph{Linear DB}); the database of persistent facts
+(\emph{Persistent DB}); the rule matching structures (\emph{Rule
+Engine}); and a auxiliary buffer for storing intermediate facts
+coming from other threads (\emph{Fact Buffer}).
 
-The main thread loop is shown in Fig.~\ref{code:work_loop}, where the threads
-inspects their work queue for active nodes. Procedure \texttt{process\_node()} takes a node with new candidate rules
-and executes them. If the work queue is empty, the thread attempts to steal one node from another thread before
-becoming idle (\emph{Steal node} in Fig.~\ref{fig:overview}). Starting from a random thread, it cycles through all the threads to find one active node.
-Eventually, there will be no more work to do and the threads will go idle. There is a global atomic counter, a global
-boolean flag and one boolean flag for each thread that are used to detect termination.
-Once a thread goes idle, it decrements the global counter and changes its flag to idle. If the counter
-reaches zero, the global flag is set to idle. Since every thread will be busy-waiting and checking
-the global flag, they will detect the change and stop executing.
+Whenever a new fact is derived through rule derivation, we need to
+update the data structures for the corresponding node. This is trivial
+if the thread that derived the fact also owns the node. If that is not
+the case, then we have to synchronize since multiple threads might be
+updating the same node's data structures. We added a lock and a
+boolean flag to each node to protect the access to its data
+structures. When the flag is activated, it means that the node is
+currently being executed by the owning thread. For example, in
+Fig.~\ref{fig:overview}, if thread 2 derives a fact to
+node \texttt{@1} (owned by thread 1), then thread 2 checks the node's
+flag and if not activated, will lock node \texttt{@1} and perform the
+required updates (\emph{New fact (1)}). If the flag is activated, it
+will not touch the main node data structures, but instead will add the
+new fact to \emph{Fact Buffer} (\emph{New fact (2)}). The facts stored
+in \emph{Fact Buffer} will then be processed whenever the
+corresponding node is about to be executed.
 
-\begin{figure}[h!]
-\scriptsize\begin{Verbatim}
-void work_loop(thread_id tid):
-   while (true):
-      current_node = NULL;
-      if(has_work(tid)):
-         current_node = pop_work(tid); // take node from the queue
-      else:
-         target_thread = random(NUM_THREADS);
-         for (i = 0; i < NUM_THREADS && current_node == NULL; ++i):
-            // need to steal a node
-            target_thread = (target_thread + 1) % NUM_THREADS;
-            current_node = steal_node_from_thread(target_thread)
-      if(current_node == NULL):
-         become_idle(tid);
-         if(!synchronize_termination(tid)):
-            return;
-         become_active(tid);
-      else:
-         process_node(current_node, tid);
-\end{Verbatim}
-  \caption{Thread work loop.}
-  \label{code:work_loop}
-\end{figure}
+There is another thread interaction that might happen during fact
+derivation if the node receiving a new fact is not active. In such
+case, the sending thread needs to activate the node by pushing it to
+the \emph{Work Queue} of the target thread. For example, consider
+again the situation in which thread 2 sends a new fact to
+node \texttt{@1}. If node \texttt{@1} is not active, then thread 2
+also needs to activate it by pushing it to the \emph{Work Queue} of
+thread 1. After this synchronization point, if the target thread is
+currently idle, it will become active and with a new node to process.
 
-\subsection{Nodes}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-In Fig.~\ref{fig:overview}, we also present the internal structure of a node (node \texttt{@1}). Each node
-contains the following: a database composed of \emph{Linear DB} (linear facts) and \emph{Persistent DB} (persistent facts) (further explained in \ref{sec:database});
-rule matching structures (\emph{Rule Matching Structures}); and a \emph{Fact Buffer}, for storing intermediate facts coming
-from other threads.
+\subsection{Database Data Structures}
+\label{sec:database}
 
-Whenever a new fact is derived through rule derivation, we need to update the data structures for the corresponding node.
-This is trivial if the thread that owns the node derived the fact also. However, if that is not the case, then we have to synchronize
-since multiple threads need to update the same data structures. We added a lock and a boolean flag to each node to protect the access to its data structures.
-For example, in Fig.~\ref{fig:overview}, if thread 2 derives a fact to node \texttt{@1} (owned by thread 1), then thread 2 checks the node's flag
-and if not activated, will lock node \texttt{@1} and then perform the required updates (\emph{New fact (1)}). If the flag is activated, it means
-that thread 1 is currently executing the node, therefore it will not touch the main node data structures, but instead will add the new fact
-to \emph{Fact Buffer} (\emph{New fact (2)}). The facts stored in \emph{Fact Buffer} will then be processed by thread 1 whenever the node is about to be executed.
+We said before that LM rules are constrained by the first argument and
+that each node has its own database of linear and persistent
+facts. Moreover, since only one thread at a time will be using a
+node's database, we do not need to deal with synchronization
+issues. Note also that the first argument of each fact is not stored.
 
-Another thread interaction happens during fact derivation. In the example, after the fact is derived and sent to node \texttt{@1}, thread 2 also may need
-to activate node \texttt{@1} by pushing it to the \emph{Work Queue} of thread 1 (\emph{Activate node}). Each node has another flag called \emph{In Queue} that indicates
-if the node is currently placed in the \emph{Work Queue}. If the node is not in the queue, then we push it to the queue and then check if the target thread is currently idle.
-After this synchronization point, the target thread will be active and with a new node to process.
-
-\subsubsection{Database Data Structures}\label{sec:database}
-
-We said before that LM rules are constrained by the first argument. Because nodes can execute
-independently, our database is indexed by the node address and each sub-database does not
-need to deal with synchronization issues since at any given point, only one thread will be using
-the database. Note that the first argument of each fact is not stored.
-
-The database must be implemented efficiently because during matching of rules we need
-to restrict the facts using a given \emph{match object}, which fixes arguments of the target predicate to instantiated values.
-Each sub-database is implemented using three kinds of data structures:
+The database must be implemented efficiently because during matching
+of rules we need to restrict the facts using a given \emph{match
+object}, which fixes arguments of the target predicate to instantiated
+values. Each node's database is implemented using three kinds of data
+structures:
 
 \begin{itemize}
-   \item \emph{Trie Data Structures} are used exclusively to store persistent facts.
-   Tries are trees where facts are indexed by the common arguments.
-
-   \item \emph{Doubly Linked List Data Structures} are used to store linear facts.
-   We use a double linked list because it is very efficient to add and remove facts.
-
-   \item \emph{Hash Table Data Structures} are used to improve lookup when linked lists are too long and when we need to do search filtered by a fixed argument.
-The virtual machine decides which arguments are best to be indexed
-   (see "Indexing") and then
-   uses an hash table indexed by the appropriate argument.
-If we need to go through all the facts, we just iterate through all the facts in the table.
-For collisions, we use the above doubly linked list data structure.
+\item \emph{Trie Data Structures} are used exclusively to store 
+  persistent facts. Tries are trees where facts are indexed by common
+  prefix arguments.
+\item \emph{Doubly Linked List Data Structures} are used to store 
+  linear facts. We use a double linked list because it is a very
+   efficient way to add and remove facts.
+\item \emph{Hash Table Data Structures} are used to improve lookup when 
+  linked lists are too long and when we need to do search filtered by
+  a fixed argument. The virtual machine decides which arguments are
+  best to be indexed (see Section~\ref{indexing}) and then uses a hash table
+  indexed by the appropriate argument. If we need to go through all
+  the facts, we just iterate through all the facts in the table. For
+  collisions, we use the above doubly linked list data structure.
 \end{itemize}
 
-\begin{figure}[]
+Figure~\ref{fig:hash_table} shows an example for a hash table data
+structure for a \texttt{a(int,int)} predicate with 3 linear facts
+indexed by the second argument and stored as a doubly linked list in
+bucket \texttt{2}. Each linear fact contains the regular list
+pointers, a \texttt{flags} field and the fact arguments. Those are all
+stored continuously to improve data locality. One use of
+the \texttt{flags} field is to mark that a fact is already being
+used. For example, consider the rule \texttt{a(A,B), a(C,D) -o
+...}. When we first pick a fact for \texttt{a(A, B)} from the hash
+table, we mark it as being used in order to ensure that, when we
+retrieve facts for \texttt{a(C, D)}, the first one cannot be used
+twice since that would violate linearity.
+
+\begin{figure}[ht]
 \centering
-   \includegraphics[width=0.40\textwidth]{hash_table.pdf}
-   \caption{\small{Hash table data structure for storing predicate \texttt{a(int,int)}.}}
+\includegraphics[width=0.4\textwidth]{figures/hash_table.pdf}
+\caption{Hash table and doubly linked data structures for 
+  a \texttt{a(int,int)} predicate}
 \label{fig:hash_table}
@@
 
-Figure~\ref{fig:hash_table} shows an example for a hash table data structure
-with 3 linear facts indexed by the second argument and stored as doubly linked list
-in bucket \texttt{2}. Each linear fact contains the regular list pointers, a \texttt{flags} field
-and the fact arguments. Those are all stored continuously to improve data
-locality. One use of the \texttt{flags} field is to mark that a fact is already being used. For example,
-consider the rule body \texttt{a(A,B), a(C,D) -o ...}. When we first pick a fact for \texttt{a(A, B)} from the hash table,
-we mark it as being used in order to ensure that, when we retrieve facts for \texttt{a(C, D)}, the first one
-cannot be used since that would violate linearity.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\subsubsection{Rule Engine}\label{rule_engine}
+\subsection{Rule Engine}
+\label{rule_engine}
 
-The rule engine decides which rules may need to be executed while taking into account rule priorities.
-There are 5 main data structures for scheduling rule execution;
-\texttt{Rule Queue} is the bitmap representing the rules that will be run;
-\texttt{Active Bitmap} contains the rules that have enough facts to be fired;
-\texttt{Inactive Bitmap} contains the rules that must be dropped from \texttt{Rule Queue};
-\texttt{Predicates Bitmap} marks the newly derived facts;
-and \texttt{Predicates Count} counts the number of facts per predicate.
-To understand how our engine works, consider the following rules and axioms:
+The rule engine decides which rules may need to be executed while
+taking into account rule priorities. Figure~\ref{fig:rule_engine}
+shows the rule engine data structures in more detail. There are 4 main
+data structures for scheduling rule execution: \texttt{Rule Queue} is
+the bitmap representing the rules scheduled to run; \texttt{Active
+Bitmap} contains the rules that have enough facts to be
+fired; \texttt{Dropped Bitmap} contains the rules that must be dropped
+from \texttt{Rule Queue} and \texttt{Active Bitmap} if the rule being
+executed succeeds; and \texttt{Predicates Count} counts the number of
+facts per predicate. To understand how our engine works, consider the
+following set of rules and facts:
 
-{\footnotesize\begin{Verbatim}
+{\footnotesize
+\begin{Verbatim}
 // rules
@@
 
-// axioms
+// facts
 a.
@@
 
-\begin{figure*}[]
+\begin{figure*}[t]
 \centering
-   \scalebox{0.7}{\includegraphics[]{rule_queue1.pdf}}
-   \caption{Program and rule engine data structures: before and after applying the \nth{2} rule.}
+\includegraphics[width=11cm]{figures/rule_queue.pdf}
+\caption{Rule engine data structures (a) before and (b) after applying 
+  the rule \texttt{a -o c.}}
 \label{fig:rule_engine}
@@
 
-In our example, since we have facts \texttt{a} and \texttt{e(0)},
-we will execute the second rule \texttt{a -o c}.
-In order to pick such rule, we need to take the least significant bit from the \texttt{Rule Queue} bitmap.
-Because the derivation is successful, we will consume \texttt{a} and derive \texttt{c}.
-We thus mark the \texttt{c} predicate in the \texttt{Predicates Bitmap} and the first and second rules
-in \texttt{Dropped Bitmap} since such rules are no longer applicable (\texttt{a} is gone). To update the \texttt{Rule Queue},
-we remove the bits marked in \texttt{Dropped Bitmap} and add the active rules marked in \texttt{Active Bitmap} that are affected
-by predicates in \texttt{Predicates Map}. The engine thus schedules the fourth and fifth rules to run (Fig.~\ref{fig:rule_engine}b).
+Since we have facts for predicates \texttt{a/0} and \texttt{e/1},
+the \texttt{Active Bitmap} starts with rules 1, 2 and 4 marked as
+having enough facts to be fired. The \texttt{Rule Queue} bitmap also
+starts with the same three rules. In order to pick rules for
+execution, we take the rule corresponding to the least significant bit
+from the \texttt{Rule Queue} bitmap, initially the first
+rule \texttt{a, e(1) -o b}. However, since we don't have
+fact \texttt{e(1)}, this rule fails and we execute the second
+rule \texttt{a -o c}. Figure~\ref{fig:rule_engine}(a) shows the rule
+engine data structures at that point.
 
-Note that every node in the program has the same set of data structures present in Fig.~\ref{fig:rule_engine}.
-We use 32 bits integers to implement bitmaps and an array of 16 bits integers to count facts, resulting in
-$32 + 2P$ bytes per node, where $P$ is the number of predicates.
+Because the derivation for the second rule succeeds, we will consume
+fact \texttt{a} and derive fact \texttt{c}. We thus update
+\texttt{Predicates Count} accordingly, mark the first 
+and second rules in \texttt{Dropped Bitmap} since such rules are no
+longer applicable (\texttt{a} was consumed) and mark the fifth rule
+in \texttt{Active Bitmap} since \texttt{c} was derived. Finally, to
+update the \texttt{Rule Queue}, we remove the bits marked
+in \texttt{Dropped Bitmap} and add the active rules marked
+in \texttt{Active Bitmap} that use the newly derived predicates, rule
+5 in this case. In the continuation, the engine will schedule the
+fourth and fifth rules to run. Figure~\ref{fig:rule_engine}(b) shows
+the rule engine data structures at that point.
 
-We do a small optimization to reduce the number of derivations of persistent facts. We
-divide the program rules into two sets: \emph{persistent rules} and \emph{non persistent rules}.
-Persistent rules are rules where only persistent facts are involved. We compile such rules
-incrementally, that is, we attempt to fire all rules where a persistent fact is used. This is called
-the \emph{pipelined semi-naive} evaluation and it originated in the P2 system~\cite{Loo-condie-garofalakis-p2}.
-This evaluation method avoids excessing re-derivations of the same fact. The order of derivation does not matter for those rules, since
-only persistent facts are used.
+Note that every node in the program has the same set of data
+structures presented in Fig.~\ref{fig:rule_engine}. We use 32 bits
+integers to implement bitmaps and an array of 16 bits integers to
+count facts, resulting in $32 + 2P$ bytes per node, where $P$ is the
+number of predicates.
 
-\subsection{Byte-Code}
+We do a small optimization to reduce the number of derivations of
+persistent facts and, for that, we divide the program rules into two
+sets: \emph{persistent rules} and \emph{non persistent rules}.
+Persistent rules are rules where only persistent facts are
+involved. We compile such rules incrementally, i.e., we attempt to
+fire all rules where a persistent fact is used. This is called
+the \emph{pipelined semi-naive} evaluation and it originated in the P2
+system~\cite{Loo-condie-garofalakis-p2}. This evaluation method avoids
+excessing re-derivations of the same fact. The order of derivation
+does not matter for those rules, since only persistent facts are used.
 
-A byte-code file contains meta-data about the program's predicates, initial nodes, partitioning
-information, and code for each rule.
-Each VM thread has 32 registers that are used during rule execution.
-Registers can store facts, integers, floats, node addresses and pointers to runtime 
-data structures (lists and structures). When registers store facts, we can reference
-fields in the fact through the register.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-Consider a rule \texttt{!a(X,Y), b(X,Z), c(X,Y) -o d(Y)} and a database with
-\texttt{!a(1,2)}, \texttt{!a(2,3)}, \texttt{b(1,3)}, \texttt{b(5,3)}, \texttt{c(1,2)}, \texttt{c(1,3)},
-\texttt{c(5,3)}. Rule execution proceeds in a series of recursive loops, as follows: the first loop retrieves an
-iterator for the persistent facts of \texttt{!a/2} and moves the first valid fact, \texttt{!a(1,2)},
-to register 0; the inner loop retrieves linear facts that match \texttt{b(1,Z)} (from the
-\emph{join constraint}) and moves \texttt{b(1,3)} to register 1; in the final
-loop we move \texttt{c(1,2)} to register 2 and the body of the rule is successfully matched. Next, we
-derive \texttt{d(2)}, where \texttt{2} comes from register 0.
-Fig.~\ref{fig:byte_code} shows the byte-code for this example.
+\subsection{Rule Execution}
 
-\begin{figure}[]
-\scriptsize\begin{Verbatim}
+A byte-code file contains meta-data about the program's predicates,
+initial nodes, partitioning information, and code for each rule. Each
+VM thread has 32 registers that are used during rule
+execution. Registers can store facts, integers, floats, node addresses
+and pointers to runtime data structures (lists and structures). When
+registers store facts, we can reference fields in the fact through the
+register.
+
+Consider the example in Fig.~\ref{fig:byte_code} that shows the LM
+byte-code for the rule \texttt{!a(X,Y), b(X,Z), c(X,Y) -o
+d(Y)}. Consider also a database with the
+facts \texttt{!a(1,2)}, \texttt{!a(2,3)}, \texttt{b(1,3)},
+\texttt{b(5,3)}, \texttt{c(1,2)}, \texttt{c(1,3)}, \texttt{c(5,3)}. Rule
+execution for this rule and facts proceeds in a series of recursive
+loops, as follows. The first loop retrieves an iterator for the
+persistent facts of \texttt{!a/2} and moves the first valid
+fact, \texttt{!a(1,2)}, to register 0. The inner loop retrieves linear
+facts that match \texttt{b(1,Z)} (from the
+\emph{join constraint}) and moves \texttt{b(1,3)} to register 1. The 
+final loop moves \texttt{c(1,2)} to register 2 and the body of the
+rule is successfully matched. Next, we derive \texttt{d(2)},
+where \texttt{2} comes from register 0.
+
+\begin{figure}[ht]
+{\footnotesize
+\begin{Verbatim}
 PERSISTENT ITERATE a MATCHING TO reg 0
    LINEAR ITERATE b MATCHING TO reg 1
-      (match).0=0.0
+      (match).0=0.0             // match argument X
       LINEAR ITERATE c MATCHING TO reg 2
-        (match).0=0.0
-        (match).1=0.1
+         (match).0=0.0          // match argument X
+         (match).1=0.1          // match argument Y
          ALLOC d TO reg 3
-      MVFIELDFIELD 0.1 TO 3.0
-      ADDLINEAR reg 3
+         MVFIELDFIELD 0.1 TO 3.0  // get argument Y
+         ADDLINEAR reg 3             // derive d(Y)
          REMOVE reg 2
@@
 \end{Verbatim}
-\caption{\small{Byte-code for rule \texttt{!a(X,Y), b(X,Z), c(X,Y) -o d(Y).}}}
+}
+\caption{LM byte-code for rule \texttt{!a(X,Y), b(X,Z), c(X,Y) -o d(Y)}}
 \label{fig:byte_code}
@@
 
-In case of failure, we jump to the previous outer loop in order to try the next candidate fact.
-If a rule matches and the head is derived, we backtrack to the inner most \emph{valid loop}, i.e.,
-the first inner loop that uses linear facts or, if there are no linear facts involved, to the previous 
-inner loop. We need to jump to a valid loop because we may have loops with linear facts that are now invalid.
-In our example, we would jump to the
-loop of \texttt{b(X,Z)} and not \texttt{c(X,Y)}, since \texttt{b(1,3)} was consumed.
+In case of failure, we jump to the previous loop in order to try the
+next candidate fact. In case of rule success, the head is derived and
+we should backtrack to the inner most \emph{valid loop}, i.e., the
+older loop that uses linear facts or, if there are no linear facts
+involved, to the previous loop. We need to jump to a valid loop
+because we may have loops with linear facts that are now invalid. In
+our example, we would jump to the loop of \texttt{b(X,Z)} and
+not \texttt{c(X,Y)}, since \texttt{b(1,3)} was consumed.
 
-\begin{figure}[]
-\scriptsize\begin{Verbatim}
+As an optimization, the compiler re-orders the fact expressions used
+in the body in order to make execution more efficient. For example, it
+forces the join constraints in rules to appear first so that matching
+will fail sooner rather than later. It also does the same for
+constraints. Note also that for every loop, the compiler adds
+the \emph{match object}, which contains information about which
+arguments need to match, so that runtime matching be efficient.
+
+Our compiler also detects cases where we re-derive a linear fact with
+new arguments. For example, as shown in Fig.~\ref{code:update}, the LM
+byte-code for rule \texttt{a(N) -o a(N+1)} will compile to code that
+reuses/updates the old \texttt{a(N)} fact when deriving the
+new \texttt{a(N+1)} fact.
+
+\begin{figure}[ht]
+{\footnotesize
+\begin{Verbatim}
 LINEAR ITERATE a MATCHING TO reg 0
-  MVFIELDREG 0.0 TO reg 1
+  MVFIELDREG 0.0 TO reg 1       // initial argument
   MVINTREG INT 1 TO reg 2
   reg 1 INT PLUS reg 2 TO reg 3
-  MVREGFIELD reg 3 TO 0.0
+  MVREGFIELD reg 3 TO 0.0  // reuse/update argument
   UPDATE reg 0
@@
 \end{Verbatim}
-\caption{\small{Byte-code for rule \texttt{a(N) -o a(N+1)}.}}
+}
+\caption{\small{LM byte-code for rule \texttt{a(N) -o a(N+1)}}}
 \label{code:update}
@@
 
-The compiler re-orders the fact expressions used in the body in order to make execution more
-efficient. For example, it forces the join constraints in rules to appear at the beginning so
-that matching will fail sooner rather than later. It also does the same for constraints.
-Note that for every loop, the compiler adds the \emph{match object}, which contains information
-about which arguments need to match, so that runtime matching is efficient.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-Our compiler also detects cases where we re-derive a linear fact with new arguments.
-For example, as shown in Fig.~\ref{code:update}, the rule \texttt{a(N) -o a(N+1)}
-will compile to code that reuses the old \texttt{a(N)} fact.
-We use a \texttt{flags} field to mark updated nodes (presented next).
+\subsection{Indexing}
+\label{indexing}
 
-\subsection{Indexing}\label{indexing}
+To improve fact lookup, the VM employs a fully dynamic mechanism to
+decide which argument may be optimal to index.  The algorithm is
+performed in the beginning of the execution and empirically tries to
+assess the argument of each predicate that more equally spreads the
+database across the values of the argument.  A single thread performs
+the algorithm for all predicates.
 
-To improve fact lookup,
-the VM employs a fully dynamic mechanism to decide which argument may be optimal to index.
-The algorithm is performed in the beginning of the execution and empirically tries to assess the argument
-of each predicate that more equally spreads the database across the values of the argument.
-A single thread performs the algorithm for all predicates.
+The indexing algorithm is performed in three main steps. First, it
+gathers statistics of lookup data by keeping a counter for each
+predicate's argument.  Every time a fact search is performed where
+arguments are fixed to a value, the counter of such arguments is
+incremented. This phase is performed during rule execution for a small
+fraction of the nodes in the program.
 
-The indexing algorithm is performed in three main steps. First, it gathers statistics of lookup data by keeping a counter
-for each predicate's argument.
-Every time a fact search is performed where arguments are fixed to a value, the counter of such arguments is incremented.
-This phase is performed during rule execution for a small fraction of the nodes in the program.
-
-The second step of the algorithm then decides the candidate arguments of each predicate.
-If a predicate was not searched with any fixed arguments, then it will be not indexed.
-If only one argument was fixed, then such argument is set as the indexing argument. Otherwise, the top 2 arguments
-are selected for the third phase, where \emph{entropy statistics} are collected dynamically.
+The second step of the algorithm then decides the candidate arguments
+of each predicate.  If a predicate was not searched with any fixed
+arguments, then it will be not indexed.  If only one argument was
+fixed, then such argument is set as the indexing argument. Otherwise,
+the top 2 arguments are selected for the third phase,
+where \emph{entropy statistics} are collected dynamically.
 
@@
 
-{\scriptsize
+{\tiny
 \[
@@
 
-Where $A$ is the target argument, $F$ is the multi-set of linear facts for the target predicate, $values(F, A)$ is set of values of the argument $A$, $count(F, A = v)$ counts the number
-of linear facts where argument $A$ is equal to $v$ and $total(F)$ counts the number of linear facts in $F$.
-The entropy value is a good metric because it tells us how much information is needed to describe an argument.
-If more information is needed, then that must be the best argument to index.
+Where $A$ is the target argument, $F$ is the multi-set of linear facts
+for the target predicate, $values(F, A)$ is set of values of the
+argument $A$, $count(F, A = v)$ counts the number of linear facts
+where argument $A$ is equal to $v$ and $total(F)$ counts the number of
+linear facts in $F$.  The entropy value is a good metric because it
+tells us how much information is needed to describe an argument.  If
+more information is needed, then that must be the best argument to
+index.
 
-For one of the arguments to score, $Entropy(A, F)$ multiplied by the number of times it has been used for lookup must be larger than the other argument.
+For one of the arguments to score, $Entropy(A, F)$ multiplied by the
+number of times it has been used for lookup must be larger than the
+other argument. The argument with the best score is selected and then
+a global variable called \texttt{indexing\_epoch} is updated.  In
+order to convert the node's linked lists into hash tables, each node
+also has a local variable called \texttt{indexing\_epoch} that is
+compared to the global variable in order to rebuild the node database
+according to the new indexing information.
 
-The argument with the best score is selected and then
-a global variable called \texttt{indexing\_epoch} is updated.
-In order to convert the node's linked lists into hash tables, each node also has a local variable called \texttt{indexing\_epoch}
-that is compared to the global variable in order to rebuild the node database according to the new indexing
-information.
+Our VM also dynamically resizes the hash table if necessary. When the
+hash table becomes too dense, it is resized to the double. When it
+becomes too sparse, it is reduced in half or simply transformed back
+into a doubly linked list. This is done once in a while, before a node
+executes.
 
-Our VM also dynamically resizes the hash table if necessary. When the hash table becomes
-too dense, it is resized to the double. When it becomes too sparse, it is reduced in half
-or simply transformed back into a doubly linked list. This is done once in a while, before a node executes.
+We have seen very good results with this scheme. For example, for the
+all-pairs shortest paths program, we obtained a 2 to 5-fold
+improvement in sequential execution time.  The overhead of dynamic
+indexing is negligible since programs run almost as fast as if the
+indices have been added from the start.
 
-We have seen very good results with this scheme. For example, for the all-pairs shortest paths program,
-we obtained a 2 to 5-fold improvement in sequential execution time.
-The overhead of dynamic indexing is negligible since programs run almost as fast
-as if the indices have been added from the start.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-LM supports recursive types such as lists and pairs. These complex data structures are stored in
-the heap of the VM and are managed through reference counting. For instance, each list
-is a \emph{cons cell} with 3 fields: \texttt{tail}, the pointer to the next element of the list;
-\texttt{head}, the element stored by this element of the list; and \texttt{refs} that counts the
-number of pointers to this list element in the VM. The list is deleted from the heap whenever
-\texttt{refs} is decremented to 0.
+LM also supports recursive types such as lists and pairs. These
+complex data structures are stored in the heap of the VM and are
+managed through reference counting. For instance, each list is
+a \emph{cons cell} with 3 fields: \texttt{tail}, the pointer to the
+next element of the list; \texttt{head}, the element stored by this
+element of the list; and \texttt{refs} that counts the number of
+pointers to this list element in the VM. The list is deleted from the
+heap whenever \texttt{refs} is decremented to 0.

### 2014-07-25 paper.tex
Update.

@@
 
+\toappear{Permission to make digital or hard copies of all or part of this work for personal or classroom use is granted without fee provided that copies are not made or distributed for profit or commercial advantage and that copies bear this notice and the full citation on the first page.
+Copyrights for components of this work owned by others than ACM must be honored.
+Abstracting with credit is permitted.
+To copy otherwise, or republish, to post on servers or to redistribute to lists, requires prior specific permission and/or a fee.
+Request permissions from Permissions@acm.org.
+
+PPDP '14, September 08 - 10 2014, Canterbury, United Kingdom
+Copyright 2014 ACM 978-1-4503-2947-7/14/09\$15.00.
+http://dx.doi.org/10.1145/2643135.2643150 }
 \title{Design and Implementation of a Multithreaded Virtual Machine for Executing Linear Logic Programs}

### 2014-07-25 virtual_machine.tex
Update.

@@
 number of threads) and then each thread will work on their own
-subgraph. Reduction of inter-thread communication is achieved by first
-ordering the node addresses present in the code in such a way that
+subgraph. Reduction of communication between nodes in different threads
+is achieved by first ordering the node addresses present in the code in such a way that
 closer nodes are clustered together and then partitioning them to
@@
 (\emph{Persistent DB}); the rule matching structures (\emph{Rule
-Engine}); and a auxiliary buffer for storing intermediate facts
+Engine}); and an auxiliary buffer for storing intermediate facts
 coming from other threads (\emph{Fact Buffer}).
@@
 // rules
-a, e(1) -o b.
-a -o c.
-b -o d.
-e(0) -o f.
-c -o e(1).
+1: a, e(1) -o b.
+2: a -o c.
+3: b -o d.
+4: e(0) -o f.
+5: c -o e(1).
 
@@
 structures presented in Fig.~\ref{fig:rule_engine}. We use 32 bits
-integers to implement bitmaps and an array of 16 bits integers to
-count facts, resulting in $32 + 2P$ bytes per node, where $P$ is the
-number of predicates.
+integers to implement the 3 bitmaps and an array of 16 bits integers to
+count facts.
 
@@
 heap whenever \texttt{refs} is decremented to 0.
+
+We avoid garbage collection schemes since objects are created and discarded
+in very specific points of the virtual machine and our objects cannot contain circular references.
+A reference counting mechanism is thus more appropriate than a parallel garbage collector which
+would entail pausing the execution of the program to garbage collect all the unused objects.

### 2014-07-25 language.tex
Fixed.

@@
 provides several aggregate operations, including the \emph{minimum},
-\emph{maximum}, \emph{sum} and \emph{count}.
+\emph{maximum}, \emph{sum} and \emph{count}. Note that the \texttt{.} syntax indicates
+an empty list of variables and \texttt{1} is borrowed from linear logic
+and represents an empty derivation.
+
+Comprehensions and aggregates are logically justified by the underlying
+proof system of the language. Our proof system is extended with greatest
+fixed points~\cite{Baelde:2012:LGF:2071368.2071370}, which allows us to describe
+recursive definitions such as comprehensions or aggregates.
 
@@
 concurrency, the programmer is encouraged to write rules that take
-advantage of the non-deterministic nature of execution.
+advantage of the non-deterministic nature of execution since too much
+determinism will naturally imply less scalability and more synchronization
+when executing the programs.

### 2014-07-25 virtual_machine.tex
Fixed.

@@
 is achieved by first ordering the node addresses present in the code in such a way that
-closer nodes are clustered together and then partitioning them to
+connected nodes are clustered together and then partitioning them to
 threads. During compilation, we take note of predicates that are used
-in rules with communication (arguments with type \emph{node}) and
-then build a graph of nodes from the program's axioms. The nodes of
+for communication (arguments with type \emph{node}) and
+then build a graph of nodes from them. The nodes of
 the graph are then ordered by using a breadth-first search algorithm
@@
 in \emph{Fact Buffer} will then be processed whenever the
-corresponding node is about to be executed.
+corresponding node's flag becomes active.
 
@@
 
-Where $A$ is the target argument, $F$ is the multi-set of linear facts
+Where $A$ is the target argument, $F$ is the set of linear facts
 for the target predicate, $values(F, A)$ is set of values of the

### 2014-07-25 related_work.tex
Fixed.
More data

@@
 based on a data-flow model.
+
+Several linear logic programming languages have been developed in the past~\cite{Miller85anoverview}.
+Lolli, a programming language based on a fragment of intuitionistic linear logic~\cite{Hodas94logicprogramming}, 
+proves goals by lazily managing the context of linear resources during backward-chaining proof search.
+LolliMon~\cite{Lopez:2005:MCL:1069774.1069778} is a concurrent linear logic programming language that integrates
+both forward and backward-chaining search, where the backward-chaining phase is done sequentially but
+the forward-chaining is done concurrently inside a monad. The backward-chaining phase is suspended
+between the forward-chaining phases, where a fix-point is computed.
+LolliMon is derived from the concurrent logical framework called CLF~\cite{Watkins:2004uq}.

### 2014-08-07 experiments.tex
Update.
Fixed.
ricroc's changes.

@@
 graph-based problems in machine learning. C and GraphLab perform about
-the same since both use C/C++. Python runs very slowly since it is a
+the same since they are both compiled to machine code, although
+the GraphLab version is highly optimized to run on multicore architectures.
+Python runs very slowly since it is a
 dynamic programming language and BP has many mathematical
@@
 \hline\hline
-\textbf{10}  & 0.67 & 0.03 & 1.00 \\
+\textbf{10}  & 1.00 & 0.03 & 1.00 \\
 \textbf{50}  & 1.77 & 0.04 & 1.73 \\
@@
    {\label{exp:ggc}\includegraphics[width=\plotsize]{figures/results-ggc.pdf}}
-\caption{Experimental results for the GGC and PageRank algorithms}
+\caption{Experimental results for the PageRank and GGC  algorithms}
 \end{figure*}
@@
 \hspace{0.5cm}
-\subfigure[BP (400x400~image)]
+\subfigure[Belief Propagation (400x400~image)]
    {\label{exp:bp}\includegraphics[width=\plotsize]{figures/results-bp.pdf}}
@@
 
-In this section, we present the amount of \textbf{Memory} used by the VM after the
+In this section, we present the amount of memory used by the VM after the
 programs presented in the previous section complete. We also count the number of
-\textbf{Facts} stored in the database. This serves as an indication of
+facts stored in the database. This serves as an indication of
 how much memory is needed in terms of the number of facts stored in the database.
@@
 \hline\hline
-\multicolumn{3}{l}{Average (W/ MiniMax) } & 288 bytes \\
+\multicolumn{3}{l}{Average (without MiniMax) } & 288 bytes \\
 \end{tabular}}
@@
 positions of the queens in the board. Storing lists is far more expensive than storing integral
-values such as integers or floating point numbers. The same thing happens with BP because
+values such as integers or floating point numbers. The same happens with BP because
 the belief values are stored as lists of floating point numbers.

### 2014-08-07 introduction.tex
ricroc's changes.

@@
 
-The last decade has seen a tremendous growth in content available in the World Wide Web, and, more specifically,
-in information generated from online social networks. The structure of such content is usually a graph, a very
-flexible structure suited to represent content where pairs of items are linked.
-In order to process such information, there has been an increased interest in running graph-based
-algorithms concurrently and efficiently on top of distributed networks and computer architectures (multicores).
-Currently available libraries and frameworks are built on top of imperative programming languages, which
-require the programmer to know how to properly use the framework and the language. Reasoning about such
-programs requires knowing the intricacies of the framework, how computation is scheduled and how processing
-units coordinate between each other.
+The last decade has seen a tremendous growth in content available in
+the World Wide Web, and, more specifically, in information generated
+from online social networks. The structure of such content is usually
+a graph, a very flexible structure suited to represent content where
+pairs of items are linked.  In order to process such information,
+there has been an increased interest in running graph-based algorithms
+concurrently and efficiently on top of distributed networks and
+computer architectures (multicores).  Currently available libraries
+and frameworks are built on top of imperative programming languages,
+which require the programmer to know how to properly use the framework
+and the language. Reasoning about such programs requires knowing the
+intricacies of the framework, how computation is scheduled and how
+processing units coordinate between each other.
 
-Some well known frameworks include Dryad, Pregel and GraphLab.
-The Dryad system~\cite{Isard:2007:DDD:1272996.1273005} combines computational vertices
-with communication channels (edges) to form a data-flow graph. The program is scheduled to
-run on multiple processors or cores and data is partitioned during runtime. Routines that run on computational vertices
-are sequential, with no synchronization.
-The Pregel system~\cite{Malewicz:2010:PSL:1807167.1807184} is also graph based, although programs have a more strict
-structure. They must be represented as a sequence of iterations where each iteration is composed of computation and message passing.
-Pregel is specially suited to work on big graphs and to scale to large architectures.
-GraphLab~\cite{GraphLab2010} is a C++ framework for developing parallel machine learning
-algorithms. While Pregel uses message passing, GraphLab allows nodes to have read/write
-access to different scopes through different concurrent access models in order to balance
-performance and data consistency. While some programs only need to access the local node's
-data, others may need to update edge information. Each consistency model will provide different
-guarantees that are better adapted to some algorithms. GraphLab also provides different
-schedulers that dictate the order in which node's are computed.
+Some well known frameworks include Dryad, Pregel and GraphLab.  The
+Dryad system~\cite{Isard:2007:DDD:1272996.1273005} combines
+computational vertices with communication channels (edges) to form a
+data-flow graph. The program is scheduled to run on multiple
+processors or cores and data is partitioned during runtime. Routines
+that run on computational vertices are sequential, with no
+synchronization.  The Pregel
+system~\cite{Malewicz:2010:PSL:1807167.1807184} is also graph based,
+although programs have a more strict structure. They must be
+represented as a sequence of iterations where each iteration is
+composed of computation and message passing.  Pregel is specially
+suited to work on big graphs and to scale to large architectures.
+GraphLab~\cite{GraphLab2010} is a C++ framework for developing
+parallel machine learning algorithms. While Pregel uses message
+passing, GraphLab allows nodes to have read/write access to different
+scopes through different concurrent access models in order to balance
+performance and data consistency. While some programs only need to
+access the local node's data, others may need to update edge
+information. Each consistency model will provide different guarantees
+that are better adapted to some algorithms. GraphLab also provides
+different schedulers that dictate the order in which node's are
+computed.
 
@@
 
-\iffalse
-Logic programming languages split into two main
-groups: \emph{backwards-chaining} and \emph{forward-chaining}. In a
-backwards-chaining logic programming language, programs are composed
-of a set of rules that can be activated by inputting a query. Given a
-query $q(\hat{x})$, an interpreter will work backwards by matching
+Logic programming split into two main groups:
+\emph{backwards-chaining} and \emph{forward-chaining} languages. In a
+backwards-chaining programming language, programs are composed of a
+set of rules that can be activated by inputting a query. Given a query
+$q(\hat{x})$, an interpreter will work backwards by matching
 $q(\hat{x})$ against the head of a rule. If found, the interpreter
@@
 parallel. Researchers took advantage of Prolog's non-determinism to
-evaluate subgoals in parallel with models such
-as \emph{And-parallelism}
-and \emph{Or-parallelism}~\cite{Gupta:2001:PEP:504083.504085}.
-\fi
+evaluate subgoals in parallel with models such as
+\emph{And-parallelism} and
+\emph{Or-parallelism}~\cite{Gupta:2001:PEP:504083.504085}.
 
-In forward-chaining logic programming, programs start with
+In a forward-chaining logic programming language, programs start with
 a database of facts (filled with the program's axioms) and a set of
@@
 The LM virtual machine was designed from scratch to run LM programs on
-multicore machines in an efficient manner\footnote{Source code available at \url{http://github.com/flavioc/meld}}
+multicore machines in an efficient manner\footnote{Source code available at \url{http://github.com/flavioc/meld}}.
 The virtual machine is multithreaded and executes byte-code that is generated
@@
 
-This paper is organized as follows. First, we briefly introduce the LM
-language. Then, we present an overview of the virtual machine and
-describe in more detail the code organization, thread management, rule
-execution and database organization. Finally, we present our experiments
-and outline some conclusions.
+This remainder of the paper is organized as follows. First, we briefly
+introduce the LM language. Then, we present an overview of the virtual
+machine and describe in more detail the code organization, thread
+management, rule execution and database organization. Finally, we
+present our experiments and outline some conclusions.

### 2014-08-07 language.tex
More data
ricroc's changes.

@@
 
-\begin{figure}[t]
+\begin{figure}[ht]
 {\footnotesize
@@
 Finally, with the \texttt{replace(@0, 6, x)} axiom instantiated at the
-root node $@0$, we intend to change the value of key 6 to value
-`$x$'. Note that when writing rules or axioms, persistent facts are
-preceded with a \texttt{!}.
+root node \texttt{@0} (line 36), we intend to change the value of key
+6 to value \texttt{x}. Note that when writing rules or axioms,
+persistent facts are preceded with a `\texttt{!}'.
+
+Figure~\ref{fig:btree_trace} illustrates the trace of the
+execution. Note that the program database is partitioned by the tree
+nodes using the first argument of each fact. In
+Fig.~\ref{fig:btree_trace1}, we present the database filled with the
+program's axioms. Next, we follow the right branch using rule 3 since
+$6 > 3$ (Fig.~\ref{fig:btree_trace2}). We then use the same rule
+again in Fig.~\ref{fig:btree_trace3} where we finally reach the key
+6. Here, we apply rule 1 and \texttt{value(@6, 6, g)} is updated to
+\texttt{value(@6, 6, x)} (Fig.~\ref{fig:btree_trace4}).
 
@@
 \end{figure*}
-
-Figure~\ref{fig:btree_trace} illustrates the trace of the
-execution. Note that the program database is partitioned by the tree
-nodes using the first argument of each fact. In
-Fig.~\ref{fig:btree_trace1}, we present the database filled with the
-program's axioms. Next, we follow the right branch using rule 3 since
-$6 > 3$ (Fig.~\ref{fig:btree_trace2}). We then use the same rule
-again in Fig.~\ref{fig:btree_trace3} where we finally reach the key
-6. Here, we apply rule 1 and \texttt{value(@6, 6, g)} is updated to
-\texttt{value(@6, 6, x)} (Fig.~\ref{fig:btree_trace4}).
 
@@
 \begin{Verbatim}
-!edge(@1, @2).   !edge(@1, @3).
+!edge(@1, @2).
+!edge(@1, @3).
 iterate(@1).
@@
 \begin{Verbatim}
-price(@1, 3).   price(@1, 4).   price(@1, 5).
+price(@1, 3).
+price(@1, 4).
+price(@1, 5).
 count-prices(@1).
@@
 recursive definitions such as comprehensions or aggregates.
+A more detailed description is found in~\cite{cruz-iclp14}.
 

### 2014-08-07 paper.tex
ricroc's changes.

--- 2014-07-25 paper.tex
@@
-\documentclass{acm_proc_article-sp}
+\documentclass{sigplanconf}
 \usepackage[utf8]{inputenc}
@@
 
+\special{papersize=8.5in,11in}
+\setlength{\pdfpageheight}{\paperheight}
+\setlength{\pdfpagewidth}{\paperwidth}
+
+\conferenceinfo{PPDP '14}{September 08--10, 2014, Canterbury, United Kingdom}
+\copyrightyear{2014} 
+\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
+\doi{nnnnnnn.nnnnnnn}
+
+% Uncomment one of the following two, if you are not going for the 
+% traditional copyright transfer agreement.
+\exclusivelicense                % ACM gets exclusive license to publish, 
+                                  % you retain copyright
+%\permissiontopublish             % ACM gets nonexclusive license to publish
+                                  % (paid open-access papers, 
+                                  % short abstracts)
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \title{Design and Implementation of a Multithreaded Virtual Machine for Executing Linear Logic Programs}
 
-\numberofauthors{3}
-\author{
-  \alignauthor Flavio Cruz \\
-    \affaddr{Carnegie Mellon University}
-    \affaddr{Pittsburgh, PA 15213, USA}\\
-    \email{fmfernan@cs.cmu.edu} \\
-  \alignauthor Ricardo Rocha \\
-    \affaddr{CRACS \& INESC TEC}\\
-    \affaddr{University of Porto}\\
-    \affaddr{Rua Campo Alegre 1021/1055}\\
-    \affaddr{4169-007 Porto, Portugal}\\
-    \email{ricroc@dcc.fc.up.pt}
-  \alignauthor Seth Copen Goldstein \\
-    \affaddr{Carnegie Mellon University}
-    \affaddr{Pittsburgh, PA 15213, USA}\\
-    \email{seth@cs.cmu.edu} \\
-}
+\authorinfo{Flavio Cruz}
+           {Carnegie Mellon University\\Pittsburgh, PA 15213, USA}
+           {fmfernan@cs.cmu.edu}
+
+\authorinfo{Ricardo Rocha}
+           {CRACS \& INESC TEC\\University of Porto\\Rua Campo Alegre 1021/1055\\4169-007 Porto, Portugal}
+           {ricroc@dcc.fc.up.pt}
+
+\authorinfo{Seth Copen Goldstein}
+           {Carnegie Mellon University\\Pittsburgh, PA 15213, USA}
+           {seth@cs.cmu.edu}
 
@@
 \input{conclusions}
+\input{ack}
 
@@
 
-\bibliographystyle{abbrv}
-\bibliography{refs}
+\bibliographystyle{abbrvnat}
+
+% The bibliography should be embedded for final submission.
+%\bibliography{refs}
+
+\begin{thebibliography}{35}
+\providecommand{\natexlab}[1]{#1}
+\providecommand{\url}[1]{\texttt{#1}}
+\expandafter\ifx\csname urlstyle\endcsname\relax
+  \providecommand{\doi}[1]{doi: #1}\else
+  \providecommand{\doi}{doi: \begingroup \urlstyle{rm}\Url}\fi
+
+\bibitem[Ali and Karlsson(1990)]{Ali:1990fk}
+K.~Ali and R.~Karlsson.
+\newblock Full prolog and scheduling or-parallelism in muse.
+\newblock \emph{International Journal of Parallel Programming}, 19\penalty0
+  (6):\penalty0 445--475, 1990.
+\newblock ISSN 0885-7458.
+\newblock \doi{10.1007/BF01397627}.
+\newblock URL \url{http://dx.doi.org/10.1007/BF01397627}.
+
+\bibitem[Ali(1988)]{Ali88}
+K.~A.~M. Ali.
+\newblock Or-parallel execution of prolog on bc-machine.
+\newblock In R.~A. Kowalski and K.~A. Bowen, editors, \emph{ICLP/SLP}, pages
+  1531--1545. MIT Press, 1988.
+\newblock ISBN 0-262-61056-6.
+\newblock URL \url{http://dblp.uni-trier.de/db/conf/iclp/iclp88.html#Ali88}.
+
+\bibitem[Ashley-Rollman et~al.(2007)Ashley-Rollman, Rosa, Srinivasa, Pillai,
+  Goldstein, and Campbell]{ashley-rollman-derosa-iros07wksp}
+M.~P. Ashley-Rollman, M.~D. Rosa, S.~S. Srinivasa, P.~Pillai, S.~C. Goldstein,
+  and J.~D. Campbell.
+\newblock Declarative programming for modular robots.
+\newblock In \emph{Workshop on Self-Reconfigurable Robots/Systems and
+  Applications at IROS 2007}, 2007.
+
+\bibitem[Ashley-Rollman et~al.(2009)Ashley-Rollman, Lee, Goldstein, Pillai, and
+  Campbell]{ashley-rollman-iclp09}
+M.~P. Ashley-Rollman, P.~Lee, S.~C. Goldstein, P.~Pillai, and J.~D. Campbell.
+\newblock A language for large ensembles of independently executing nodes.
+\newblock In \emph{International Conference on Logic Programming (ICLP)}, 2009.
+
+\bibitem[Baelde(2012)]{Baelde:2012:LGF:2071368.2071370}
+D.~Baelde.
+\newblock Least and greatest fixed points in linear logic.
+\newblock \emph{ACM Transactions on Computational Logic}, 13\penalty0
+  (1):\penalty0 1--44, 2012.
+
+\bibitem[Betz and Fr{\"u}hwirth(2005)]{Betz:2005kx}
+H.~Betz and T.~Fr{\"u}hwirth.
+\newblock A linear-logic semantics for constraint handling rules.
+\newblock In \emph{Principles and Practice of Constraint Programming - CP
+  2005}, volume 3709 of \emph{Lecture Notes in Computer Science}, pages
+  137--151. 2005.
+
+\bibitem[Butler et~al.(1988)Butler, Disz, Lusk, Olson, Overbeek, and
+  Stevens]{ButlerDLOOS88}
+R.~Butler, T.~Disz, E.~L. Lusk, R.~Olson, R.~A. Overbeek, and R.~L. Stevens.
+\newblock Scheduling or-parallelism: An argonne perspective.
+\newblock In R.~A. Kowalski and K.~A. Bowen, editors, \emph{ICLP/SLP}, pages
+  1590--1605. MIT Press, 1988.
+\newblock ISBN 0-262-61056-6.
+\newblock URL
+  \url{http://dblp.uni-trier.de/db/conf/iclp/iclp88.html#ButlerDLOOS88}.
+
+\bibitem[Colmerauer and Roussel(1993)]{Colmerauer:1993:BP:154766.155362}
+A.~Colmerauer and P.~Roussel.
+\newblock The birth of prolog.
+\newblock In \emph{The Second ACM SIGPLAN Conference on History of Programming
+  Languages}, pages 37--52, New York, NY, USA, 1993.
+
+\bibitem[Costa et~al.(2011)Costa, Damas, and
+  Rocha]{DBLP:journals/corr/abs-1102-3896}
+V.~S. Costa, L.~Damas, and R.~Rocha.
+\newblock The yap prolog system.
+\newblock \emph{CoRR}, abs/1102.3896, 2011.
+
+\bibitem[Cruz et~al.(2014)Cruz, Rocha, Goldstein, and Pfenning]{cruz-iclp14}
+F.~Cruz, R.~Rocha, S.~Goldstein, and F.~Pfenning.
+\newblock {A Linear Logic Programming Language for Concurrent Programming over
+  Graph Structures}.
+\newblock \emph{Journal of Theory and Practice of Logic Programming, 30th
+  International Conference on Logic Programming (ICLP 2014), Special Issue},
+  14\penalty0 (4 \& 5):\penalty0 493--507, July 2014.
+
+\bibitem[Culler et~al.(1993)Culler, Goldstein, Schauser, and von
+  Eicken]{CullerGSvE93}
+D.~E. Culler, S.~C. Goldstein, K.~E. Schauser, and T.~von Eicken.
+\newblock {TAM --- a compiler controlled threaded abstract machine}.
+\newblock \emph{Journal of Parallel and Distributed Computing}, 18:\penalty0
+  347--370, July 1993.
+\newblock URL \url{http://www.cs.cmu.edu/~seth/papers/CullerGSvE93.pdf}.
+
+\bibitem[Girard(1987)]{girard-87}
+J.-Y. Girard.
+\newblock Linear logic.
+\newblock \emph{Theoretical Computer Science}, 50\penalty0 (1):\penalty0
+  1--102, 1987.
+
+\bibitem[Goldstein(1994)]{goldstein-tr94}
+S.~C. Goldstein.
+\newblock The implementation of a threaded abstract machine.
+\newblock Technical Report UCB/CSD-94-818, EECS Department, University of
+  California, Berkeley, 1994.
+\newblock URL \url{http://www.cs.cmu.edu/~seth/papers/goldstein-tr94.pdf}.
+
+\bibitem[Gonzalez et~al.(2009)Gonzalez, Low, and
+  Guestrin]{Gonzalez+al:aistats09paraml}
+J.~Gonzalez, Y.~Low, and C.~Guestrin.
+\newblock Residual splash for optimally parallelizing belief propagation.
+\newblock In \emph{Artificial Intelligence and Statistics (AISTATS)}, 2009.
+
+\bibitem[Gupta et~al.(2001)Gupta, Pontelli, Ali, Carlsson, and
+  Hermenegildo]{Gupta:2001:PEP:504083.504085}
+G.~Gupta, E.~Pontelli, K.~A.~M. Ali, M.~Carlsson, and M.~V. Hermenegildo.
+\newblock Parallel execution of prolog programs: {A} survey.
+\newblock \emph{ACM Transactions on Programming Languages and Systems
+  (TOPLAS)}, 23\penalty0 (4):\penalty0 472--602, 2001.
+
+\bibitem[Hermenegildo(1986)]{Hermenegildo:1986:AMB:913061}
+M.~V. Hermenegildo.
+\newblock \emph{An Abstract Machine Based Execution Model for Computer
+  Architecture Design and Efficient Implementation of Logic Programs in
+  Parallel}.
+\newblock PhD thesis, 1986.
+\newblock AAI8700203.
+
+\bibitem[Hodas and Miller(1994)]{Hodas94logicprogramming}
+J.~S. Hodas and D.~Miller.
+\newblock Logic programming in a fragment of intuitionistic linear logic.
+\newblock \emph{Information and Computation}, 110:\penalty0 32--42, 1994.
+
+\bibitem[Holzbaur et~al.(2004)Holzbaur, de~la Banda, Stuckey, and
+  Duck]{DBLP:journals/corr/cs-PL-0408025}
+C.~Holzbaur, M.~J.~G. de~la Banda, P.~J. Stuckey, and G.~J. Duck.
+\newblock Optimizing compilation of constraint handling rules in hal.
+\newblock \emph{CoRR}, cs.PL/0408025, 2004.
+
+\bibitem[Isard et~al.(2007)Isard, Budiu, Yu, Birrell, and
+  Fetterly]{Isard:2007:DDD:1272996.1273005}
+M.~Isard, M.~Budiu, Y.~Yu, A.~Birrell, and D.~Fetterly.
+\newblock Dryad: distributed data-parallel programs from sequential building
+  blocks.
+\newblock In \emph{European Conference on Computer Systems (EuroSys)}, pages
+  59--72, 2007.
+
+\bibitem[Kacsuk(1990)]{Kacsuk:1990:EMP:533578}
+P.~Kacsuk.
+\newblock \emph{Execution Models of PROLOG for Parallel Computers}.
+\newblock MIT Press, Cambridge, MA, USA, 1990.
+\newblock ISBN 0262111497.
+
+\bibitem[Kara et~al.(1997)Kara, Davy, Goodeve, and Nash]{Kara:1997:AMM:265274}
+M.~Kara, J.~R. Davy, D.~Goodeve, and J.~Nash, editors.
+\newblock \emph{Abstract Machine Models for Parallel and Distributed
+  Computing}.
+\newblock IOS Press, Amsterdam, The Netherlands, The Netherlands, 1997.
+\newblock ISBN 90-5199-267-X.
+
+\bibitem[Lam and Sulzmann(2010)]{DBLP:journals/corr/abs-1006-3039}
+E.~S.~L. Lam and M.~Sulzmann.
+\newblock Concurrent goal-based execution of constraint handling rules.
+\newblock \emph{CoRR}, abs/1006.3039, 2010.
+
+\bibitem[Lin and Kumar(1988)]{Lin:1988:AEL:900478}
+Y.~Lin and V.~Kumar.
+\newblock And-parallel execution of logic programs on a shared memory
+  multiprocessor:a summary of results.
+\newblock Technical report, Austin, TX, USA, 1988.
+
+\bibitem[Liu(1998)]{Liu98extendingdatalog}
+M.~Liu.
+\newblock Extending datalog with declarative updates.
+\newblock In \emph{International Conference on Database and Expert System
+  Applications (DEXA)}, volume 1873, pages 752--763, 1998.
+
+\bibitem[Loo et~al.(2006)Loo, Condie, Garofalakis, Gay, and
+  Hellerstein]{Loo-condie-garofalakis-p2}
+B.~T. Loo, T.~Condie, M.~Garofalakis, D.~E. Gay, and J.~M. Hellerstein.
+\newblock Declarative networking: Language, execution and optimization.
+\newblock In \emph{International Conference on Management of Data (SIGMOD)},
+  pages 97--108, 2006.
+
+\bibitem[L\'{o}pez et~al.(2005)L\'{o}pez, Pfenning, Polakow, and
+  Watkins]{Lopez:2005:MCL:1069774.1069778}
+P.~L\'{o}pez, F.~Pfenning, J.~Polakow, and K.~Watkins.
+\newblock Monadic concurrent linear logic programming.
+\newblock In \emph{Proceedings of the 7th ACM SIGPLAN International Conference
+  on Principles and Practice of Declarative Programming}, PPDP '05, pages
+  35--46, New York, NY, USA, 2005.
+
+\bibitem[Low et~al.(2010)Low, Gonzalez, Kyrola, Bickson, Guestrin, and
+  Hellerstein]{GraphLab2010}
+Y.~Low, J.~Gonzalez, A.~Kyrola, D.~Bickson, C.~Guestrin, and J.~M. Hellerstein.
+\newblock Graphlab: {A} new framework for parallel machine learning.
+\newblock In \emph{Conference on Uncertainty in Artificial Intelligence (UAI)},
+  pages 340--349, 2010.
+
+\bibitem[Malewicz et~al.(2010)Malewicz, Austern, Bik, Dehnert, Horn, Leiser,
+  and Czajkowski]{Malewicz:2010:PSL:1807167.1807184}
+G.~Malewicz, M.~H. Austern, A.~J.~C. Bik, J.~C. Dehnert, I.~Horn, N.~Leiser,
+  and G.~Czajkowski.
+\newblock Pregel: a system for large-scale graph processing.
+\newblock In \emph{International Conference on Management of Data (SIGMOD)},
+  pages 135--146, 2010.
+
+\bibitem[Miller(1985)]{Miller85anoverview}
+D.~Miller.
+\newblock An overview of linear logic programming.
+\newblock In \emph{in Computational Logic}, pages 1--5, 1985.
+
+\bibitem[Ramakrishnan and Ullman(1993)]{Ramakrishnan93asurvey}
+R.~Ramakrishnan and J.~D. Ullman.
+\newblock A survey of research on deductive database systems.
+\newblock \emph{Journal of Logic Programming}, 23:\penalty0 125--149, 1993.
+
+\bibitem[Sunderam(1990)]{Sunderam90pvm:a}
+V.~S. Sunderam.
+\newblock Pvm: A framework for parallel distributed computing.
+\newblock \emph{Concurrency: Practice and Experience}, 2:\penalty0 315--339,
+  1990.
+
+\bibitem[{van Rossum}(1995)]{vanRossum:1995:PRM}
+G.~{van Rossum}.
+\newblock {Python} reference manual.
+\newblock Report CS-R9525, Centrum voor Wiskunde en Informatica, Amsterdam, the
+  Netherlands, Apr. 1995.
+\newblock URL \url{http://www.python.org/doc/ref/ref-1.html}.
+
+\bibitem[Warren(1983)]{AICPub641:1983}
+D.~H.~D. Warren.
+\newblock An abstract prolog instruction set.
+\newblock Technical Report 309, AI Center, SRI International, 333 Ravenswood
+  Ave., Menlo Park, CA 94025, Oct 1983.
+
+\bibitem[Warren(1987)]{Warren:1987:OEM:67683.67699}
+D.~H.~D. Warren.
+\newblock Or-parallel execution models of prolog.
+\newblock In \emph{II and Colloquium on Functional and Logic Programming and
+  Specifications (CFLP) on TAPSOFT '87: Advanced Seminar on Foundations of
+  Innovative Software Development}, pages 243--259, New York, NY, USA, 1987.
+  Springer-Verlag New York, Inc.
+\newblock ISBN 0-387-17611-X.
+\newblock URL \url{http://dl.acm.org/citation.cfm?id=67683.67699}.
+
+\bibitem[Watkins et~al.(2004)Watkins, Cervesato, Pfenning, and
+  Walker]{Watkins:2004uq}
+K.~Watkins, I.~Cervesato, F.~Pfenning, and D.~Walker.
+\newblock A concurrent logical framework: The propositional fragment.
+\newblock In \emph{Types for Proofs and Programs}, volume 3085 of \emph{Lecture
+  Notes in Computer Science}, pages 355--377. 2004.
+\newblock \doi{10.1007/978-3-540-24849-1_23}.
+
+\end{thebibliography}
+
 \end{document}

### 2014-08-07 virtual_machine.tex
ricroc's changes.

@@
 We have developed a compiler that compiles LM programs to byte-code
-and a multi-threaded virtual machine (VM) using POSIX threads
+and a multithreaded virtual machine (VM) using POSIX threads
 that runs the byte-code.
@@
             node = steal_node_from_thread(target)
+            if (node) break
          if (!node)
@@
             become_idle(tid)
-            if (!synchronize_termination(tid))
+            if (synchronize_termination(tid))
                return
@@
 the \texttt{flags} field is to mark that a fact is already being
-used. For example, consider the rule \texttt{a(A,B), a(C,D) -o
-...}. When we first pick a fact for \texttt{a(A, B)} from the hash
+used. For example, consider the rule:
+
+{\footnotesize
+\begin{Verbatim}
+a(A,B), a(C,D) -o ...
+\end{Verbatim}
+}
+
+When we first pick a fact for \texttt{a(A, B)} from the hash
 table, we mark it as being used in order to ensure that, when we
@@
 facts per predicate. To understand how our engine works, consider the
-following set of rules and facts:
+following set of facts and rules:
 
@@
 \begin{Verbatim}
-// rules
-1: a, e(1) -o b.
-2: a -o c.
-3: b -o d.
-4: e(0) -o f.
-5: c -o e(1).
-
-// facts
 a.
 e(0).
+
+a, e(1) -o b.  // rule 1
+a -o c.        // rule 2
+b -o d.        // rule 3
+e(0) -o f.     // rule 4
+c -o e(1).     // rule 5
 \end{Verbatim}
@@
 
-Where $A$ is the target argument, $F$ is the set of linear facts
+\noindent
+where $A$ is the target argument, $F$ is the set of linear facts
 for the target predicate, $values(F, A)$ is set of values of the
@@
 pointers to this list element in the VM. The list is deleted from the
-heap whenever \texttt{refs} is decremented to 0.
+heap whenever \texttt{refs} is decremented to zero.
 

### 2014-08-07 paper.tex
flatten latex file.

@@
 \lstset{language=C, basicstyle=\ttfamily, breaklines=true, frame=lines}
-\input{fp-macros}
+\makeatletter{}\usepackage{latexsym}
+\usepackage{amssymb}            \usepackage{stmaryrd}           
+\newcommand{\m}[1]{\mathsf{#1}}
+\newcommand{\f}[1]{\framebox{#1}}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newcommand{\eph}{\mathit{eph}}
+\newcommand{\pers}{\mathit{pers}}
+\newcommand{\um}[1]{\underline{\m{#1}}}
+
+\newcommand{\seq}{\vdash}
+\newcommand{\semi}{\mathrel{;}}
+\newcommand{\lequiv}{\mathrel{\dashv\vdash}}
+
+\newcommand{\lolli}{\multimap}
+\newcommand{\tensor}{\otimes}
+\newcommand{\with}{\mathbin{\binampersand}}
+\newcommand{\paar}{\mathbin{\bindnasrepma}}
+\newcommand{\one}{\mathbf{1}}
+\newcommand{\zero}{\mathbf{0}}
+\newcommand{\bang}{{!}}
+\newcommand{\whynot}{{?}}
+\newcommand{\bilolli}{\mathrel{\raisebox{1pt}{\ensuremath{\scriptstyle\circ}}{\lolli}}}
 
@@
 \setlength{\pdfpagewidth}{\paperwidth}
+\conferenceinfo{PPDP '14}{September 8--10, 2014, Canterbury, UK}
+\copyrightyear{2014}
+\copyrightdata{978-1-4503-2947-7/14/09}
+\doi{2643135.2643150}
 
-\conferenceinfo{PPDP '14}{September 08--10, 2014, Canterbury, United Kingdom}
-\copyrightyear{2014} 
-\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
-\doi{nnnnnnn.nnnnnnn}
-
-% Uncomment one of the following two, if you are not going for the 
-% traditional copyright transfer agreement.
-\exclusivelicense                % ACM gets exclusive license to publish, 
-                                  % you retain copyright
-%\permissiontopublish             % ACM gets nonexclusive license to publish
-                                  % (paid open-access papers, 
-                                  % short abstracts)
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\exclusivelicense
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \maketitle
 
-\input{abstract}
-\input{introduction}
-\input{language}
-\input{virtual_machine}
-\input{experiments}
-\input{related_work}
-\input{conclusions}
-\input{ack}
+\makeatletter{}\begin{abstract}
+Linear Meld is a concurrent forward-chaining linear logic programming
+language where logical facts can be asserted and retracted in a
+structured way. In Linear Meld, a program is seen as a database of
+logical facts and a set of derivation rules. The database of facts is
+partitioned by the nodes of a graph structure which leads to
+parallelism when nodes are executed simultaneously. Due to the
+foundations on linear logic, rules can retract facts in a declarative
+and structured fashion, leading to more expressive programs. We
+present the design and implementation of the virtual machine that we
+implemented to run Linear Meld on multicores, with particular focus on
+thread management, code organization, fact indexing, rule execution,
+and database organization for efficient fact insertion, lookup and
+deletion. Our results show that the virtual machine is
+capable of scaling programs with up to 16 threads and also exhibits
+interesting scalar performance results due to our indexing optimizations.
+\end{abstract}
+
+\category{D.1.3}{PROGRAMMING TECHNIQUES}{Concurrent Programming}[Parallel Programming]
+\category{D.3.4}{PROCESSORS}{Interpreters}
+\category{D.3.4}{PROCESSORS}{Run-time environments}
+
+\terms{Design, Languages, Performance}
+
+\keywords{Linear Logic, Virtual Machine, Implementation}
+
+\makeatletter{}\section{Introduction}
+
+The last decade has seen a tremendous growth in content available in
+the World Wide Web, and, more specifically, in information generated
+from online social networks. The structure of such content is usually
+a graph, a very flexible structure suited to represent content where
+pairs of items are linked.  In order to process such information,
+there has been an increased interest in running graph-based algorithms
+concurrently and efficiently on top of distributed networks and
+computer architectures (multicores).  Currently available libraries
+and frameworks are built on top of imperative programming languages,
+which require the programmer to know how to properly use the framework
+and the language. Reasoning about such programs requires knowing the
+intricacies of the framework, how computation is scheduled and how
+processing units coordinate between each other.
+
+Some well known frameworks include Dryad, Pregel and GraphLab.  The
+Dryad system~\cite{Isard:2007:DDD:1272996.1273005} combines
+computational vertices with communication channels (edges) to form a
+data-flow graph. The program is scheduled to run on multiple
+processors or cores and data is partitioned during runtime. Routines
+that run on computational vertices are sequential, with no
+synchronization.  The Pregel
+system~\cite{Malewicz:2010:PSL:1807167.1807184} is also graph based,
+although programs have a more strict structure. They must be
+represented as a sequence of iterations where each iteration is
+composed of computation and message passing.  Pregel is specially
+suited to work on big graphs and to scale to large architectures.
+GraphLab~\cite{GraphLab2010} is a C++ framework for developing
+parallel machine learning algorithms. While Pregel uses message
+passing, GraphLab allows nodes to have read/write access to different
+scopes through different concurrent access models in order to balance
+performance and data consistency. While some programs only need to
+access the local node's data, others may need to update edge
+information. Each consistency model will provide different guarantees
+that are better adapted to some algorithms. GraphLab also provides
+different schedulers that dictate the order in which node's are
+computed.
+
+Logic programming is an attractive approach to the graph-based algorithms,
+since logic-based languages provide a high-level, declarative approach
+to programming. An important characteristic of logic programming is
+that it offers great potential for implicit parallelism, thus making
+logic programs much easier to parallelize than imperative
+programs. First, logic programs are easier to reason about since they
+are based on logical foundations. Second, logic programmers do not
+need to use low level programming constructs such as locks or
+semaphores to coordinate parallel execution, because logic systems
+hide such details from the programmer.
+
+Logic programming split into two main groups:
+\emph{backwards-chaining} and \emph{forward-chaining} languages. In a
+backwards-chaining programming language, programs are composed of a
+set of rules that can be activated by inputting a query. Given a query
+$q(\hat{x})$, an interpreter will work backwards by matching
+$q(\hat{x})$ against the head of a rule. If found, the interpreter
+will then try to match the body of the rule, recursively, until it
+finds the program axioms (rules without body). If the search procedure
+succeeds, the interpreter finds a valid substitution for the $\hat{x}$
+variables. A popular backwards-chaining programming language is
+Prolog~\cite{Colmerauer:1993:BP:154766.155362}, which has been a
+productive research language for executing logic programs in
+parallel. Researchers took advantage of Prolog's non-determinism to
+evaluate subgoals in parallel with models such as
+\emph{And-parallelism} and
+\emph{Or-parallelism}~\cite{Gupta:2001:PEP:504083.504085}.
+
+In a forward-chaining logic programming language, programs start with
+a database of facts (filled with the program's axioms) and a set of
+logical rules. The database of facts is then used to fire the
+program's rules and derive new facts that are then added to the
+database. This process is repeated recursively until the database
+reaches \emph{quiescence} and no more information can be derived from
+the program. A popular forward-chaining programming language is
+Datalog~\cite{Ramakrishnan93asurvey}.
+
+We have designed Linear Meld~(LM), a forward-chaining logic programming
+language that is specially suited for
+concurrent programming over graph structures~\cite{cruz-iclp14}. LM
+differs from Datalog-like languages because it integrates both
+classical logic and linear logic~\cite{girard-87} into the language, allowing some
+facts to be retracted and asserted in a logical fashion. Although most
+Datalog and Prolog-like programming languages allow some kind of state
+manipulation~\cite{Liu98extendingdatalog}, those features are
+extra-logical, reducing the advantages brought forward by logic programming.
+
+The roots of LM are the P2 system~\cite{Loo-condie-garofalakis-p2} and
+the original Meld
+language~\cite{ashley-rollman-derosa-iros07wksp,ashley-rollman-iclp09}. P2
+is a Datalog-like language that maps a computer network to a graph,
+where each computer node can perform computations locally and
+communicate with neighbors. Meld is inspired by the P2 system
+but adapted to the concept of massively distributed systems made of
+modular robots with a dynamic topology. LM still follows the same
+graph model of computation of Meld, which makes LM programs naturally
+concurrent since the graph of nodes can be easily partitioned to be
+executed by different threads. As a forward-chaining linear logic
+programming language, LM also shares similarities with Constraint
+Handling Rules
+(CHR)~\cite{Betz:2005kx,DBLP:journals/corr/abs-1006-3039}.  CHR is a
+concurrent committed-choice constraint language used to write
+constraint solvers. A CHR program is a set of rules and a set of
+constraints (which can be seen as facts). Constraints can be consumed
+or generated during the application of rules. Some optimizations
+used in LM such as join optimizations and the use of different data
+structures for indexing facts were inspired by research done in
+CHR~\cite{DBLP:journals/corr/cs-PL-0408025}.
+
+In this paper, we present the design and implementation of the LM virtual machine
+and compiler.
+The LM virtual machine was designed from scratch to run LM programs on
+multicore machines in an efficient manner\footnote{Source code available at \url{http://github.com/flavioc/meld}}.
+The virtual machine is multithreaded and executes byte-code that is generated
+by the LM compiler. To test our language and virtual machine we have implemented several graph algorithms, search
+algorithms and machine learning algorithms, including: belief
+propagation~\cite{Gonzalez+al:aistats09paraml}, belief propagation
+with residual splash~\cite{Gonzalez+al:aistats09paraml}, PageRank,
+graph coloring, N-Queens, shortest path, diameter estimation, map
+reduce, quick-sort, neural network training, minimax, etc.
+
+Our results show that our virtual machine is scalable and presents some interesting
+execution times when compared with other competing systems.
+The virtual machine uses a simple, but effective, work stealing algorithm
+that is able to balance the load across threads, improving scalability.
+Another important feature of our virtual machine is the dynamic indexing algorithm.
+It is a runtime algorithm that decides how to index logical facts and which
+data structure is best to use, so that database lookup and insertion time during rule application
+is effectively reduced.
+
+This remainder of the paper is organized as follows. First, we briefly
+introduce the LM language. Then, we present an overview of the virtual
+machine and describe in more detail the code organization, thread
+management, rule execution and database organization. Finally, we
+present our experiments and outline some conclusions.
+
+\makeatletter{}\newcommand{\selector}[0]{[~S~\Rightarrow~y;~BE~] \lolli HE}
+\newcommand{\comprehension}[0]{\{~\widehat{x};~BE;~SH~\}}
+\newcommand{\aggregate}[0]{[~A~\Rightarrow~y;~\widehat{x};~BE;~SH_1;~SH_2~]}
+
+\section{The LM Language}
+\label{lm_language}
+
+Linear Meld (LM) is a logic programming language that offers a
+declarative and structured way to manage mutable state.  A program consists of
+a database of facts and a set of derivation rules. The database
+includes persistent and linear facts. Persistent facts cannot be
+deleted, while linear facts can be asserted and retracted.
+
+The dynamic (or operational) semantics of LM is identical to
+Datalog. Initially, we populate the database with the \emph{program's axioms} (initial facts)
+and then determine which derivation rules can be applied by
+using the current database. Once a rule is applied, new facts can be
+derive, which are then added to the database. If a rule uses linear
+facts, they are retracted from the database. The program stops when
+\emph{quiescence} is achieved, i.e., when rules no longer apply.
+
+Each fact is a predicate on a tuple of \emph{values}, where the type
+of the predicate prescribes the types of the arguments.  LM rules are
+type-checked using the predicate declarations in the header of the
+program. LM has a simple type system that includes types such as
+\emph{node}, \emph{int}, \emph{float}, \emph{string},
+\emph{bool}. Recursive types such as \emph{list X} and \emph{pair X;
+  Y} are also allowed.  Each rule in LM has a defined priority that is
+inferred from its position in the source file.  Rules at the beginning
+of the file have higher priority.
+We consider all the new facts that have been not used yet to create a
+set of \emph{candidate rules}.
+The set of candidate rules is then applied (by priority)
+and updated as new facts are derived.
+
+\subsection{Example}
+
+\begin{figure}[ht]
+{\footnotesize
+\begin{Verbatim}[numbers=right]
+type left(node, node).
+type right(node, node).
+type linear value(node, int, string).
+type linear replace(node, int, string).
+
+// set of rules
+replace(A, K, New),
+value(A, K, Old)
+   -o value(A, K, New). // we found our key
+
+replace(A, RKey, RValue),
+value(A, Key, Value),
+!left(A, B),
+RKey < Key
+   -o value(A, Key, Value),
+      replace(B, RKey, RValue). // go left
+
+replace(A, RKey, RValue),
+value(A, Key, Value),
+!right(A, B),
+RKey > Key
+   -o value(A, Key, Value),
+      replace(B, RKey, RValue). // go right
+
+// initial configuration
+!left(@0, @1).     !right(@0, @2).
+!left(@1, @3).     !right(@1, @4). 
+!left(@2, @5).     !right(@2, @6).
+
+value(@0, 3, a).   value(@1, 1, b).
+value(@2, 5, c).   value(@3, 0, d).
+value(@4, 2, e).   value(@5, 4, f).
+value(@6, 6, g).
+
+// update key 6 to value x
+replace(@0, 6, x).
+\end{Verbatim}
+}
+\caption{LM program for replacing a key's value in a binary tree dictionary}
+\label{code:btree_replace}
+\end{figure}
+
+We now present in Fig.~\ref{code:btree_replace}, a LM program that
+implements the update operation for a binary tree dictionary
+represented as key/value pairs. We first declare the predicates (lines
+1-4) which represent the facts we are going to use. Predicate
+\texttt{left/2} and \texttt{right/2} are persistent while predicates
+\texttt{value/3} and \texttt{replace/3} are linear. Predicate
+\texttt{value/3} assigns a key/value pair to a tree node and predicate
+\texttt{replace/3} represents an update operation that updates the key
+in the second argument to the value in the third argument.
+
+The algorithm uses three rules for the three cases of updating a key's
+value: the first rule performs the update (lines 6-9); the second rule
+recursively picks the left branch for the update operation (lines
+11-16); and the third rule picks the right branch (lines 18-23). The
+initial axioms are presented in lines 26-33 and they describe the
+initial binary tree configuration, including keys and values.
+Finally, with the \texttt{replace(@0, 6, x)} axiom instantiated at the
+root node \texttt{@0} (line 36), we intend to change the value of key
+6 to value \texttt{x}. Note that when writing rules or axioms,
+persistent facts are preceded with a `\texttt{!}'.
+
+Figure~\ref{fig:btree_trace} illustrates the trace of the
+execution. Note that the program database is partitioned by the tree
+nodes using the first argument of each fact. In
+Fig.~\ref{fig:btree_trace1}, we present the database filled with the
+program's axioms. Next, we follow the right branch using rule 3 since
+$6 > 3$ (Fig.~\ref{fig:btree_trace2}). We then use the same rule
+again in Fig.~\ref{fig:btree_trace3} where we finally reach the key
+6. Here, we apply rule 1 and \texttt{value(@6, 6, g)} is updated to
+\texttt{value(@6, 6, x)} (Fig.~\ref{fig:btree_trace4}).
+
+\begin{figure*}[t]
+\centering
+\subfigure[Initial database]
+   {\label{fig:btree_trace1}\includegraphics[width=0.44\textwidth]{figures/btree_trace1.pdf}}
+\subfigure[After applying rule 3 at node $@0$]
+   {\label{fig:btree_trace2}\includegraphics[width=0.44\textwidth]{figures/btree_trace2.pdf}}
+\subfigure[After applying rule 3 at node $@2$]
+   {\label{fig:btree_trace3}\includegraphics[width=0.44\textwidth]{figures/btree_trace3.pdf}}
+\subfigure[After applying rule 1 at node $@6$]
+   {\label{fig:btree_trace4}\includegraphics[width=0.44\textwidth]{figures/btree_trace4.pdf}}
+\caption{An execution trace for the binary tree dictionary program}
+\label{fig:btree_trace}
+\end{figure*}
+
+\subsection{Syntax}
+
+Table~\ref{tbl:ast} shows the abstract syntax for rules in LM.  An LM
+program $Prog$ consists of a set of derivation rules $\Sigma$ and a
+database $D$.  Each derivation rule $R$ can be written as $BE \lolli
+HE$ where $BE$ is the body of a rule and $HE$ is the head. Rules
+without bodies are allowed in LM and they are called
+\textit{axioms}. Rules without heads are also allowed.  The body of a
+rule, $BE$, may contain linear ($L$) and persistent ($P$) \emph{fact
+  expressions} and constraints ($C$). Fact expressions are template
+facts that instantiate variables (from facts in the
+database). Variables can be used again in the body for matching and
+also in the head when instantiating facts. Constraints are boolean
+expressions that must be true in order for the rule to be
+fired. Constraints use variables from fact expressions and are built
+using a small functional language that includes mathematical
+operations, boolean operations, external functions and literal values.
+The head of a rule, $HE$, contains linear ($L$) and persistent ($P$)
+\emph{fact templates} which are uninstantiated facts to derive new
+facts. Head expressions may use the variables instantiated in the
+body. The head can also have \emph{comprehensions} ($CE$) and
+\emph{aggregates} ($AE$).
+
+\begin{table}[ht]
+\centering
+{\scriptsize
+\begin{tabular}{llcl}
+Program          & $Prog$ & $::=$ & $\Sigma, D$ \\
+Set of Rules     & $\Sigma$ & $::=$ & $\cdot \; | \; \Sigma, R$\\
+Database         & $D$ & $::=$ & $\Gamma; \Delta$ \\
+Rule             & $R$ & $::=$ & $BE \lolli HE \; | \; \forall_{x}. R$ \\
+Body Expression  & $BE$ & $::=$ & $L \; | \; P \; | \; C \; | \; BE, BE \; | \; \exists_{x}. BE \; | \; 1$\\
+Head Expression  & $HE$ & $::=$ & $L \; | \; P \; | \; HE, HE \; | \; CE \; | \; AE \; | \; 1$\\
+Linear Fact      & $L$ & $::=$ & $l(\hat{x})$\\
+Persistent Fact  & $P$ & $::=$ & $\bang p(\hat{x})$\\
+Constraint       & $C$ & $::=$ & $c(\hat{x})$ \\
+Comprehension    & $CE$ & $::=$ & $\comprehension$ \\
+Aggregate        & $AE$ & $::=$ & $\aggregate$ \\
+Operations       & $A$ & $::=$ & $\mathtt{min} \; | \; \mathtt{max} \; | \; \mathtt{sum} \; | \; \mathtt{count}$ \\
+Sub-Head         & $SH$ & $::=$ & $L \; | \; P \; | \; SH, SH \; | \; 1$\\
+Linear Facts     & $\Delta$ & $::=$ & $\cdot \; | \; \Delta, l(\hat{t})$ \\
+Persistent Facts & $\Gamma$ & $::=$ & $\cdot \; | \; \Gamma, \bang p(\hat{t})$ \\
+\end{tabular}
+}
+\caption{Abstract syntax of LM}
+\label{tbl:ast}
+\end{table}
+
+We created the concept of comprehensions to be used when the
+consumption of a linear fact should generate a set of facts
+accordingly to the current contents of the database. In a
+comprehension $\comprehension$, $\widehat{x}$ is a list of variables,
+$BE$ is the body of the comprehension and $SH$ is the head. The body
+$BE$ is used to generate all possible combinations for the head $SH$,
+according to the facts in the database. The following program shows a
+simple example that uses comprehensions:
+
+{\footnotesize
+\begin{Verbatim}
+!edge(@1, @2).
+!edge(@1, @3).
+iterate(@1).
+
+iterate(A)
+   -o {B | !edge(A, B) | perform(B)}.
+\end{Verbatim}
+}
+
+When the rule is fired, we consume \texttt{iterate(@1)} and then
+generate the comprehension. Here, we iterate through all the
+\texttt{edge/2} facts that match \texttt{!edge(@1, B)}, which are
+\texttt{!edge(@1, @2)} and \texttt{!edge(@1, @3)}. For each fact, we
+then derive \texttt{perform(B)}, namely \texttt{perform(@2)} and
+\texttt{perform(@3)} in this example.
+
+Another useful feature is the ability to reduce several facts into a
+single fact. For that, we have aggregates, a special kind of sub-rule
+that works very similarly to comprehensions. In a aggregate
+$\aggregate$, $A$ is the aggregate operation, $\widehat{x}$ is the
+list of variables introduced in $BE$, $SH_1$ and $SH_2$ and $y$ is the
+variable in the body $BE$ that represents the values to be aggregated
+using $A$. Like comprehensions, we use $\widehat{x}$ to try all the
+combinations of $BE$, but, in addition to deriving $SH_1$ for each
+combination, we aggregate the values represented by $y$ and derive
+$SH_2$ only once using $y$. As an example, consider the following
+program:
+
+{\footnotesize
+\begin{Verbatim}
+price(@1, 3).
+price(@1, 4).
+price(@1, 5).
+count-prices(@1).
+
+count-prices(A)
+   -o [sum => P | . | price(A, P) | 1 | total(A, P)].
+\end{Verbatim}
+}
+
+By applying the rule, we consume \texttt{count-prices(@1)} and derive
+the aggregate which consumes all the \texttt{price(@1, P)} linear
+facts. These are summed up and \texttt{total(@1,~12)} is derived. LM
+provides several aggregate operations, including the \emph{minimum},
+\emph{maximum}, \emph{sum} and \emph{count}. Note that the \texttt{.} syntax indicates
+an empty list of variables and \texttt{1} is borrowed from linear logic
+and represents an empty derivation.
+
+Comprehensions and aggregates are logically justified by the underlying
+proof system of the language. Our proof system is extended with greatest
+fixed points~\cite{Baelde:2012:LGF:2071368.2071370}, which allows us to describe
+recursive definitions such as comprehensions or aggregates.
+A more detailed description is found in~\cite{cruz-iclp14}.
+
+\subsection{Concurrency}
+
+LM is at its core a concurrent programming language. The database of
+facts can be seen as a graph data structure where each node contains a
+fraction of the database.  To accomplish this, we force the first
+argument of each predicate to be typed as a \emph{node}. We then
+restrict the derivation rules to only manipulate facts belonging to a
+single node.  However, the expressions in the head may refer to other
+nodes, as long as those nodes are instantiated in the body of the
+rule.
+
+Due to the restrictions on LM rules, nodes are able to run rules
+independently without using other node's facts. Node computation
+follows a \emph{don't care} or \emph{committed choice} non-determinism
+since any node can be picked to run as long as it contains enough
+facts to fire a derivation rule.  Facts coming from other nodes will
+arrive in order of derivation but may be considered partially and
+there is no particular order among the neighborhood. To improve
+concurrency, the programmer is encouraged to write rules that take
+advantage of the non-deterministic nature of execution since too much
+determinism will naturally imply less scalability and more synchronization
+when executing the programs.
+
+\makeatletter{}\section{The Virtual Machine}
+\label{virtual_machine}
+
+We have developed a compiler that compiles LM programs to byte-code
+and a multithreaded virtual machine (VM) using POSIX threads
+that runs the byte-code.
+
+\subsection{Threads}
+
+A key goal of our design is to keep the threads as busy as possible
+and to reduce inter-thread communication. When the VM starts, it reads
+the byte-code file and starts all threads. Initially, the VM will
+partition the application graph of $N$ nodes into $T$ subgraphs (the
+number of threads) and then each thread will work on their own
+subgraph. Reduction of communication between nodes in different threads
+is achieved by first ordering the node addresses present in the code in such a way that
+connected nodes are clustered together and then partitioning them to
+threads. During compilation, we take note of predicates that are used
+for communication (arguments with type \emph{node}) and
+then build a graph of nodes from them. The nodes of
+the graph are then ordered by using a breadth-first search algorithm
+that changes the nodes of addresses to the domain $[0, n[$, where $n$
+is the number of nodes. Once the VM starts, we simply partition the
+range $[0, n[$ into $T$ subgraphs.
+
+During execution, threads can steal nodes of other threads to keep
+themselves busy. The load balancing aspect of the system is performed
+by our work scheduler that is based on a simple work stealing
+algorithm. The pseudo-code for the main thread loop is shown in
+Fig.~\ref{code:work_loop}. In each round, a thread inspects its work
+queue for active nodes with new candidate rules and, if there is any,
+procedure \texttt{process\_node()} executes it. If the work queue is
+empty, the thread attempts to steal one node from another
+thread. Starting from a random thread, it cycles through all the
+threads to find one active node. Eventually, there will be no more
+work to do and the threads will go idle. There is a global atomic
+counter, a global boolean flag and one boolean flag for each thread
+that are used to detect termination. Once a thread goes idle, it
+decrements the global counter and changes its flag to idle. If the
+counter reaches zero, the global flag is set to idle. Since every
+thread will be busy-waiting and checking the global flag, they will
+detect the change and stop executing.
+
+\begin{figure}[ht]
+{\footnotesize
+\begin{Verbatim}
+void thread_work_loop(thread_id tid)
+   while (true)
+      node = pop_node_from_work_queue(tid)
+      if (node)
+         process_node(tid, node)
+      else
+         // need to steal a node
+         target = random(NUM_THREADS)
+         for (i = 0; i < NUM_THREADS && !node; i++)
+            target = (target + 1)             node = steal_node_from_thread(target)
+            if (node) break
+         if (!node)
+            // try to terminate
+            become_idle(tid)
+            if (synchronize_termination(tid))
+               return
+            become_active(tid)
+\end{Verbatim}
+}
+\caption{Thread work loop}
+\label{code:work_loop}
+\end{figure}
+
+\subsection{Nodes}
+
+\begin{figure*}[t]
+\centering
+\includegraphics[width=\textwidth]{figures/overview.pdf}
+\caption{Layout of the virtual machine}
+\label{fig:overview}
+\end{figure*}
+
+Figure~\ref{fig:overview} presents the layout of our virtual machine
+for a program with six nodes and two running threads. Each thread
+space includes the nodes owned by the thread (the dotted arrows
+represent the edges between nodes) and a \emph{Work Queue}, a single
+linked list containing \emph{active nodes}, i.e., nodes that have new
+facts to process. Initially, the \emph{Work Queue} is filled with all
+the nodes of the thread in order to derive the
+axioms. Figure~\ref{fig:overview} also illustrates the internal
+structure layout of a node, which includes: the database of linear
+facts (\emph{Linear DB}); the database of persistent facts
+(\emph{Persistent DB}); the rule matching structures (\emph{Rule
+Engine}); and an auxiliary buffer for storing intermediate facts
+coming from other threads (\emph{Fact Buffer}).
+
+Whenever a new fact is derived through rule derivation, we need to
+update the data structures for the corresponding node. This is trivial
+if the thread that derived the fact also owns the node. If that is not
+the case, then we have to synchronize since multiple threads might be
+updating the same node's data structures. We added a lock and a
+boolean flag to each node to protect the access to its data
+structures. When the flag is activated, it means that the node is
+currently being executed by the owning thread. For example, in
+Fig.~\ref{fig:overview}, if thread 2 derives a fact to
+node \texttt{@1} (owned by thread 1), then thread 2 checks the node's
+flag and if not activated, will lock node \texttt{@1} and perform the
+required updates (\emph{New fact (1)}). If the flag is activated, it
+will not touch the main node data structures, but instead will add the
+new fact to \emph{Fact Buffer} (\emph{New fact (2)}). The facts stored
+in \emph{Fact Buffer} will then be processed whenever the
+corresponding node's flag becomes active.
+
+There is another thread interaction that might happen during fact
+derivation if the node receiving a new fact is not active. In such
+case, the sending thread needs to activate the node by pushing it to
+the \emph{Work Queue} of the target thread. For example, consider
+again the situation in which thread 2 sends a new fact to
+node \texttt{@1}. If node \texttt{@1} is not active, then thread 2
+also needs to activate it by pushing it to the \emph{Work Queue} of
+thread 1. After this synchronization point, if the target thread is
+currently idle, it will become active and with a new node to process.
+
+\subsection{Database Data Structures}
+\label{sec:database}
+
+We said before that LM rules are constrained by the first argument and
+that each node has its own database of linear and persistent
+facts. Moreover, since only one thread at a time will be using a
+node's database, we do not need to deal with synchronization
+issues. Note also that the first argument of each fact is not stored.
+
+The database must be implemented efficiently because during matching
+of rules we need to restrict the facts using a given \emph{match
+object}, which fixes arguments of the target predicate to instantiated
+values. Each node's database is implemented using three kinds of data
+structures:
+
+\begin{itemize}
+\item \emph{Trie Data Structures} are used exclusively to store 
+  persistent facts. Tries are trees where facts are indexed by common
+  prefix arguments.
+\item \emph{Doubly Linked List Data Structures} are used to store 
+  linear facts. We use a double linked list because it is a very
+   efficient way to add and remove facts.
+\item \emph{Hash Table Data Structures} are used to improve lookup when 
+  linked lists are too long and when we need to do search filtered by
+  a fixed argument. The virtual machine decides which arguments are
+  best to be indexed (see Section~\ref{indexing}) and then uses a hash table
+  indexed by the appropriate argument. If we need to go through all
+  the facts, we just iterate through all the facts in the table. For
+  collisions, we use the above doubly linked list data structure.
+\end{itemize}
+
+Figure~\ref{fig:hash_table} shows an example for a hash table data
+structure for a \texttt{a(int,int)} predicate with 3 linear facts
+indexed by the second argument and stored as a doubly linked list in
+bucket \texttt{2}. Each linear fact contains the regular list
+pointers, a \texttt{flags} field and the fact arguments. Those are all
+stored continuously to improve data locality. One use of
+the \texttt{flags} field is to mark that a fact is already being
+used. For example, consider the rule:
+
+{\footnotesize
+\begin{Verbatim}
+a(A,B), a(C,D) -o ...
+\end{Verbatim}
+}
+
+When we first pick a fact for \texttt{a(A, B)} from the hash
+table, we mark it as being used in order to ensure that, when we
+retrieve facts for \texttt{a(C, D)}, the first one cannot be used
+twice since that would violate linearity.
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=0.4\textwidth]{figures/hash_table.pdf}
+\caption{Hash table and doubly linked data structures for 
+  a \texttt{a(int,int)} predicate}
+\label{fig:hash_table}
+\end{figure}
+
+\subsection{Rule Engine}
+\label{rule_engine}
+
+The rule engine decides which rules may need to be executed while
+taking into account rule priorities. Figure~\ref{fig:rule_engine}
+shows the rule engine data structures in more detail. There are 4 main
+data structures for scheduling rule execution: \texttt{Rule Queue} is
+the bitmap representing the rules scheduled to run; \texttt{Active
+Bitmap} contains the rules that have enough facts to be
+fired; \texttt{Dropped Bitmap} contains the rules that must be dropped
+from \texttt{Rule Queue} and \texttt{Active Bitmap} if the rule being
+executed succeeds; and \texttt{Predicates Count} counts the number of
+facts per predicate. To understand how our engine works, consider the
+following set of facts and rules:
+
+{\footnotesize
+\begin{Verbatim}
+a.
+e(0).
+
+a, e(1) -o b.  // rule 1
+a -o c.        // rule 2
+b -o d.        // rule 3
+e(0) -o f.     // rule 4
+c -o e(1).     // rule 5
+\end{Verbatim}
+}
+
+\begin{figure*}[t]
+\centering
+\includegraphics[width=11cm]{figures/rule_queue.pdf}
+\caption{Rule engine data structures (a) before and (b) after applying 
+  the rule \texttt{a -o c.}}
+\label{fig:rule_engine}
+\end{figure*}
+
+Since we have facts for predicates \texttt{a/0} and \texttt{e/1},
+the \texttt{Active Bitmap} starts with rules 1, 2 and 4 marked as
+having enough facts to be fired. The \texttt{Rule Queue} bitmap also
+starts with the same three rules. In order to pick rules for
+execution, we take the rule corresponding to the least significant bit
+from the \texttt{Rule Queue} bitmap, initially the first
+rule \texttt{a, e(1) -o b}. However, since we don't have
+fact \texttt{e(1)}, this rule fails and we execute the second
+rule \texttt{a -o c}. Figure~\ref{fig:rule_engine}(a) shows the rule
+engine data structures at that point.
+
+Because the derivation for the second rule succeeds, we will consume
+fact \texttt{a} and derive fact \texttt{c}. We thus update
+\texttt{Predicates Count} accordingly, mark the first 
+and second rules in \texttt{Dropped Bitmap} since such rules are no
+longer applicable (\texttt{a} was consumed) and mark the fifth rule
+in \texttt{Active Bitmap} since \texttt{c} was derived. Finally, to
+update the \texttt{Rule Queue}, we remove the bits marked
+in \texttt{Dropped Bitmap} and add the active rules marked
+in \texttt{Active Bitmap} that use the newly derived predicates, rule
+5 in this case. In the continuation, the engine will schedule the
+fourth and fifth rules to run. Figure~\ref{fig:rule_engine}(b) shows
+the rule engine data structures at that point.
+
+Note that every node in the program has the same set of data
+structures presented in Fig.~\ref{fig:rule_engine}. We use 32 bits
+integers to implement the 3 bitmaps and an array of 16 bits integers to
+count facts.
+
+We do a small optimization to reduce the number of derivations of
+persistent facts and, for that, we divide the program rules into two
+sets: \emph{persistent rules} and \emph{non persistent rules}.
+Persistent rules are rules where only persistent facts are
+involved. We compile such rules incrementally, i.e., we attempt to
+fire all rules where a persistent fact is used. This is called
+the \emph{pipelined semi-naive} evaluation and it originated in the P2
+system~\cite{Loo-condie-garofalakis-p2}. This evaluation method avoids
+excessing re-derivations of the same fact. The order of derivation
+does not matter for those rules, since only persistent facts are used.
+
+\subsection{Rule Execution}
+
+A byte-code file contains meta-data about the program's predicates,
+initial nodes, partitioning information, and code for each rule. Each
+VM thread has 32 registers that are used during rule
+execution. Registers can store facts, integers, floats, node addresses
+and pointers to runtime data structures (lists and structures). When
+registers store facts, we can reference fields in the fact through the
+register.
+
+Consider the example in Fig.~\ref{fig:byte_code} that shows the LM
+byte-code for the rule \texttt{!a(X,Y), b(X,Z), c(X,Y) -o
+d(Y)}. Consider also a database with the
+facts \texttt{!a(1,2)}, \texttt{!a(2,3)}, \texttt{b(1,3)},
+\texttt{b(5,3)}, \texttt{c(1,2)}, \texttt{c(1,3)}, \texttt{c(5,3)}. Rule
+execution for this rule and facts proceeds in a series of recursive
+loops, as follows. The first loop retrieves an iterator for the
+persistent facts of \texttt{!a/2} and moves the first valid
+fact, \texttt{!a(1,2)}, to register 0. The inner loop retrieves linear
+facts that match \texttt{b(1,Z)} (from the
+\emph{join constraint}) and moves \texttt{b(1,3)} to register 1. The 
+final loop moves \texttt{c(1,2)} to register 2 and the body of the
+rule is successfully matched. Next, we derive \texttt{d(2)},
+where \texttt{2} comes from register 0.
+
+\begin{figure}[ht]
+{\footnotesize
+\begin{Verbatim}
+PERSISTENT ITERATE a MATCHING TO reg 0
+   LINEAR ITERATE b MATCHING TO reg 1
+      (match).0=0.0                  // match argument X
+      LINEAR ITERATE c MATCHING TO reg 2
+         (match).0=0.0               // match argument X
+         (match).1=0.1               // match argument Y
+         ALLOC d TO reg 3
+         MVFIELDFIELD 0.1 TO 3.0       // get argument Y
+         ADDLINEAR reg 3                  // derive d(Y)
+         REMOVE reg 2
+         REMOVE reg 1
+         TRY NEXT
+      NEXT
+   NEXT
+RETURN
+\end{Verbatim}
+}
+\caption{LM byte-code for rule \texttt{!a(X,Y), b(X,Z), c(X,Y) -o d(Y)}}
+\label{fig:byte_code}
+\end{figure}
+
+In case of failure, we jump to the previous loop in order to try the
+next candidate fact. In case of rule success, the head is derived and
+we should backtrack to the inner most \emph{valid loop}, i.e., the
+older loop that uses linear facts or, if there are no linear facts
+involved, to the previous loop. We need to jump to a valid loop
+because we may have loops with linear facts that are now invalid. In
+our example, we would jump to the loop of \texttt{b(X,Z)} and
+not \texttt{c(X,Y)}, since \texttt{b(1,3)} was consumed.
+
+As an optimization, the compiler re-orders the fact expressions used
+in the body in order to make execution more efficient. For example, it
+forces the join constraints in rules to appear first so that matching
+will fail sooner rather than later. It also does the same for
+constraints. Note also that for every loop, the compiler adds
+the \emph{match object}, which contains information about which
+arguments need to match, so that runtime matching be efficient.
+
+Our compiler also detects cases where we re-derive a linear fact with
+new arguments. For example, as shown in Fig.~\ref{code:update}, the LM
+byte-code for rule \texttt{a(N) -o a(N+1)} will compile to code that
+reuses/updates the old \texttt{a(N)} fact when deriving the
+new \texttt{a(N+1)} fact.
+
+\begin{figure}[ht]
+{\footnotesize
+\begin{Verbatim}
+LINEAR ITERATE a MATCHING TO reg 0
+   MVFIELDREG 0.0 TO reg 1           // initial argument
+   MVINTREG INT 1 TO reg 2
+   reg 1 INT PLUS reg 2 TO reg 3
+   MVREGFIELD reg 3 TO 0.0      // reuse/update argument
+   UPDATE reg 0
+   TRY NEXT
+RETURN
+\end{Verbatim}
+}
+\caption{\small{LM byte-code for rule \texttt{a(N) -o a(N+1)}}}
+\label{code:update}
+\end{figure}
+
+\subsection{Indexing}
+\label{indexing}
+
+To improve fact lookup, the VM employs a fully dynamic mechanism to
+decide which argument may be optimal to index.  The algorithm is
+performed in the beginning of the execution and empirically tries to
+assess the argument of each predicate that more equally spreads the
+database across the values of the argument.  A single thread performs
+the algorithm for all predicates.
+
+The indexing algorithm is performed in three main steps. First, it
+gathers statistics of lookup data by keeping a counter for each
+predicate's argument.  Every time a fact search is performed where
+arguments are fixed to a value, the counter of such arguments is
+incremented. This phase is performed during rule execution for a small
+fraction of the nodes in the program.
+
+The second step of the algorithm then decides the candidate arguments
+of each predicate.  If a predicate was not searched with any fixed
+arguments, then it will be not indexed.  If only one argument was
+fixed, then such argument is set as the indexing argument. Otherwise,
+the top 2 arguments are selected for the third phase,
+where \emph{entropy statistics} are collected dynamically.
+
+During the third phase, each candidate argument has an entropy score.
+Before a node is executed, the facts of the target predicate
+are used in the following formula applied for the two arguments:
+
+{\tiny
+\[
+Entropy(A, F) = - \sum_{v \in values(F, A)} \frac{count(F, A = v)}{total(F)} \log_2 \frac{count(F, A = v)}{total(F)}
+\]
+}
+
+\noindent
+where $A$ is the target argument, $F$ is the set of linear facts
+for the target predicate, $values(F, A)$ is set of values of the
+argument $A$, $count(F, A = v)$ counts the number of linear facts
+where argument $A$ is equal to $v$ and $total(F)$ counts the number of
+linear facts in $F$.  The entropy value is a good metric because it
+tells us how much information is needed to describe an argument.  If
+more information is needed, then that must be the best argument to
+index.
+
+For one of the arguments to score, $Entropy(A, F)$ multiplied by the
+number of times it has been used for lookup must be larger than the
+other argument. The argument with the best score is selected and then
+a global variable called \texttt{indexing\_epoch} is updated.  In
+order to convert the node's linked lists into hash tables, each node
+also has a local variable called \texttt{indexing\_epoch} that is
+compared to the global variable in order to rebuild the node database
+according to the new indexing information.
+
+Our VM also dynamically resizes the hash table if necessary. When the
+hash table becomes too dense, it is resized to the double. When it
+becomes too sparse, it is reduced in half or simply transformed back
+into a doubly linked list. This is done once in a while, before a node
+executes.
+
+We have seen very good results with this scheme. For example, for the
+all-pairs shortest paths program, we obtained a 2 to 5-fold
+improvement in sequential execution time.  The overhead of dynamic
+indexing is negligible since programs run almost as fast as if the
+indices have been added from the start.
+
+\subsection{Runtime Data Structures}
+
+LM also supports recursive types such as lists and pairs. These
+complex data structures are stored in the heap of the VM and are
+managed through reference counting. For instance, each list is
+a \emph{cons cell} with 3 fields: \texttt{tail}, the pointer to the
+next element of the list; \texttt{head}, the element stored by this
+element of the list; and \texttt{refs} that counts the number of
+pointers to this list element in the VM. The list is deleted from the
+heap whenever \texttt{refs} is decremented to zero.
+
+We avoid garbage collection schemes since objects are created and discarded
+in very specific points of the virtual machine and our objects cannot contain circular references.
+A reference counting mechanism is thus more appropriate than a parallel garbage collector which
+would entail pausing the execution of the program to garbage collect all the unused objects.
+
+\makeatletter{}\newcommand{\plotsize}{0.45\textwidth}
+
+\section{Experimental Results}
+\label{results}
+
+This section presents initial results for our VM. First, we present a
+comparison with similar programs written in other programming
+languages in order to show evidence that our VM is viable. Then, we
+present scalability results in order to show that LM programs can take
+advantage of multicore architectures.
+
+For our experimental setup, we used a machine with 32 (2x16) Core AMD
+Opteron (tm) Processor 6274 $@$ 2.2 GHz with 32 GBytes of RAM memory
+and running the Linux kernel 3.8.3-1.fc17.x86\_64.  We compiled our VM
+using GCC 4.7.2 (g++) with the flags \texttt{-O3 -std=c+0x
+  -march=x86-64}. We run all experiments 3 times and averaged the
+execution time.
+
+\subsection{Absolute Execution Time}
+
+To put our VM in perspective, we first compare it in terms of absolute
+execution time with other competing systems using a single thread.
+
+In Table~\ref{tbl:comp_nqueens}, we compare LM's version of the
+classic N-Queens puzzle against 3 other versions: a straightforward
+sequential program implemented in C using backtracking; a sequential
+Python implementation~\cite{vanRossum:1995:PRM}; and a Prolog
+implementation executed in YAP
+Prolog~\cite{DBLP:journals/corr/abs-1102-3896}, an efficient
+implementation of Prolog. Numbers less than 1 mean that LM is faster
+and larger than 1 mean that LM is slower. We can observe that LM
+easily beats Python, but is 5 to 10 times slower than YAP Prolog and
+around 15 times slower than C. Note however that, as we will see next,
+if we use at least 16 threads in LM, we can beat the sequential
+implementation written in C.
+
+\begin{table}[ht]
+\centering
+{\begin{tabular}{c|c|c|c}
+\textbf{Problem} & \multicolumn{3}{c}{\textbf{System}} \\
+\textbf{Size} & \textbf{C} & \textbf{Python} & \textbf{YAP Prolog} \\
+\hline\hline
+\textbf{10x10} & 16.92 & 0.62 & 5.42 \\
+\textbf{11x11} & 21.59 & 0.64 & 6.47 \\
+\textbf{12x12} & 10.32 & 0.73 & 7.61 \\
+\textbf{13x13} & 14.35 & 0.88 & 10.38 \\
+\end{tabular}}
+\caption{Comparing the absolute execution times (LM/System) for the
+  N-Queens program}
+\label{tbl:comp_nqueens}
+\end{table}
+
+In Table~\ref{tbl:comp_bp}, we compare LM's Belief Propagation (BP)
+program, a machine learning algorithm to denoise images, against a
+sequential C, Python and GraphLab~\cite{GraphLab2010} version of the
+algorithm. GraphLab is a parallel C++ library used to solve
+graph-based problems in machine learning. C and GraphLab perform about
+the same since they are both compiled to machine code, although
+the GraphLab version is highly optimized to run on multicore architectures.
+Python runs very slowly since it is a
+dynamic programming language and BP has many mathematical
+computations. We should note, however, that LM's version uses some
+external functions written in C++ in order to improve execution time,
+therefore the comparison is not totally fair.
+
+\begin{table}[ht]
+\centering
+{\begin{tabular}{c|c|c|c}
+\textbf{Problem} & \multicolumn{3}{c}{\textbf{System}} \\
+\textbf{Size} & \textbf{C} & \textbf{Python} & \textbf{GraphLab} \\
+\hline\hline
+\textbf{10}  & 1.00 & 0.03 & 1.00 \\
+\textbf{50}  & 1.77 & 0.04 & 1.73 \\
+\textbf{200} & 1.99 & 0.05 & 1.79 \\
+\textbf{400} & 2.00 & 0.04 & 1.80 \\
+\end{tabular}}
+\caption{Comparing the absolute execution times (LM/System) for the
+  Belief Propagation program}
+\label{tbl:comp_bp}
+\end{table}
+
+We also compared a LM's version of the PageRank program against a
+similar GraphLab version and LM showed to be around 4 to 6 times
+slower. Our worse results were obtained for the all-pairs shortest
+distance algorithm where a LM's version of the problem was around 50
+times slower than a C sequential implementation of the Dijkstra
+algorithm, but almost twice as fast when compared to the same
+implementation in Python.
+
+\subsection{Scalability}
+
+\begin{figure*}[ht]
+\centering
+\subfigure[PageRank using a graph of web pages with around 12,000 nodes and 292,000 edges]
+   {\label{exp:pagerank}\includegraphics[width=\plotsize]{figures/results-pagerank-search-engines.pdf}}
+\hspace{0.5cm}
+\subfigure[GGC using a random graph with 2,000 nodes and 600,000 edges]
+   {\label{exp:ggc}\includegraphics[width=\plotsize]{figures/results-ggc.pdf}}
+\caption{Experimental results for the PageRank and GGC  algorithms}
+\end{figure*}
+
+\begin{figure*}[ht]
+\centering
+\subfigure[Shortest Distance for a graph with around 5,000 nodes and 13,000 edges]
+   {\label{exp:sd}\includegraphics[width=\plotsize]{figures/results-shortest-uspowergrid.pdf}}
+\hspace{0.5cm}
+\subfigure[MiniMax algorithm for the Tic-Tac-Toe game (complete tree)]
+   {\label{exp:minimax}\includegraphics[width=\plotsize]{figures/results-minimax.pdf}}
+\caption{Experimental results for the Shortest Distance and MiniMax algorithm}
+\end{figure*}
+
+\begin{figure*}[ht]
+\centering
+\subfigure[N-Queens program (13x13~board)]
+   {\label{exp:13queens}\includegraphics[width=\plotsize]{figures/results-13queens.pdf}}
+\hspace{0.5cm}
+\subfigure[Belief Propagation (400x400~image)]
+   {\label{exp:bp}\includegraphics[width=\plotsize]{figures/results-bp.pdf}}
+\caption{Experimental results for the N-Queens and Belief Propagation programs}
+\end{figure*}
+
+In this section we measure the scalability of the VM along with the
+performance gains due to work stealing and dynamic indexing.  For
+this purpose, we used 4 configurations for the VM: \textbf{WI}, the
+full configuration that includes work stealing and dynamic indexing;
+\textbf{WN}, with work stealing but without dynamic
+indexing; \textbf{NI}, with indexing but without work stealing; and
+\textbf{NN}, without work stealing and without dynamic indexing.  We
+ran each configuration using 1, 2, 4, 6, 8, 10, 12, 14 and 16 threads
+and compared the run time against the sequential execution (1 thread)
+of \textbf{WI}. We used the following set of programs:
+
+\begin{itemize}
+\item PageRank implements a PageRank algorithm without synchronization
+  between iterations. Every time a node sends a new rank to its
+  neighbors and the change was significant, the neighbors are
+  scheduled to recompute their ranks.
+\item Greedy Graph Coloring (GGC) colors nodes in a graph so that no
+  two adjacent nodes have the same color. We start with a small number
+  of colors and then we expand the number of colors when we cannot
+  color the graph.
+\item Shortest Distance (SD) computes the shortest distance of all
+  nodes to all nodes.
+\item MiniMax, the AI algorithm for selecting the best player move in a game of Tic-Tac-Toe.
+\item N-Queens, the classic puzzle for a 13x13 board.
+\item Belief Propagation, a machine learning algorithm to denoise
+  images.
+\end{itemize}
+
+The PageRank results are shown in Fig.~\ref{exp:pagerank}. We used a
+search engine graph of 12,000 webpages\footnote{Available from
+  \url{http://www.cs.toronto.edu/~tsap/experiments/download/download.html}}. Since
+this dataset follows the power law, that is, there is a small number
+of pages with a lots of links (1\% of the nodes have 75\% of the
+edges), it can be difficult to parallelize. Our results show that the
+VM is able to scale the program with up to 14 threads.  We also notice
+the huge performance drop when we run the VM without work
+stealing. Dynamic indexing is also an advantage, since it detects that
+the facts for the pagerank of neighboring nodes need to be indexed
+efficiently.
+
+Figure~\ref{exp:ggc} presents the results for the GGC program with a
+random dataset of 2,000 nodes with an uniform distribution of edges.
+There is a slight drop in scalability as the number of threads goes
+up, but the VM is still capable of reducing the run time. We note that
+in this program, the work available is reduced as the graph becomes
+increasingly colored.
+
+In Fig.~\ref{exp:sd} we show the results for the Shortest Distance program.
+We attain a 13-fold speedup for 16 threads with both work stealing and dynamic
+indexing (\textbf{WI}). We note that indexing is more advantageous than work stealing
+because indexing the distance facts according to the source node is more crucial
+than improved load balancing.
+
+The results for the MiniMax algorithm are presented in Fig.~\ref{exp:minimax}.
+MiniMax is very different than the other algorithms because the graph of nodes is dynamic
+and is created during program execution. The load balancing is also problematic since there
+is little work to do in the initial and final phases of the algorithm. Still, our VM has decent
+performance, with almost a 7-fold speedup for 14 threads. The scalability drops with 16 threads
+but we think that is due to the simplicity of the current work stealing algorithm.
+Dynamic indexing has no effects in this program.
+
+The results for the N-Queens program are shown in
+Fig.~\ref{exp:13queens}. The program is not regular since computation
+starts at the top of the grid and then rolls down, until only the last
+row be doing computation. Because the number of valid states for the
+nodes in the upper rows is much less than the nodes in the lower rows,
+this may potentially lead to load balancing problems. The results show
+that our system is able to scale well. When work stealing is left out
+(\textbf{NI} and \textbf{NN}), we see a serious drop in performance with 16 threads.
+
+Finally, we shown the results for the Belief Propagation~(BP) program
+in Fig.~\ref{exp:bp}. BP is a regular and asynchronous program and
+benefits (as expected) from having multiple threads executing since
+the belief values of each node will converge faster.  The super-linear
+results prove this assertion.
+
+\subsection{Memory Statistics}
+
+In this section, we present the amount of memory used by the VM after the
+programs presented in the previous section complete. We also count the number of
+facts stored in the database. This serves as an indication of
+how much memory is needed in terms of the number of facts stored in the database.
+Results are shown in Table~\ref{tbl:memory}.
+
+\begin{table}[ht]
+\centering
+{\begin{tabular}{c|c|c|c}
+\textbf{Program} & \textbf{Facts} & \textbf{Memory} & \textbf{Per Fact} \\
+\hline\hline
+PageRank & 1,180,603 & 203,832 KB & 176 bytes \\
+GGC & 2,363,536 & 292,682 KB & 127 bytes \\
+SD & 502,347 & 45,387 KB & 92 bytes \\
+MiniMax & 3 & 886,443 KB & 295,481 KB \\
+N-Queens & 74,557 & 55,408 KB & 760 bytes \\
+BP & 2,235,200 & 617,417 KB & 283 bytes \\
+\hline\hline
+\multicolumn{3}{l}{Average (without MiniMax) } & 288 bytes \\
+\end{tabular}}
+\caption{Memory usage of programs after completion using 1 thread.}
+\label{tbl:memory}
+\end{table}
+
+The most unexpected result in Table~\ref{tbl:memory} is that of the MiniMax program. There is a
+very low number of facts (that indicate the final player decision) and a huge
+amount of memory. How can this be explained? We note that MiniMax builds a huge
+decision tree of nodes with almost $9!$ leaves. Although these nodes do not participate
+in the computation after the MiniMax result is computed for them, the VM does not garbage collect
+nodes that do not contain facts and that do not have external references from other nodes.
+This is a potential improvement that can be done in the future.
+
+PageRank, GGC and SD show a low memory usage per fact. This makes sense since the predicates
+used in those programs are relatively simple with only a few arguments.
+This changes with the N-Queens problem where the predicates have lists representing the valid
+positions of the queens in the board. Storing lists is far more expensive than storing integral
+values such as integers or floating point numbers. The same happens with BP because
+the belief values are stored as lists of floating point numbers.
+
+\makeatletter{}\section{Related Work}
+\label{related_work}
+
+Virtual machines are a popular technique for implementing interpreters
+for high level programming languages.  Due to the increased
+availability of parallel machines and distributed architectures,
+several machine models have been developed with parallelism in
+mind~\cite{Kara:1997:AMM:265274}.  One example of such machine is the
+Parallel Virtual Machine (PVM)~\cite{Sunderam90pvm:a}, which serves as
+an abstraction to program heterogeneous computers as a single
+machine. Another important machine is the Threaded Abstract Machine
+(TAM)~\cite{CullerGSvE93,goldstein-tr94}, which defines a
+self-scheduled machine language of parallel threads where a program is
+represented as conventional control flow.
+
+Prolog, the most prominent logic programming language, has a rich
+history of virtual machine research centered around the Warren
+Abstract Machine (WAM)~\cite{AICPub641:1983}. 
+Prolog is naturally parallel because several clauses for the same goal
+(AND-parallelism) or all goals in a clause (OR-parallelism) can be
+tried in parallel. Different abstract machines for AND-parallelism has
+been developed on top of the
+WAM~\cite{Hermenegildo:1986:AMB:913061,Lin:1988:AEL:900478}.  For
+OR-parallelism we have several models such as: the SRI
+model~\cite{Warren:1987:OEM:67683.67699}, the Argonne
+model~\cite{ButlerDLOOS88}, the MUSE model~\cite{Ali:1990fk} and the
+BC machine~\cite{Ali88}. While all those models have been developed
+using the WAM, there are also some parallel machines totally different
+from the WAM, such as the PPAM~\cite{Kacsuk:1990:EMP:533578}, which is
+based on a data-flow model.
+
+Several linear logic programming languages have been developed in the past~\cite{Miller85anoverview}.
+Lolli, a programming language based on a fragment of intuitionistic linear logic~\cite{Hodas94logicprogramming}, 
+proves goals by lazily managing the context of linear resources during backward-chaining proof search.
+LolliMon~\cite{Lopez:2005:MCL:1069774.1069778} is a concurrent linear logic programming language that integrates
+both forward and backward-chaining search, where the backward-chaining phase is done sequentially but
+the forward-chaining is done concurrently inside a monad. The backward-chaining phase is suspended
+between the forward-chaining phases, where a fix-point is computed.
+LolliMon is derived from the concurrent logical framework called CLF~\cite{Watkins:2004uq}. 
+\makeatletter{}\section{Conclusions}
+
+We have presented a parallel virtual machine for executing
+forward-chaining linear logic programs, with particular focus on
+thread management, code organization, fact indexing, rule execution
+and database organization for fast insertion, lookup, and deletion of
+linear facts. Experimental results show that our VM is able to scale
+the execution of programs when run with up to 16 threads.
+Our results also show the importance of having an efficient indexing mechanisms
+for facts. With our dynamic indexing, the VM automatically detects which predicates
+need to be indexed in order to improve performance.
+Due to these and other optimizations, the VM fairs relatively well against other programming languages, including
+compiled languages. Moreover, since LM programs are
+concurrent by default, we can easily get better performance from the
+start by executing them with multiple threads.
+As further work, we want to improve parallel scalability and
+take advantage of linear logic to perform whole-program
+optimizations, including computing program invariants, loop detection
+in rules and bypass of rule priorities.
+
+We think that our virtual machine is an excellent starting point to make
+logic programming more desirable in the data-mining, machine learning and
+distributed/parallel programming community. Moreover, our virtual machine can
+be easily extended to execute over computer networks or to execute programs on really big datasets.
+In a nutshell, LM provides a concise way to describe
+graph-based algorithms that can be more easily reasoned about, a clear advantage over
+competing systems.
+
+\makeatletter{}\acks
+
+This work is partially funded by the ERDF (European Regional
+Development Fund) through the COMPETE Programme; by FCT (Portuguese
+Foundation for Science and Technology) through the Carnegie Mellon
+Portugal Program and project SIBILA (NORTE-07-0124-FEDER-000059); and
+by the Qatar National Research Fund under grant NPRP
+09-667-1-100. Flavio Cruz is funded by the FCT grant
+SFRH/BD/51566/2011.
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \bibliographystyle{abbrvnat}
-
-% The bibliography should be embedded for final submission.
-%\bibliography{refs}
 

### 2014-08-08 paper.tex
final changes.

@@
 Persistent Facts & $\Gamma$ & $::=$ & $\cdot \; | \; \Gamma, \bang p(\hat{t})$ \\
+Terms            & $t$ & $::=$ & $\m{node}(n) \; | \; \m{int}(n) \; | \; \m{float}(f) \; | $\\
+                 &     &       & $ \m{string}(s) \; | \; \m{bool}(b) \; | \; l \; | \; \m{pair}(t,t)$ \\
+Lists            & $l$ & $::=$ & $\m{nil} \; | \; [ t \; | \; l ]$\\
 \end{tabular}
@@
 
+Figure~\ref{fig:overview} presents the layout of our virtual machine
+for a program with six nodes and two running threads. Each thread
+space includes the nodes owned by the thread (the dotted arrows
+represent the edges between nodes) and a \emph{Work Queue}, a single
+linked list containing \emph{active nodes}, i.e., nodes that have new
+facts to process. Initially, the \emph{Work Queue} is filled with all
+the nodes of the thread in order to derive the axioms.
+
+\begin{figure*}[t]
+\centering
+\includegraphics[width=\textwidth]{figures/overview.pdf}
+\caption{Layout of the virtual machine}
+\label{fig:overview}
+\end{figure*}
+
 During execution, threads can steal nodes of other threads to keep
@@
 algorithm. The pseudo-code for the main thread loop is shown in
-Fig.~\ref{code:work_loop}. In each round, a thread inspects its work
-queue for active nodes with new candidate rules and, if there is any,
-procedure \texttt{process\_node()} executes it. If the work queue is
-empty, the thread attempts to steal one node from another
+Fig.~\ref{code:work_loop}. In each round, a thread inspects its \emph{Work Queue}
+for active nodes with new candidate rules and, if there is any,
+procedure \texttt{process\_node()} is called on the target node.
+Otherwise, if the \emph{Work Queue} is empty, the thread attempts to
+steal one node from another
 thread. Starting from a random thread, it cycles through all the
@@
          for (i = 0; i < NUM_THREADS && !node; i++)
-            target = (target + 1)             node = steal_node_from_thread(target)
+            target = (target + 1)
+            node = steal_node_from_thread(target)
             if (node) break
@@
 
-\begin{figure*}[t]
-\centering
-\includegraphics[width=\textwidth]{figures/overview.pdf}
-\caption{Layout of the virtual machine}
-\label{fig:overview}
-\end{figure*}
-
-Figure~\ref{fig:overview} presents the layout of our virtual machine
-for a program with six nodes and two running threads. Each thread
-space includes the nodes owned by the thread (the dotted arrows
-represent the edges between nodes) and a \emph{Work Queue}, a single
-linked list containing \emph{active nodes}, i.e., nodes that have new
-facts to process. Initially, the \emph{Work Queue} is filled with all
-the nodes of the thread in order to derive the
-axioms. Figure~\ref{fig:overview} also illustrates the internal
+Figure~\ref{fig:overview} also illustrates the internal
 structure layout of a node, which includes: the database of linear
@@
 proves goals by lazily managing the context of linear resources during backward-chaining proof search.
-LolliMon~\cite{Lopez:2005:MCL:1069774.1069778} is a concurrent linear logic programming language that integrates
+Forum~\cite{Miller94} is an extension of Lolli that is based on a classical proof system.
+While Lolli uses an intuitionistic proof system that supports only one conclusion, Forum allows multiple conclusions.
+Forum is therefore better adapted to represent concurrent computations than Lolli, since many multiple goals need to be proved at once.
+
+Another extension of Lolli is LolliMon~\cite{Lopez:2005:MCL:1069774.1069778}, a concurrent linear logic programming language that integrates
 both forward and backward-chaining search, where the backward-chaining phase is done sequentially but
@@
 LolliMon is derived from the concurrent logical framework called CLF~\cite{Watkins:2004uq}.
+
+Lygon~\cite{Harland96} is a backward-chaining linear logic programming language that extends Prolog with linear resources.
+Lygon is not only able to run Prolog programs but also allows the body of clauses to be linear that can be used exactly once in each query.
+In contrast to Prolog, where there is no clause selection, Lygon needs to use heuristics to decide which clause to prove first, since that will help finding a proof and finding it efficiently.
+Notably, Lygon is very suitable to solve graph-based problems due to the mutable state provided by linear clauses.
+
 \makeatletter{}\section{Conclusions}
@@
 
+\bibitem[Harland(1996)]{Harland96}
+J.~Harland, D.~Pym and M.~Winikoff
+\newblock Programming in Lygon: An Overview
+\newblock In \emph{International Conference on Algebraic Methodology and Software Technology (AMAST)}, 391-405,  1996.
+
 \bibitem[Hermenegildo(1986)]{Hermenegildo:1986:AMB:913061}
@@
 \newblock Monadic concurrent linear logic programming.
-\newblock In \emph{Proceedings of the 7th ACM SIGPLAN International Conference
-  on Principles and Practice of Declarative Programming}, PPDP '05, pages
+\newblock In \emph{International Conference on Principles and Practice of Declarative Programming}, PPDP '05, pages
   35--46, New York, NY, USA, 2005.
@@
 \newblock In \emph{in Computational Logic}, pages 1--5, 1985.
+
+\bibitem[Miller(1994)]{Miller94}
+D.~Miller.
+\newblock A multiple-conclusion meta logic
+\newblock In \emph{Logic in Computer Science (LICS)}, pages 272--281, 1994.
 

### 2014-08-12 paper.tex
update

@@
 several machine models have been developed with parallelism in
-mind~\cite{Kara:1997:AMM:265274}.  One example of such machine is the
+mind~\cite{Kara:1997:AMM:265274}.  One example is the
 Parallel Virtual Machine (PVM)~\cite{Sunderam90pvm:a}, which serves as
@@
 tried in parallel. Different abstract machines for AND-parallelism has
-been developed on top of the
+been developed for the
 WAM~\cite{Hermenegildo:1986:AMB:913061,Lin:1988:AEL:900478}.  For
 OR-parallelism we have several models such as: the SRI
-model~\cite{Warren:1987:OEM:67683.67699}, the Argonne
-model~\cite{ButlerDLOOS88}, the MUSE model~\cite{Ali:1990fk} and the
-BC machine~\cite{Ali88}. While all those models have been developed
-using the WAM, there are also some parallel machines totally different
-from the WAM, such as the PPAM~\cite{Kacsuk:1990:EMP:533578}, which is
-based on a data-flow model.
+model~\cite{Warren:1987:OEM:67683.67699}, the MUSE model~\cite{Ali:1990fk} and the
+BC machine~\cite{Ali88}. Although these models are built on top of the WAM,
+some machines use different approaches, such as the
+PPAM~\cite{Kacsuk:1990:EMP:533578} with its
+data-flow model.
 
@@
 Forum is therefore better adapted to represent concurrent computations than Lolli, since many multiple goals need to be proved at once.
-
 Another extension of Lolli is LolliMon~\cite{Lopez:2005:MCL:1069774.1069778}, a concurrent linear logic programming language that integrates
@@
 between the forward-chaining phases, where a fix-point is computed.
-LolliMon is derived from the concurrent logical framework called CLF~\cite{Watkins:2004uq}.
+LolliMon is derived from the logical framework called CLF~\cite{Watkins:2004uq}.
 
@@
 
+We thank the anonymous reviewers for their insightful comments and suggestions
+that helped us improve the quality of the paper.
+
 This work is partially funded by the ERDF (European Regional
@@
   137--151. 2005.
-
-\bibitem[Butler et~al.(1988)Butler, Disz, Lusk, Olson, Overbeek, and
-  Stevens]{ButlerDLOOS88}
-R.~Butler, T.~Disz, E.~L. Lusk, R.~Olson, R.~A. Overbeek, and R.~L. Stevens.
-\newblock Scheduling or-parallelism: An argonne perspective.
-\newblock In R.~A. Kowalski and K.~A. Bowen, editors, \emph{ICLP/SLP}, pages
-  1590--1605. MIT Press, 1988.
-\newblock ISBN 0-262-61056-6.
-\newblock URL
-  \url{http://dblp.uni-trier.de/db/conf/iclp/iclp88.html#ButlerDLOOS88}.
 

### 2014-09-15 paper.tex
fix

@@
 
-We think that our virtual machine is an excellent starting point to make
+We think that our virtual machine is a promising starting point to make
 logic programming more desirable in the data-mining, machine learning and

### 2015-11-19 presentation/slides.tex
Add presentation.

\documentclass{beamer}

\usepackage[english]{babel}
\usepackage{beamerthemeBoadilla}
%\usepackage{beamerthemeCambridgeUS}
%\usepackage{beamerthemeRochester}
%\usepackage{beamerthemeSzeged}
%\usepackage{beamerthemeMontpellier}
%\usepackage{beamerthemedefault}
\usepackage{url}
\usepackage{verbatim}
\usepackage[utf8]{inputenc}
\usepackage{multirow}
\usepackage{fancyvrb}
\usepackage{proof-dashed}
\input{macros}
\input{fp-macros}

\newsavebox{\mysavebox}

\def\Tiny{\fontsize{6pt}{6pt}\selectfont}

\title{Design and Implementation of a Multithreaded Virtual Machine for Executing Linear Logic Programs}
\author[Flávio Cruz]{Flávio Cruz {\small \texttt{<fmfernan@cs.cmu.edu>}}\\
\scriptsize{\textbf{Authors}:\\
Flavio Cruz (CMU/UP)\\
Ricardo Rocha (UP)\\
Seth Goldstein (CMU)}}

\institute[CMU/UP]{Carnegie Mellon University \\ Pittsburgh, PA 15213, USA \and
CRACS \& INESC TEC, Faculty of Sciences, University Of Porto\\
Rua do Campo Alegre, 1021/1055, 4169-007 Porto, Portugal}
\date{\today}

\let\oldalert\alert
\renewcommand{\alert}[2][]{%
  \if\relax\detokenize{#1}\relax% http://tex.stackexchange.com/q/53068/5764
    \oldalert{#2}% Default overlay
  \else
    \oldalert<#1>{#2}% Specific overlay
  \fi}

\begin{document}

\frame{\titlepage}

\AtBeginSection[] { \begin{frame}<beamer>
\frametitle{Plan} \tableofcontents[currentsection]
\end{frame}}

\section{Introduction}

\frame
{
  \frametitle{Logic and Parallel Programming}
  \begin{itemize}
     \item Logic programs are declarative and easier to parallelize than imperative programs
     \begin{itemize}
      \item Logic program: set of rules
      \item Rules can be explored in parallel
      \item Imperative programming enforces a specific sequence of execution
     \end{itemize}
     \item Two paradigms:
     \begin{itemize}
      \item Backwards-chaining: Prolog
      \begin{itemize}
         \item Or-parallelism
         \item And-parallelism
      \end{itemize}
      \item Forwards-chaining: Datalog, \textbf{Linear Meld (LM)}
     \end{itemize}
  \end{itemize} 
}

\frame
{
   \frametitle{Forward Chaining Logic Programming}
   \begin{itemize}
      \item We start with a set of rules and a database of logical facts
      \item The database of facts is used to apply more rules
      \item Program stops when no more information can be derived
      \item Datalog: since all facts are persistent, concurrent rule application is trivial
      \item However, Datalog falls short when trying to express more stateful programs
   \end{itemize}
}

\subsection{Linear Meld}

\frame
{
   \frametitle{Linear Meld}
   \begin{itemize}
      \item Linear Meld (LM) is a Datalog-like language that incorporates Linear Logic
      \begin{itemize}
         \item Two kinds of facts: persistent facts and linear facts
         \item Linear facts can be retracted
      \end{itemize}
      \item Fully declarative language without extra-logical features
      \item LM programs are represented as a graph data structure
      \begin{itemize}
         \item Logical facts reside on a node of the graph
         \item Rules are locally restricted so that only node facts are manipulated
         \item No synchronization problems
      \end{itemize}
      %\item For a more in-depth view read our ICLP paper "A Linear Logic Programming Language for Concurrent Programming over Graph Structures"
   \end{itemize}
}

\subsection{Example}

\begin{frame}[fragile]
  \frametitle{Bipartiteness Checking}
  \begin{columns}[t]
     \column{.45\textwidth}
     \begin{block}{Program}
       \begin{Verbatim}[fontsize=\tiny,commandchars=\\\{\},frame=single]
\alert[1]{type route edge(node, node).}
\alert[1]{type linear mark(node, int).}
\alert[1]{type linear uncolored(node).}
\alert[1]{type linear colored(node, int).}
\alert[1]{type linear fail(node).}

fun next(int X) : int =
   if X <> 1 then 1 else 2 end.

\alert[2,6]{mark(@1, 1).}
\alert[2]{!edge(@1, @2). !edge(@1, @3).}
\alert[2]{!edge(@2, @4). !edge(@3, @4).}
\alert[2]{uncolored(A).}

\alert[3,7,8,9,10]{mark(A, P), uncolored(A)}
   \alert[3,7,8,9,10]{-o \{B | !edge(A, B) | mark(B, next(P))\},}
      \alert[3,7,8,9,10]{colored(A, P).}

\alert[4,11]{mark(A, P), colored(A, P)}
   \alert[4,11]{-o colored(A, P).}
\alert[5]{mark(A, P1), colored(A, P2), P1 <> P2}
   \alert[5]{-o fail(A).}
\alert[5]{mark(A, P), fail(A)}
   \alert[5]{-o fail(A).}
       \end{Verbatim}
     \end{block}
      \column{.5\textwidth}
      \begin{block}{\only<1>{Predicates}\only<2>{Axiom}\only<3>{First rule}\only<4>{Second rule}\only<5>{Third and fourth rule}\only<6-11>{Execution}}
         \centering
         {\scriptsize
         \only<1>{\begin{itemize}
                \item The first argument of every predicate must be typed as \texttt{node}.
                \item Predicates specified as \texttt{route} inform the compiler about the graph data structure.
                \item Predicates specified as \texttt{linear} turns facts of the predicate into linear facts, which can be asserted or retracted.
                \item Predicates not specified as \texttt{linear} are persistent.
                \item Nodes are either \texttt{colored/2} or \texttt{uncolored/1}.
             \end{itemize}}
         \only<2>{\begin{itemize}
                \item Axioms are rules without bodies that are added to the database as soon as the program starts.
                \item Node literals are written as \texttt{@X}, where \texttt{X} is the node number.
             \end{itemize}}
         \only<3>{\begin{itemize}
               \item If a node is scheduled to be \texttt{mark/1}'ed and is \texttt{uncolored/1}, then we can assign the color \texttt{P} to the node by deriving \texttt{colored(A,P)}.
               \item We use a comprehension to mark the neighbor nodes \\
               \texttt{\{B|!edge(A,B)|mark(B,next(P))\}}:
               \begin{itemize}
                  \item {\tiny For every \texttt{!edge(A, B)} in the database we derive a \texttt{mark(B,next(P))}.}
                  \item {\tiny With \texttt{next(P)} we attempt to color the neighbor node with the opposite color.}
               \end{itemize}
         \end{itemize}}
         \only<4>{\begin{itemize}
            \item If a node is to be marked with color \texttt{P} and has color \texttt{P}, then we keep it that way.
         \end{itemize}}
         \only<5>{\begin{itemize}
            \item However, if the colors are different then we derive \texttt{fail/1}.
            \item ... same if the coloring process has already failed.
         \end{itemize}}
         \only<6-11>{
         \begin{figure}[ht]
            \includegraphics<6>[height=4.5cm]{bipartiteness1.pdf}
            \includegraphics<7>[height=4.5cm]{bipartiteness2.pdf}
            \includegraphics<8>[height=4.5cm]{bipartiteness3.pdf}
            \includegraphics<9>[height=4.5cm]{bipartiteness4.pdf}
            \includegraphics<10>[height=4.5cm]{bipartiteness5.pdf}
            \includegraphics<11>[height=4.5cm]{bipartiteness6.pdf}
         \end{figure}
         }
      }
      \end{block}
  \end{columns}
\end{frame}

\section{Virtual Machine}

\begin{frame}[fragile]
   \frametitle{Virtual Machine}
   \begin{itemize}
      \item Register based machine implemented in C++
      \item Separate compiler compiles programs to byte-code
      \item Uses POSIX Threads for multithreading
      \item Optimized for:
      \begin{itemize}
         \item Parallelism
         \item Database storage and lookup of linear facts
         \item Fast rule execution
      \end{itemize}
   \end{itemize}
\end{frame}

\begin{frame}[fragile]
   \frametitle{Parallelism}
  \begin{columns}[t]
     \column{.45\textwidth}
{\small
      \begin{itemize}
      \item Graph is partitioned into P sub-graphs, where P is the number of threads
      \item Each thread processes the nodes of its sub-graph
      \item Work stealing is employed whenever threads start to starve
      \end{itemize}
}
     \column{.45\textwidth}
       \begin{block}{Partitioning}
       \begin{center}
          \includegraphics[height=5.5cm]{graph_coordination.pdf}
          \end{center}
       \end{block}
   \end{columns}
\end{frame}

\begin{frame}[fragile]
   \frametitle{Overview}
   \includegraphics[height=7.5cm]{overview.pdf}
\end{frame}

\begin{frame}[fragile]
   \frametitle{Overview}
   \includegraphics[height=7.5cm]{overview2.pdf}
\end{frame}

\begin{frame}[fragile]
   \frametitle{Byte Code}
   \begin{columns}[t]
   \column{.45\textwidth}
     \begin{block}{Program}
       \begin{Verbatim}[fontsize=\tiny,commandchars=\\\{\},frame=single]
type route edge(node, node).
type linear mark(node, int).
type linear uncolored(node).
type linear colored(node, int).
type linear fail(node).

fun next(int X) : int =
   if X <> 1 then 1 else 2 end.

mark(@1, 1).
!edge(@1, @2). !edge(@1, @3).
!edge(@2, @4). !edge(@3, @4).
uncolored(A).

\alert[1]{mark(A, P), uncolored(A)}
   \alert[1]{-o \{B | !edge(A, B) | mark(B, next(P))\},}
      \alert[1]{colored(A, P).}

mark(A, P), colored(A, P)
   -o colored(A, P).
mark(A, P1), colored(A, P2), P1 <> P2
   -o fail(A).
mark(A, P), fail(A)
   -o fail(A).
\end{Verbatim}
     \end{block}
   \column{.5\textwidth}
   \begin{block}{Byte Code}
\begin{Verbatim}[fontsize=\tiny]
LINEAR ITERATE OVER uncolored MATCHING TO reg 0
  LINEAR ITERATE OVER mark MATCHING TO reg 1
    ALLOC colored TO reg 2
    MVFIELDFIELD 1.0 TO 2.0
    ADDLINEAR reg 2
    RESET LINEAR
      PERSISTENT ITERATE OVER edge MATCHING TO reg 2
        ALLOC mark TO reg 3
        PUSH
        PUSH
        PUSH REGS
        MVFIELDREG 1.0 TO reg 0
        MVPCOUNTERSTACK TO STACK 33
        CALLF 0
        POP REGS
        MVSTACKFIELD STACK 0 TO 3.0
        POP
        POP
        MVFIELDREG 2.0 TO reg 4
        SEND reg 3 TO reg 4
        TRY NEXT
    END LINEAR
    REMOVE reg 1
    REMOVE reg 0
    TRY NEXT
  NEXT
\end{Verbatim}
   \end{block}
   \end{columns}
\end{frame}

\begin{frame}[fragile]
   \frametitle{Overview}
   \includegraphics[height=7.5cm]{overview3.pdf}
\end{frame}

\begin{frame}[fragile]
   \frametitle{Rule Engine}
   \begin{itemize}
      \item Compact engine to detect which rules need to be executed (by priority)
      \item Structures:
      \begin{itemize}
         \item Predicates Count: compact bitmap to keep count of predicates
         \item Active Bitmap: rules "satisfied" with the current database
         \item Rule Queue: queue of rules to be executed
         \item Dropped Bitmap: rules that no longer are satisfied with the current database after a rule application
         \item Predicates Bitmap: bitmap of predicates/facts derived during a rule application.
      \end{itemize}
      \item How to update Rule Queue?
      \begin{itemize}
         \item Remove rules from Dropped Bitmap (XOR)
         \item Iterate through Predicates Bitmap and add all satisfied rules (Activate Bitmap) (OR)
      \end{itemize}
      \item Execute next rule? Get the least significant index
      \begin{itemize}
         \item \texttt{rule = FFS(rule\_queue)}
         \item \texttt{rule\_queue = rule\_queue \& (rule\_queue - 1)}
      \end{itemize}
   \end{itemize}
\end{frame}

\begin{frame}[fragile]
   \frametitle{Overview}
   \includegraphics[height=7.5cm]{overview3.pdf}
\end{frame}

\begin{frame}[fragile]
   \frametitle{Database Structures}
   \begin{itemize}
      \item Tries are used for storing persistent facts
      \item Doubly linked lists for linear facts
      \begin{itemize}
         \item $\mathcal{O}(1)$ retraction
         \item $\mathcal{O}(1)$ assertion
         \item $\mathcal{O}(N)$ lookup
      \end{itemize}
      \item Hash tables also used for linear facts
      \begin{itemize}
         \item Indexing of facts using one of the arguments
         \item $\mathcal{O}(1)$ retraction and assertion
         \item $\mathcal{O}(1)$ lookup on average
         \item Full dynamic support
      \end{itemize}
   \end{itemize}
\end{frame}

\begin{frame}[fragile]
   \frametitle{Storing linear facts}
   \begin{center}
      \includegraphics[height=6cm]{hash_table.pdf}
   \end{center}
   \begin{itemize}
      \item {\large Predicate \texttt{a(int, int)} indexed by the second argument}
   \end{itemize}
\end{frame}

\begin{frame}[fragile]
   \frametitle{Database Indexing}
   \begin{itemize}
      \item The virtual machine identifies during runtime the best indexing argument of each predicate
      \item Indexing algorithm is performed during the first few seconds of execution in 3 phases:
      \begin{itemize}
         \item Gathering of lookup data by keeping a counter for each argument
         \item Selecting the two best candidate arguments for each predicate
         \begin{itemize}
            \item If a predicate only uses 1 argument for lookup, then we use that argument
         \end{itemize}
         \item Compute entropy score for the two top arguments
      \end{itemize}
      \item Entropy: the more disorder there is for a given argument, the better
      \begin{itemize}
         \item The algorithm samples a few node databases to compute the entropy
      \end{itemize}
   \end{itemize}
\end{frame}

\section{Experimental Results}

\begin{comment}
\begin{frame}[fragile]
   \frametitle{Implemented programs}
   \begin{itemize}
      \item Belief Propagation
      \item MiniMax
      \item N-Queens
      \item PageRank
      \item Neural Networks
      \item Graph algorithms: shortest path, search, coloring
   \end{itemize}
\end{frame}
\end{comment}

\begin{frame}[fragile]
   \frametitle{Experimental Results: Scalability}
   \begin{columns}[t]
      \column{.5\textwidth}
      \begin{figure}[b]
         \includegraphics[width=\textwidth]{../figures/results-shortest-uspowergrid.pdf}
         \caption{Shortest Distance for a graph with around 5,000 nodes and 13,000 edges.}
      \end{figure}
      \column{.5\textwidth}
      \begin{figure}[b]
         \includegraphics[width=\textwidth]{../figures/results-minimax.pdf}
         \caption{MiniMax algorithm for the Tic-Tac-Toe game.}
      \end{figure}
   \end{columns}
\end{frame}

\begin{frame}[fragile]
   \frametitle{Experimental Results: Scalability}
   \begin{columns}[t]
      \column{.5\textwidth}
      \begin{figure}[b]
         \includegraphics[width=\textwidth]{../figures/results-pagerank-search-engines.pdf}
         \caption{PageRank for a webpage graph.}
      \end{figure}
      \column{.5\textwidth}
      \begin{figure}[b]
         \includegraphics[width=\textwidth]{../figures/results-ggc.pdf}
         \caption{Greedy Graph Coloring for a random graph.}
      \end{figure}
   \end{columns}
\end{frame}

\begin{frame}[fragile]
   \frametitle{Experimental Results: Scalability}
   \begin{columns}[t]
      \column{.5\textwidth}
      \begin{figure}[b]
         \includegraphics[width=\textwidth]{../figures/results-13queens.pdf}
         \caption{N-Queens program (13x13~board)}
      \end{figure}
      \column{.5\textwidth}
      \begin{figure}[b]
         \includegraphics[width=\textwidth]{../figures/results-bp.pdf}
         \caption{Belief propagation (400x400 image).}
      \end{figure}
   \end{columns}
\end{frame}

\begin{frame}[fragile]
   \frametitle{Experimental Results: Absolute Execution Time}
   \begin{figure}[b]
      \begin{tabular}{ | l | l | l | l |}
       \hline
       Size & C & Python & YAP Prolog \\ \hline\hline
       10 & 16.92 & \textbf{0.62} & 5,42 \\
       11 & 21.59 & \textbf{0.64} & 6.47 \\
       12 & 10.32 & \textbf{0.73} & 7.61 \\
       13 & 14.35 & \textbf{0.88} & 10.38 \\
       \hline
       \end{tabular}
       \caption{N-Queens problem.}
    \end{figure}
    \begin{figure}[b]
       \begin{tabular}{ | l | l | l | l |}
        \hline

        Size & C & Python & GraphLab \\ \hline\hline
        10 & \textbf{1.00} & \textbf{0.03} & \textbf{1.00} \\
        50 & \textit{1.77} & \textbf{0.04} & \textit{1.73} \\
        200 & \textit{1.99} & \textbf{0.05} & \textit{1.79} \\
        400 & \textit{2.00} & \textbf{0.04} & \textit{1.80} \\
        \hline
        \end{tabular}
        \caption{Belief Propagation program.}
    \end{figure}
\end{frame}

\begin{frame}[fragile]
   \frametitle{Experimental Results: Memory Usage}
   \begin{center}
\begin{tabular}{c|c|c|c}
\textbf{Program} & \textbf{Facts} & \textbf{Memory} & \textbf{Per Fact} \\
\hline\hline
PageRank & 1,180,603 & 203,832 KB & 176 bytes \\
GGC & 2,363,536 & 292,682 KB & 127 bytes \\
SD & 502,347 & 45,387 KB & 92 bytes \\
MiniMax & 3 & 886,443 KB & 295,481 KB \\
N-Queens & 74,557 & 55,408 KB & 760 bytes \\
BP & 2,235,200 & 617,417 KB & 283 bytes \\
\hline\hline
\multicolumn{3}{l}{Average (without MiniMax) } & 288 bytes \\
\end{tabular}
\end{center}
\end{frame}

\begin{frame}
   \frametitle{Conclusions}
   \begin{itemize}
      \item LM programs are succinct
      \item The LM language is general enough to be ported to many distributed architectures
      \item The virtual machine aims to bridge the gap between logic programming and multicore technology
      \begin{itemize}
         \item Goal: execute large programs scalably and efficiently
      \end{itemize}
      \item Current work: use derivation rules to schedule computation
      \begin{itemize}
         \item Nodes have priority
         \item A given computation order will give the same result but will be complete faster
      \end{itemize}
   \end{itemize}
\end{frame}

\begin{frame}{Thank You}
\begin{center}
{\Huge Thank you for your time!}
\end{center}
\end{frame}

\end{document}
